---
title: "Mastering Statistical Process Control"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee



## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(pdftools)
```



# Part 3: Putting SPC into Practice - The Cases

Page 59 - 

# 5. Investigating variation in chemical concentration (X and MR)

How control charts were used to identify, investigate and prove the cause of fluctuations in results. 

Demonstrates how one control chart was used to: 

* identify a performance problem

* generate theories as to the causes

* test and ultimately "prove" the theory that differences in measurments were due to different chemical analysts and not actual process performance. 





# Part 5: Developing SPC Skills: Organizational Review Questions, Workshops and Exercises


Part 5 is to help develop SPC understanding and skills

## 29. The Rods Experiments- A practical case study 

```{r}
# Rods run data (Chart 29.5)
rods <- data.frame(sample_no = 1:33, 
                   rod_length_in = c(5.9, 6.1, 6.2, 5.9, 6.0, 6.0, 5.8, 6.4, 6.1, 
                                    5.9, 5.6, 5.8, 6.2, 6.1, 5.8, 6.2, 6.1, 5.8, 
                                    5.9, 6.0, 6.0, 5.8, 6.5, 6.1, 5.9, 6.2, 5.8, 
                                    5.7, 5.9, 6.0, 6.1, 5.8, 6.0))

# Creating a run chart 
rods %>%
  ggplot(aes(x = sample_no, y = rod_length_in)) + 
  geom_point() + 
  geom_line()

```

## 29.1 Compleing the X/MR control chart

```{r}
# the average
x_mean <- mean(rods$rod_length_in)

# average moving range
moving_range <- data.frame(sample_no = 2:33, 
                           mr = abs(diff(rods$rod_length_in)))

moving_range_mean <- mean(moving_range$mr)

# calculate standard deviation from the moving range
s_mr <- moving_range_mean/1.128 

# Upper and lower warning limits
UWL <- x_mean + 2 * s_mr
LWL <- x_mean - 2 * s_mr

# upper and lower action limits
UAL <- x_mean + 3 * s_mr
LAL <- x_mean - 3 * s_mr

# MR chart upper action limits
UAL_mr <- 3.27 * moving_range_mean


# Create Moving range control chart
moving_range %>%
  ggplot(aes(x = sample_no, y = mr)) +
  geom_point() + 
  geom_line() + 
  geom_hline(yintercept = c(moving_range_mean, UAL_mr), 
             linetype = c("solid", "dashed")) 


rods %>%
  ggplot(aes(x = sample_no, y = rod_length_in)) +
  geom_point() + 
  geom_line() + 
  geom_hline(yintercept = c(x_mean, UWL, LWL, UAL, LAL), 
             linetype = c("solid", "dashed", "dashed", "twodash", "twodash")) +
  labs(x = "Sample No", y = "Rod Length (in)")


```

The process is in a state of control, and we can extend the average, warning and control limits beyong the data as these represent teh expected process performance into the future. 


## 29.2 Drawing and interpreting a cusum chart 

```{r}
# create a process change data frame
rods_process_changed <- data.frame(sample_no = 34:47, 
                                   rod_length_in = c(6.2, 6.1, 5.9, 6.3, 6.4, 5.9, 
                                                     6.4, 6.3, 6.2, 6.2, 6.5, 6.2, 
                                                     6.1, 6.7))

# Cusum data calculaiton 
rods_with_process_changed <- rods %>%
  bind_rows(rods_process_changed) %>%
   # minus the target value 6 (or average)
  mutate(rod_length_in_scaled = rod_length_in - 6) %>%
  # calculate cusum
  mutate(cusum = cumsum(rod_length_in_scaled))


# plot
p <- rods_with_process_changed %>%
  ggplot(aes(x = sample_no, y = rod_length_in)) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 34, linetype = "dashed") 

p

# cusum plot 
rods_with_process_changed %>%
  ggplot(aes(x = sample_no, y = cusum)) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 34, linetype = "dashed") 


```


