---
title: "Applied Predictive Modeling - Max Kuhn and Kjell Johnson"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


**Course Description:**
 
weblinke related with the book: http://appliedpredictivemodeling.com/

## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)

library(AppliedPredictiveModeling)
library(caret)
library(corrplot)
library(e1071)
```

# 1. Introduction 


# 2. A Short Tour of the Predictive Modeling Process

* The first step in any model building is to understand the data. 

```{r}
# Fuel Economy Data
data(FuelEconomy)
head(cars2011)

library(lattice)

# Plot shown in the text
cars2010 <- cars2010[order(cars2010$EngDispl), ]
cars2011 <- cars2011[order(cars2011$EngDispl), ]

# Add the "year"
plotData <- cars2010 %>% mutate(Year = "2010 Model Year") %>%
  bind_rows(cars2011 %>% mutate(Year = "2011 Model Year")) %>%
  mutate(Year = factor(Year))

str(plotData)

## Plot
plotData %>%
  ggplot(aes(x = EngDispl, y = FE)) +
  geom_point(alpha = 0.5) +
  labs(x = "Engine Displacement", 
       y = "Fuel Efficiency (MPG)", 
       title = "The relationship between engine displacement and fuel efficiency")  +
  facet_wrap(~ Year)
```

The relationship is somewhat linear but does exhibit some curvature towards the extreme ends of the displacement axis. 


After first understanding the data, the next step is to build and evaluate a model on the data. 


A simple model: 2010 as Training set

```{r}
model_lm <- lm(FE ~ EngDispl, data = cars2010)

summary(model_lm)

# predicted value
predicted <- predict(object = model_lm, data = cars2010)

# plot
cars2010 %>%
  mutate(predicted = predict(object = model_lm, data = .)) %>%
  ggplot(aes(x = FE, y = predicted)) +
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", size = 1, col = "red") +
  scale_x_continuous(limits = c(10, 70)) + 
  scale_y_continuous(limits = c(10, 70))

```

# 3. Data Pre-processing

## 3.1 Case Study: Cell Segmentation in High-Content Screeening


## 3.2 Data Transformations for Individual Predictors

* Centering and Scaling

* Transformations to Resolve Skewness


## 3.3 Data Transformations for Multiple Predictors 

* Transformations to Resolve Outliers

Outliers are samples that are exeptionally far from the mainstream of the data. 

* Data Reduction and Feature Extraction 


## 3.4 Dealing with Missing Values

"informative missingness", which can induce significant bias in the model. 


## 3.5 Removing Predictors

A predictor variable that has a single unique value: a Zero variance predictor. 

Some predictors might have only a handful of unique values that occure with very low frequencies: "near-zero variance predictors". 

A rule of thumb for detecting near-zero variance predictors is: 

* The fraction of unique values over the sample size is low (say 10%)

* The ratio of the frequency of the most prevalent value to the frequency of the second most prevalent value is large (say around 20)


Between-predictor correlations: 

Collinearity is the technical term for the situation where a pair of predictor variables have a substantial correltion with each other. 


## 3.6 Adding Predictors


## 3.7 Binning Predictors

## 3.8 Computing 

```{r}
# Find functions for creating a confusion matrix within the currently loaded packages
apropos("confusion")

```

```{r}
# Raw segmentation data set
data("segmentationOriginal")

# feasture names
names(segmentationOriginal)

# just focus on the training set samples
segData <- subset(x = segmentationOriginal, Case == "Train")

# The Class and Cell fields are saved into separate vectors
callID <- segData$Cell
class <- segData$Class
case <- segData$Case

# Now remove the columns
segData <- segData[, -(1:3)]


# The original data contained several "status" columns and remove them-- A good method to remove columns 
statusColNum <- grep(pattern = "Status", x = names(segData))

segData <- segData[, -statusColNum]

dim(segData)
```

### 3.8.1 Transformations

```{r}
# Calculates the sample skewness statistic for each predictor: 
e1071::skewness(segData$AngleCh1)

# Calculate the skewness across columns
skewValues <- map_df(segData, .f = e1071::skewness)

head(skewValues)

max(skewValues)

# Determine which type of transformation should be used, the MASS package contains the boxcox function. Although this funciton estimates lambda, it does not creat the transformed variables. 


# BoxCox transformations
Ch1AreaTrans <- caret::BoxCoxTrans(y = segData$AreaCh1)
Ch1AreaTrans

class(Ch1AreaTrans)


# The original data
head(segData$AreaCh1)

# After transformation 
predict(object = Ch1AreaTrans, head(segData$AreaCh1))

(head(segData$AreaCh1)^(-0.9) - 1)/(-0.9)


# Another caret function, preProcess
# ?caret::preProcess

# The base R funciton *prcomp* can be used for PCA

pcaObject <- prcomp(x = segData, center = TRUE, scale. = TRUE)
names(pcaObject)

# Calculate the cumulative percentage of variance which each component account  for 
percentVariance <- pcaObject$sdev^2/sum(pcaObject$sdev^2) * 100

percentVariance[1:5]


var(pcaObject$x[, 1:5]) # PC1 has the largest variance
cor(pcaObject$x[, 1:5]) # no correlation among PC1, PC2


# the transformed values are store in pcaObject as a sub-object called x: 
head(pcaObject$x[, 1:5])

# Variable loadings
head(pcaObject$rotation[, 1:5])

# spatiaSign Transformation
spatialSign(rnorm(5))

spatialSign(matrix(rnorm(12), ncol = 3))



```

To administer a series of transformations to multiple data sets, the caret
class preProcess has the ability to transform, center, scale, or impute values,
as well as apply the spatial sign transformation and feature extraction. The
function calculates the required quantities for the transformation. After calling
the preProcess function, the predict method applies the results to a set
of data. For example, to Box–Cox transform, center, and scale the data, then
execute PCA for signal extraction, the syntax would be:

```{r}
trans <- preProcess(segData, method = c("BoxCox", "center", "scale", "pca"))

trans

# Apply the transformations:
transformed <- predict(trans, segData)
head(transformed)
```
The order in which the possible transformation are applied is *transformation*,
*centering*, *scaling*, *imputation*, *feature extraction*, and then *spatial sign*.
Many of the modeling functions have options to center and scale prior
to modeling. For example, when using the *train* function (discussed in later
chapters), there is an option to use *preProcess* prior to modeling within the
resampling iterations.

### 3.8.2 Filtering 

To filter for near-zero variance predictors, the *caret* package function *nearZeroVar* will return the column numbers of any predictors that fulfill the conditions outlined in Sect. 3.5.

```{r}
nearZeroVar(segData)
```

Similarly, to filter on between-predictor correlations, the cor function can
calculate the correlations between predictor variables:


```{r}
correlations <- cor(segData)
dim(correlations)
correlations[1:4, 1:4]

```

To visually examine the correlation structure of the data, the corrplot package
contains an excellent function of the same name. The function has many
options including one that will reorder the variables in a way that reveals
clusters of highly correlated predictors.

```{r}
corrplot::corrplot(correlations, order = "hclust")
```

To filter based on correlations, the findCorrelation function will apply the
algorithm in Sect. 3.5. For a given threshold of pairwise correlations, the function
returns column numbers denoting the predictors that are recommended
for deletion:

```{r}
highCorr <- caret::findCorrelation(correlations, cutoff = 0.75)
highCorr

filteredSegData <- segData[, -highCorr]

dim(filteredSegData)
```

### 3.8.3 Creating Dummy Variables 

