---
title: "Applied Predictive Modeling - Max Kuhn and Kjell Johnson"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


**Course Description:**
 
weblinke related with the book: http://appliedpredictivemodeling.com/

## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)

library(AppliedPredictiveModeling)
library(caret)
library(corrplot)
library(e1071)
```

# 1. Introduction 


# 2. A Short Tour of the Predictive Modeling Process

* The first step in any model building is to understand the data. 

```{r}
# Fuel Economy Data
data(FuelEconomy)
head(cars2011)

library(lattice)

# Plot shown in the text
cars2010 <- cars2010[order(cars2010$EngDispl), ]
cars2011 <- cars2011[order(cars2011$EngDispl), ]

# Add the "year"
plotData <- cars2010 %>% mutate(Year = "2010 Model Year") %>%
  bind_rows(cars2011 %>% mutate(Year = "2011 Model Year")) %>%
  mutate(Year = factor(Year))

str(plotData)

## Plot
plotData %>%
  ggplot(aes(x = EngDispl, y = FE)) +
  geom_point(alpha = 0.5) +
  labs(x = "Engine Displacement", 
       y = "Fuel Efficiency (MPG)", 
       title = "The relationship between engine displacement and fuel efficiency")  +
  facet_wrap(~ Year)
```

The relationship is somewhat linear but does exhibit some curvature towards the extreme ends of the displacement axis. 


After first understanding the data, the next step is to build and evaluate a model on the data. 


A simple model: 2010 as Training set

```{r}
model_lm <- lm(FE ~ EngDispl, data = cars2010)

summary(model_lm)

# predicted value
predicted <- predict(object = model_lm, data = cars2010)

# plot
cars2010 %>%
  mutate(predicted = predict(object = model_lm, data = .)) %>%
  ggplot(aes(x = FE, y = predicted)) +
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", size = 1, col = "red") +
  scale_x_continuous(limits = c(10, 70)) + 
  scale_y_continuous(limits = c(10, 70))

```

# 3. Data Pre-processing

## 3.1 Case Study: Cell Segmentation in High-Content Screeening


## 3.2 Data Transformations for Individual Predictors

* Centering and Scaling

* Transformations to Resolve Skewness


## 3.3 Data Transformations for Multiple Predictors 

* Transformations to Resolve Outliers

Outliers are samples that are exeptionally far from the mainstream of the data. 

* Data Reduction and Feature Extraction 


## 3.4 Dealing with Missing Values

"informative missingness", which can induce significant bias in the model. 


## 3.5 Removing Predictors

A predictor variable that has a single unique value: a Zero variance predictor. 

Some predictors might have only a handful of unique values that occure with very low frequencies: "near-zero variance predictors". 

A rule of thumb for detecting near-zero variance predictors is: 

* The fraction of unique values over the sample size is low (say 10%)

* The ratio of the frequency of the most prevalent value to the frequency of the second most prevalent value is large (say around 20)


Between-predictor correlations: 

Collinearity is the technical term for the situation where a pair of predictor variables have a substantial correltion with each other. 


## 3.6 Adding Predictors


## 3.7 Binning Predictors

## 3.8 Computing 

```{r}
# Find functions for creating a confusion matrix within the currently loaded packages
apropos("confusion")

```

```{r}
# Raw segmentation data set
data("segmentationOriginal")

# feasture names
names(segmentationOriginal)

# just focus on the training set samples
segData <- subset(x = segmentationOriginal, Case == "Train")

# The Class and Cell fields are saved into separate vectors
callID <- segData$Cell
class <- segData$Class
case <- segData$Case

# Now remove the columns
segData <- segData[, -(1:3)]


# The original data contained several "status" columns and remove them-- A good method to remove columns 
statusColNum <- grep(pattern = "Status", x = names(segData))

segData <- segData[, -statusColNum]

dim(segData)
```

### 3.8.1 Transformations

```{r}
# Calculates the sample skewness statistic for each predictor: 
e1071::skewness(segData$AngleCh1)

# Calculate the skewness across columns
skewValues <- map_df(segData, .f = e1071::skewness)

head(skewValues)

max(skewValues)
```

```{r}
# Determine which type of transformation should be used, the MASS package contains the boxcox function. Although this funciton estimates lambda, it does not creat the transformed variables. 


# BoxCox transformations
Ch1AreaTrans <- caret::BoxCoxTrans(y = segData$AreaCh1)
Ch1AreaTrans

class(Ch1AreaTrans)


# The original data
head(segData$AreaCh1)

# After transformation 
predict(object = Ch1AreaTrans, head(segData$AreaCh1))

(head(segData$AreaCh1)^(-0.9) - 1)/(-0.9)
```

```{r}

# Another caret function, preProcess
# ?caret::preProcess

# The base R funciton *prcomp* can be used for PCA

pcaObject <- prcomp(x = segData, center = TRUE, scale. = TRUE)
names(pcaObject)

# Calculate the cumulative percentage of variance which each component account  for 
percentVariance <- pcaObject$sdev^2/sum(pcaObject$sdev^2) * 100

percentVariance[1:5]


var(pcaObject$x[, 1:5]) # PC1 has the largest variance
cor(pcaObject$x[, 1:5]) # no correlation among PC1, PC2


# the transformed values are store in pcaObject as a sub-object called x: 
head(pcaObject$x[, 1:5])

# Variable loadings
head(pcaObject$rotation[, 1:5])

# spatiaSign Transformation
spatialSign(rnorm(5))

spatialSign(matrix(rnorm(12), ncol = 3))



```

To administer a series of transformations to multiple data sets, the caret
class preProcess has the ability to transform, center, scale, or impute values,
as well as apply the spatial sign transformation and feature extraction. The
function calculates the required quantities for the transformation. After calling
the preProcess function, the predict method applies the results to a set
of data. For example, to Box–Cox transform, center, and scale the data, then
execute PCA for signal extraction, the syntax would be:

```{r}
# Box-Cox transformation --> centering --> scaling --> PCA for signal extraction
trans <- preProcess(segData, method = c("BoxCox", "center", "scale", "pca"))

trans

# Apply the transformations:
transformed <- predict(trans, segData)
head(transformed)
```

The order in which the possible transformation are applied is *transformation*,
*centering*, *scaling*, *imputation*, *feature extraction*, and then *spatial sign*.
Many of the modeling functions have options to center and scale prior
to modeling. For example, when using the *train* function (discussed in later
chapters), there is an option to use *preProcess* prior to modeling within the
resampling iterations.


### 3.8.2 Filtering 

To filter for near-zero variance predictors, the *caret* package function *nearZeroVar* will return the column numbers of any predictors that fulfill the conditions.

```{r}
# add zero and near-zero variance predictors
temp <- segData
temp$zeroV <- 1

temp$nearZeroV <- c(rep(3, nrow(temp) - 10), rep(2, 10))
temp$nearZeroV2 <- c(rep(3, nrow(temp) - 50), rep(2, 25), rep(4, 25))

# find which columns have zero and near-zero variance 
near_zero_col <- nearZeroVar(temp)
near_zero_col

# remove the near-zero variance predictors
temp2 <- temp[, -near_zero_col]

# calculate how many columns were removed
ncol(temp) - ncol(temp2)
```

Similarly, to filter on between-predictor correlations, the cor function can
calculate the correlations between predictor variables:


```{r}
# correlation maxtrices
correlations <- cor(segData)
dim(correlations)

# show rows: 1:4, and cols: 1:4
correlations[1:4, 1:4]
```

To visually examine the correlation structure of the data, the corrplot package
contains an excellent function of the same name. The function has many
options including one that will reorder the variables in a way that reveals
clusters of highly correlated predictors.

```{r}
# A graphical display of a correlation matrix 
corrplot::corrplot(correlations, type = "full", order = "hclust")
```

To filter based on correlations, the findCorrelation function will apply the
algorithm in Sect. 3.5. For a given threshold of pairwise correlations, the function
returns column numbers denoting the predictors that are recommended
for deletion:

```{r}
# Determine highly correlated variables
highCorr <- caret::findCorrelation(correlations, 
                                   verbose = TRUE, 
                                   cutoff = 0.75,
                                   names = FALSE)
highCorr

filteredSegData <- segData[, -highCorr]

dim(filteredSegData)
```

### 3.8.3 Creating Dummy Variables 

When working with tree-based models, the full set is recommended. 

```{r}
# car subset data
data_test <- data.frame(x1 = 1:5, x2 = c("a", "b", "c", "c", "c"), y = 4:8)

levels(data_test$y)

# creating dummy variables based on the formula method
simpleMod <- dummyVars( ~ x1 + x2, 
                        data = data_test, 
                        # remove the variable name from the column name
                        levelsOnly = TRUE)
simpleMod

# generate the dummy variables 
predict(simpleMod, data_test)

```

```{r}
# More advance model with an interaction
withInteraction <- dummyVars(~ x1 + x2 + x1:x2, 
                             data = data_test, 
                             levelsOnly = TRUE)

withInteraction

# generate the dummy variable 
predict(withInteraction, data_test)
```

## 3.9 Exercises
### 3.9.1  Glass Samples 
```{r}
# Glass samples 
library(mlbench)
data(Glass)
head(Glass)
glimpse(Glass)
summary(Glass)

# Predictor variables distributions and relationships between predictors
Glass2 <- Glass %>% select(-Type)

GGally::ggpairs(data = Glass2)
```


# 4. Over-Fitting and Model Tuning

If models are built on small data sets, then the data quality must be sufficient and representive of the entire sample population. 

## 4.1 The Problem of Over-Fitting

## 4.2 Model Tuning

Many models have important parameters which cannot be directly estimated from the data. 

For example, in the K-nearest neighbor classification model, a new sample is predicted based on the K-closest data points in the training set. 

Model tuning parameter can not be calculated using analytical formula.  

Resampling the training set. 


## 4.3 Data Splitting

A few of the common steps in model buidling are: 

* Pre-processing the predictor data

* Estimating model parameters

* Selecting predictors for the model

* Evaluating model performance

* Fine tuning class prediction rules


When a larger amount of daa is at hand, a set of samples can be set aside to evaluate the final model. 

However, when the number of sample is not large, a strong case can be made that a test set should be avoided because every sample may be needed for model building. Additionally, the size of the test set may not have sufficient power or precision to make reasonable judgements. 

One good method: *maximum dissimilarity sampling*. 


## 4.4 Resampling Techniques

*k-Fold Cross-Validation* 

*Generalized Cross-Validatiaon*

*Repated Training/Test Splits*: is also known as "leave-group-out cross-valication" or "Monte Carlo Cross-Validation". 

*Bootstrap*: A bootstrap sample is a random sample of the data taken with replacement. Some samples will be represented multiple times in the bootstrap sample while others will not be selected at all. The samples not selected are usually referred to as the "out-of-bag" samples. 


## 4.5 Case Study: Credit Scoring 

To predict the probability that applicants have good credit. 

## 4.6 Choosing Final Tuning Parameters

## 4.7 Data Splitting Recommendations

## 4.8 Choosing Between Models

## 4.9 Computing 

### 4.9.1 Data Splitting

```{r}
# two-class data
library(AppliedPredictiveModeling)
data(twoClassData)

# summary 
summary(predictors)
summary(classes)
str(predictors)
str(classes)
```

```{r}
# Base R function sample
sample(x = 1:nrow(predictors), 
       size = round(0.8*nrow(predictors)))

# Using the createDataPartition function 

# set the random seed so the results can be reproduced. 
set.seed(1234)
trainingRows <- createDataPartition(
  # create stratified random splits 
  y = classes,
  # the percent of data for training
  p = 0.80, 
  # to generate a matrix of row numbers
  list = FALSE)

head(trainingRows)

# Subset the data into objects for training using integer sub-setting
trainPredictors <- predictors[trainingRows, ]
trainClasses <- classes[trainingRows]

# Create the test using negative integers
testPredictors <- predictors[-trainingRows, ]
testClasses <- classes[-trainingRows]

# structure of the training and the test data sets
str(trainPredictors)
str(testPredictors)

```

Generate a test set using maiximum dissimilarity sampling

```{r}
# Start with 5 data points
start <- sample(1:nrow(predictors), 5)
base <- predictors[start, ]
pool <- predictors[-start, ]

# select 35  addition 
newSampl <- maxDissim(a = base, b = pool, n = 35, randomFrac = 1, obj = minDiss)

# all sample
allSampl <- c(start, newSampl)


# Plot 
predictors[-newSampl, ] %>%
  ggplot(aes(x = PredictorA, y = PredictorB)) +
  geom_point(col = "darkgrey") + 
  geom_point(mapping = aes(x = PredictorA, y = PredictorB), data = base, shape = 16, size = 2) +
  geom_point(mapping = aes(x = PredictorA, y = PredictorB), 
             data = pool[newSampl, ], col = "darkred")

```


### 4.9.2 Resampling

Repeated training/test splits: 

```{r}
# generate 3 resampled training sets
repeatedSplits <- createDataPartition(trainClasses, 
                                       p = .80, times = 3)
str(repeatedSplits)
```

Bootstrapping: 

```{r}
createResample(trainClasses, times = 2)

```

k-fold cross-validatiaon
```{r}
# Create 10-fold cross-validation
set.seed(1)
cvSplits <- createFolds(trainClasses, 
                        k = 10, 
                        returnTrain = TRUE)
str(cvSplits)


# Get the first set of row numbers from the list
fold1 <- cvSplits[[1]]

# To Get the first 90% of the data (the first fold):
cvPredictors1 <- trainPredictors[fold1, ]
cvClasses1 <- trainClasses[fold1]
nrow(trainPredictors)
nrow(cvPredictors1)
```

### 4.9.3 Basic Model Building in R

Fit a 5-nearest neighbor classification model to the training data adn use it to predict the test set. 

*knn* from the *MASS* package

*ipredknn* from the *ipred* package

*knn3* from the *caret* package, which can produce class prediction as well as the proportion of neighbors for each class. 

```{r}
# build a 5-nearest neighbor model 
trainPredictors <- as.matrix(trainPredictors)

knnFit <- knn3(x = trainPredictors, 
               y = trainClasses, 
               # number of neighbors considered
               k = 5)

knnFit

# Predict new samples
testPredictions <- predict(knnFit, 
                           newdata = testPredictors, 
                           type = "class")
head(testPredictions)
str(testPredictions)

# accuracy 
mean(testPredictions == testClasses)
```

### 4.9.4 Determination of Tuning Parameters

* the *tune* function from the *e1071* package: can evalute four types of models across a range of parameters. 

* the *errortest* function from the *ipred* package: resample single models. 

* the *train* function from the *caret* package: has many built-in modules (144+) models, includes capabilites for different resampling methods, performances measures, and algorithms for choosing the best model from the profile. Also, this function has capabilities for parallel processing so that the resampled model fits can be executed across multiple computers or processors. 


```{r}
# Credit scoring data
data("GermanCredit")
german_credit <- GermanCredit %>%
  janitor::clean_names()

# near_zero variance
nearzero_var_index <- nearZeroVar(german_credit %>% select(-class), 
                                  freqCut = 90/10, 
                                  uniqueCut = 10)

# omit near_zero_variance
german_credit_clean <- german_credit %>% select(-nearzero_var_index)

# summary of data
head(german_credit_clean)
str(german_credit_clean)


# split data
train_index <- createDataPartition(y = german_credit_clean$class, p = 0.8, list = FALSE)
german_credit_train <- german_credit_clean[train_index, ]
german_credit_test <- german_credit_clean[-train_index, ]

```

Build SVM model:

```{r}
# build a SVM model
set.seed(1234)
svm_fit <- train(class ~ ., 
                 data = german_credit_train, 
                 # model type
                 method = "svmRadial", 
                 # pre-process the predictor data
                 preProc = c("center", "scale"), 
                 # Specify the exact cost values to investigate
                 tuneLength = 10, 
                 # Calculate performance measures
                 trControl = trainControl(method = "repeatedcv", 
                                          repeats = 5)
                 )

svm_fit

# Visualize the performance profile for the training set
plot(svm_fit, scales = list(x = list(log = 2)))

# Predict new samples 
predicted_class <- predict(svm_fit, german_credit_test)
str(predicted_class)

# Accuracy 
mean(predicted_class == german_credit_test$class)


# predict class probabilities
predicted_probs <- predict(svm_fit, 
                           newdata = german_credit_test, 
                           type = "prob")

head(predicted_probs)
```


### 4.9.5 Between-Model comparisons

```{r}
# Basic logistic regression 
set.seed(1234)
logist_reg <- train(class ~ ., 
                    data = german_credit_train, 
                    # generalized linear model
                    method = "glm", 
                    # resampling used to evaluate the performance of the model
                    trControl = trainControl(method = "repeatedcv", repeats = 5)
                    )

logist_reg
```

To compare these two models based on their cross-validatiaon statistics, the *resamples* function can be used with models that share a common set of resampled data sets. Since the random number seed was initialized prior to running the SVM and logistic models, paired accuracy measurments exist for each data set. 

```{r}
# Create a resamples object from the models
resamp <- resamples(list(SVM = svm_fit, Logistic = logist_reg))

# See the performance distributions 
summary(resamp)

# assess possible differences between the models
model_differences <- diff(resamp)

# get the p-values for the model comparisons  (p >0.05 indicates that the models fail to show any difference in performance)
summary(model_differences)


```

## 4.10 Excercise

### 4.10.1 Uncertainty of a test set

One method for understanding the uncertainty of a test set is to use a confidence interval. 

```{r}
# 16 out of 20 correct (80% accurate in small size test sample)
binom.test(x = 16, n = 20) 

# 160 out of 200 correct (80% accurate in large size test sample)
binom.test(x = 160, n = 200) 


# 1600 out of 2000 correct (80% accurate in even large size test sample)
binom.test(x = 1600, n = 2000) 
```


# 5. Measuring Performance in Regression Models

To understand the strengths and weaknesses of a particular model, relying solely on a single metric is problematic. Visualizations of the model fit, particularly residual plots, are critical to understanding whether the model is fit for purpose. 

## 5.1 Quantitative Measures of Performance

* *MSE*: Mean suqared error.  $MSE = \frac{\sum_{i = 1}^{N}(\hat{y}_i - y_i)^2}{N}$. 

* *RMSE*: the root mean squared error. $RMSE = \sqrt{\frac{\sum_{i = 1}^{N}(\hat{y}_i - y_i)^2}{N}}$.  

* *Coefficient of determination*. $R^2$ The proportion of the information in the data that is explained by the model.  It is a measure of correlation, not accuracy. 
R^2 depends on the outcome variance. For example, a model is used to model the house price (60K to 2M, note: high variance).  The model that has a $R^2= 0.9$ might be viewed as a good one, but the RMSE may be in the tens of thousands of dollars - poor predictive accuracy for anyone selling a moderately priced property. 

* *Rank correlation*: A good choice for a model that is to model biologically active new chemical compounds for pharmaceutical scientists.  The rank correlation takes the ranks of the observed outcome values and evaluates how close these are to ranks of the model predictions. To calculate this value, the ranks of the observed adn predicted outcomes are obtained and the correlation coefficient between ranks is calculated. This metric is commonly known as *Spearman's rank correlation*. 


## 5.3 The Variance-Bias Trade-off

The MSE can be decomposed into more specific pieces. Formally, the MSE of a model is:

$$MSE = \frac{\sum_{i = 1}^{n}(\hat{y}_i - y_i)^2}{n} $$

If we assume that the data points are statistically independent and that the residuals have a theoretical mean of zero and a constant variance of $\sigma^2$, then: 

$$E[MSE] = \sigma^2 + {(Model Bias)}^2 + (Model Variances)$$


## 5.4 Computing 

```{r}
# Create two vectors to illustrate the techniques
observed <- c(0.22, 0.83, -0.12, 0.89, -0.23, -1.30, -0.15, -1.4, 
              0.62, 0.99, -0.18, 0.32, 0.34, -0.30, 0.04, -0.87, 
              0.55, -1.30, -1.15, 0.20)

predicted <- c(0.24, 0.78, -0.66, 0.53, 0.70, -0.75, -0.41, -0.43, 
               0.49, 0.79, -1.19, 0.06, 0.75, -0.07, 0.43, -0.42, 
               -0.25, -0.64, -1.26, -0.07)

residual_values <- observed - predicted

# summary of residuals 
summary(residual_values)

# plot 

df <- data.frame(observed = observed, predicted = predicted, residual_values = residual_values) %>%
  pivot_long(cols = c(predicted, residual_values), names_to = "name", values_to = "value") %>%
  mutate(name = as.factor(name))

ggplot(data = df, aes(x = observed, y = value)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~name, scales = "free_y", nrow = 2)

# R2
caret::R2(predicted, observed)

# RMSE
RMSE(predicted, observed)

# Simple correlation
cor(predicted, observed)

# Rank correlation
cor(predicted, observed, method = "spearman")

```


# 6. Linear Regression and Its Cousins

Ordinary linear regression, partial least squares (PLS), and penalized models (such as ridge regression, the lasso and the elastic net)  can directly or indireectly be written in the form: 

$$y_i = b_0 + b_1x_{x1} + b_2 x_{i2} +\dots+ b_p{x_{ip} + e_i} $$

Ordinary linear regression, at one extreme, finds parameter estimates that have minimum bias, wheres ridge regression, the lasso, and the elastic net find estimates that have lower variance. 

## 6.1 Case Study: Quantitative Structure-Activity Relationship Modeling

## 6.2 Linear Regression

Ordinary least squares linear regression is to find the plane to minimize the sum-of-squared errors (SSE). 

$$SSE = \sum_{i=1}^n(y_i - \hat{y_i})^2 $$


## 6.3 Partial Least Squares

Partial Least Squares finds components that maximally summarize the variation of the predictors while simultaneously requiring these components to have maximum correlation with the response. 

PLS can be viewed as a *supervised* dimension reduction procedure; PCR is an *unsupervised* procedure. 

In practice, PCR produces models with similar predictive ability to PLS. 


## 6.4 Penalized Models

*Ridge regression* adds a penalty on the sum of the squared regression parameters:

$$SSE_{L_2} = \sum_{i = 1}^n(y_i - \hat{y_i)^2} + \lambda \sum_{j=1}^P\beta_i^2 $$

The L2 signifies that a second-order penalty. 

This method *shrinks* the estimates towards 0 as the $\lambda$ penalty becomes large (these techniques are sometimes called "shrinkage methods"). 


A popular alternative to ridge regression is the least absolute shrinkage adn selection operator model, frequently called the *lasso* (Tibshirani 1996). This model uses a similar penaly to ridge regression: 

$$SSE_{L_1} = \sum_{i=1}^n(y_i - \hat{y_i})^2 + \lambda \sum_{j=1}^{P}|{\beta_j}|$$

While this may seem like a small modification, the practical implications are significant. 


A generalization of the lasso model is the *elastic net*. 

$$SSE_{E_{net}} = \sum_{i=1}^n(y_i-\hat{y_i})^2 + \lambda_1 \sum_{j=1}^P\beta_j^2 + \lambda_2\sum_{j=1}^P|\beta_j|$$


## 6.5 Computing 

The R packages *elasticnet*, *caret*, *lars*, *MASS*, *pls*, and *stats* will be referenced. 

### 6.5.1 Solubility Data
```{r}
# solubility data
library(AppliedPredictiveModeling)
data(solubility)

# list the data objects begin with "sol": 
ls(pattern = "^solT")

# Training set
head(solTrainX)
dim(solTestX)

# test set 
head(solTestX)
dim(solTestX)

# Box-Cox transformed data frames
head(solTrainXtrans)
head(solTestXtrans)

# solubility values for each compound 
head(solTrainY)
head(solTestY)
```

### 6.5.2 Ordinary Linear Regression 

Traditional way: 

```{r}
# Creat a dataframe
training_data <- bind_cols(solTrainXtrans, Solubility = solTrainY)

# Fit all predictors (wrong ways)
lm_fit_all_predictors <- lm(Solubility ~ ., data = training_data)

summary(lm_fit_all_predictors)

# predict
lm_pred1 <- predict(lm_fit_all_predictors, solTestXtrans)
head(lm_pred1)

# collect the observed and predicted vaues into a data frame
lm_values1 <- data.frame(obs = solTestY, pred = lm_pred1)

# use the caret funtion defaultSummary to estimate the test set performance: 
defaultSummary(lm_values1)

```

Robust linear regression model: 

```{r}
# Robust linear regression model
library(MASS)
rlm_fit_all_predictors <- MASS::rlm(Solubility ~ ., data = training_data)

# Prediction on the test set
lm_pred_rlm <- predict(rlm_fit_all_predictors, newdata = solTestXtrans)

# Performance
defaultSummary(data.frame(obs = solTestY, pred = lm_pred_rlm))
```

Caret: train fuction 

```{r}
# specifies the type of resampling
ctrl <- trainControl(method = "cv", number = 10)

# Train
set.seed(100)
lm_fit1 <- train(x = solTrainXtrans, y = solTrainY, 
                 method = "lm", trControl = ctrl)

# results
lm_fit1

# Residuals check: Training set 
xyplot(solTrainY ~ predict(lm_fit1), 
       type = c("p", "g"), 
       xlab = "Predicted", ylab = "Observed")

xyplot(resid(lm_fit1) ~ predict(lm_fit1), 
       type = c("p", "g"),
       xlab = "Predicted", ylab = "Residuals")
```


Build a smaller model without predictors with extremely high correlations: 
```{r}
# filter high correlated predictors
corThresh <- 0.9
tooHigh <- findCorrelation(cor(solTrainXtrans), cutoff = corThresh)

# High correlated predictors
corrPred <- names(solTrainXtrans)[tooHigh]

# fitered data 
trainXfiltered <- solTrainXtrans[, -tooHigh]

testXfiltered <- solTestXtrans[, -tooHigh]

# build a model using the filtered data
set.seed(100)
lmFiltered <- train(x = trainXfiltered, 
                    y = solTrainY, 
                    method = "lm", 
                    trControl = ctrl)

lmFiltered

```

Robust linear regression using the *train* funciton 

```{r}
# train
set.seed(100)
rlmPCA <- train(x = solTrainXtrans, 
                y = solTrainY, 
                # robust lienar regression 
                method = "rlm", 
                # ensure that predictors are not singular, pre-process the predictors using PCA
                preProcess = "pca", 
                trControl = trainControl(method = "cv", 
                                         number = 10, 
                                         preProcOptions = list(pcaComp = 40))
                )

rlmPCA
```


### 6.5.3 Partial Least Squares

The *pls* package has functions for *PLS* and *PCR*. 

```{r}
#  Partial Least Squares and Principal Component Regression
library(pls)
pcrFit <- pcr(Solubility ~., data = training_data)
plsFit <- plsr(Solubility ~., data = training_data)

# summary of model
pcrFit
plsFit

# predict
predict(pcrFit, solTestXtrans[1:5, ], ncomp = 1:2)
```


Using the function of *train* in  the *caret*. 

```{r}
#  PLS
set.seed(100)
plsTune <- train(x = solTrainXtrans, 
                 y = solTrainY, 
                 method = "pls", 
                 # the default tuning grid evaluate components 1...tuneLength
                 tuneLength = 20, 
                 trControl = trainControl(method = "cv", number = 10 ),
                 preProcess = c("center", "scale"))

plot(plsTune)


#  PCR
set.seed(100)
pcrTune <- train(x = solTrainXtrans, 
                 y = solTrainY, 
                 method = "pcr", 
                 # the default tuning grid evaluate components 1...tuneLength
                 tuneLength = 40, 
                 trControl = trainControl(method = "cv", number = 10 ),
                 preProcess = c("center", "scale"))

plot(pcrTune)
```


### 6.5.4 Penalized Regression Models

Ridge-regression models can be creaed using the *lm.ridge* function in the *MASS* package or the *enet* function in the *elasticnet* package. 

```{r}
# When calling the enet function, the lambda argument specifies the ridge-regression penalty:
# install.packages("elasticnet")
ridgeModel <- elasticnet::enet(x = as.matrix(solTrainXtrans), 
                               y = solTrainY, 
                               # fixed the ridge penalty
                               lambda = 0.001)

# For ridge regression, we define a single lasso penalty of 0
ridgePred <- predict(ridgeModel, 
                     newx = as.matrix(solTestXtrans), 
                     s = 1, 
                     mode = "fraction", 
                     type = "fit")

head(ridgePred$fit)
```

```{r}
# to tune over the penalty
ridgeGrid <- data.frame(.lambda = seq(0, 0.1, length = 15))

set.seed(100)
ridgeRegFit <- train(x = solTrainXtrans, y = solTrainY, 
                     method = "ridge", 
                     # fit the model over many penalty values
                     tuneGrid = ridgeGrid, 
                     trControl = trainControl(method = "cv", number = 10), 
                     # put the predictors on the same scale
                     preProcess = c("center", "scale"))
                     
ridgeRegFit

```

The *lasso* model can be estimated: the *lars* function in the *lars* package; the *elasticnet* package has *enet*. 


```{r}
# fits Elastic Net regression models
library(elasticnet)
enetModel <- enet(
  # matrix of predictors
  x = as.matrix(solTrainXtrans), 
  # response
  y = solTrainY,
  # control the ridge-regression: Quadratic penalty parameter. Lambda = 0 performs the Lasso fit
  lambda = 0.01, 
                  # standardization
                  normalize = TRUE)

# returned itesm
names(enetModel)

plot(enetModel)

# The lasso penalty is specified until the time of prediction
enetPred <- predict(enetModel, 
                    newx = as.matrix(solTestXtrans), 
                    s = .1, 
                    mode = "fraction", 
                    type = "fit")
names(enetPred)

# fit values
head(enetPred$fit)

# To determin which predictors are used in the model, the predict method is used wiht type = "coefficients"

enetCoef <- predict(enetModel, newx = as.matrix(solTestXtrans), 
                     s = 0.1, mode = "fraction", 
                     type = "coefficients")

head(enetCoef$coefficients)
tail(enetCoef$coefficients)
```


Tune the elastic net model using *train*: 
```{r}
# define tune grid
enetGrid <- expand.grid(.lambda = c(0, 0.01, 0.1), 
                        .fraction = seq(0.05, 1, length = 20))

# tune the elastic net model
set.seed(100)
enetTune <- train(x = solTrainXtrans, 
                  y = solTrainY, 
                  method = "enet", 
                  tuneGrid = enetGrid, 
                  trControl = trainControl(method = "cv", number = 10), 
                  preProcess = c("center", "scale"))

plot(enetTune)

# predict
enet_predicted <- predict(object = enetTune, newdata = solTestXtrans) 
head(enet_predicted)

plot(enet_predicted, solTestY)


# collect the observed and predicted vaues into a data frame
enetTune_values1 <- data.frame(obs = solTestY, pred = enet_predicted)

# use the caret funtion defaultSummary to estimate the test set performance: 
defaultSummary(enetTune_values1)
```


# 7. Nonlinear Regression Models

Several nonlinear regression models: neural networks, multivariate adaptive regression splines (MARS), support vector machines (SVMs), and K-nearest neighbors (KNNs). 


## 7.1 Neural Networks

Each hidden unit is a linear combination of some or all the predictor variables, then this linear combination is typically transformed by a nonlinear function *g(.)*, such as the logistic (i.e., sigmoidal) function: 

$$h_x(X) = g(\beta_{0k} + \sum_{i = 1}^Px_j\beta_{jk}) $$ 

Where $g(u) = \frac{1}{1+e^{-u}}$

ANNs: the optimization prodced would try to minimize an alternative version of the sum of the squared errors: 

$$\sum_{i=1}^n(y_i - f_i(x))^2 + \lambda\sum_{k=1}^H\sum_{j=0}^P\beta_{jk}^2+\lambda\sum_{k=0}^H\gamma_k^2$$


## 7.2 Multivariate Adaptive Regression Splines (MARS)



## 7.3 Support Vector Machines (page 151)

SVMs are a class of powerful, highly fexible modeling techniques. For regression, it is *a robust regression* where the effect of outliers are minimized. Also, it is a *E-insensitive regression*. 

If the threshold is set to a relative large value, then the outliers are the only points that define the regression line! 

## 7.4 K-Nearest Neighbors

The KNN approach simply predicts a new sample using the K-closest samples from the training set. 


## 7.5 computing 

### 7.5.1 Neural Networks


```{r}
# find high correlated variables
tooHigh <- findCorrelation(cor(solTrainXtrans), cutoff = 0.75)
trainXnnet <- solTrainXtrans[, -tooHigh]
testXnnet <- solTestXtrans[, -tooHigh]

# Create a specific candidate set of models to evaluate
nnetGrid <- expand.grid(.decay = c(0, 0.01, 0.1), 
                        .size = c(1:10), 
                        # use bagging 
                        .bag = FALSE)


# train
set.seed(100)
nnetTune <- train(solTrainXtrans, 
                  solTrainY, 
                  tuneGrid = nnetGrid, 
                  method = "avNNet", 
                  trControl = trainControl(method = "cv", number = 10), 
                  # automatically standardize data 
                  preProcess = c("center", "scale"), 
                  lineout = TRUE, 
                  trace = FALSE, 
                  MaxNWts = 10 * (ncol(trainXnnet) + 1) + 10 + 1, 
                  maxit = 500)

plot(nnetTune)
# Predict 
predicted <- predict(nnetTune, newData = testXnnet)

```

### 7.5.2 Multivariate Adaptive Regression Splines (MARS)

```{r}
marsFit <- earth::earth(solTrainXtrans, solTrainY)

marsFit

# summary 
summary(marsFit)

# plot
plotmo::plotmo(marsFit)

```


```{r}
# Define the candidate models to test
marsGrid <- expand.grid(.degree = 1:2, 
                        .nprune = 2:38)

# MARS model
set.seed(100)
marsTuned <- train(solTrainXtrans, 
                   solTrainY,
                   method = "earth", 
                   # Explicitly declare the candidate models to test
                   tuneGrid = marsGrid, 
                   trControl = trainControl(method = "cv", number = 10))

# predict
head(predict(marsTuned, solTestXtrans))

# Estimate the importance of each predictor in the MARS model (*evimp*) or varImp
varImp(marsTuned)
```

### 7.5.3 Support Vector Machines

```{r}
# using the ksvm function in the kernlab package
# install.packages("kernlab")
library(kernlab)
svmFit <- ksvm(x = solTrainXtrans, 
               y = solTrainY,
               kernel = "rbfdot", 
               kpar = "automatic", 
               C = 1, 
               epsilon = 0.1)
```


```{r}
# use the "train" function
svmRTuned <- train(x = solTrainXtrans, 
                   y = solTrainY, 
                   mmethod = "svmRadial", 
                   preProcess = c("center", "scale"), 
                   tuneLength = 14, 
                   trControl = trainControl(method = "cv"))
```

### 7.5.4 K-Nearest Neighbors

The *knnreg* function in the caret package fits the KNN regression model; *train* tunes the model over K: 

```{r}
# remove a few sparse and unbalanced fingerprints first
knnDescr <- solTrainXtrans[, -nearZeroVar(solTrainXtrans)]

# KNN regression
knnTune <- train(knnDescr, 
                 solTrainY, 
                 method = "knn", 
                 # center adn scaling
                 preProcess = c("center", "scale"), 
                 tuneGrid = data.frame(.k = 1:20), 
                 trControl = trainControl(method = "cv"))

knnTune

# predict
predicted <- predict(knnTune, newdata = solTestXtrans)

# plot
plot(solTestY, predicted)

# summary
# collect the observed and predicted vaues into a data frame
knn_values1 <- data.frame(obs = solTestY, pred = predicted)

defaultSummary(knn_values1)
```


# 8. Regression Trees and Rule-Based Models
## 8.1 Basic Regression Trees 

