---
title: "Importing & Cleaning Data in R: Case Studies"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


Ref: Carchedi, Nick. https://www.datacamp.com/courses/importing-cleaning-data-in-r-case-studies. 2018.

## (I) Load Required Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(gdata)
library(stringr)
library(data.table)
```


# 1. Ticket Sales Data

## 1.1 Import and Basic Structure

```{r}

# Import sales.csv: sales
sales <- read_csv("data/sales.csv")

# View dimensions of sales
dim(sales)

# Inspect first 6 rows of sales
head(sales, n = 6)

# View column names of sales
names(sales)

# View a summary of sales
summary(sales)

# Get a glimpse of sales
glimpse(sales)

# Remove the first column of sales: sales2
sales2 <- sales[, -1]
dim(sales2)
```


## 1.2 Information not worth keeping

Many of the columns have information that's of no use to us. For example, the first four columns contain internal codes representing particular events. The last fifteen columns also aren't worth keeping; there are too many missing values to make them worthwhile.

An easy way to get rid of unnecessary columns is to create a vector containing the column indices you want to keep, then subset the data based on that vector using single bracket subsetting.

```{r}
# Define a vector of column indices: keep
keep <- 5:(ncol(sales2) - 15)

# Subset sales2 using keep: sales3
sales3 <- sales2[, keep]

```


## 1.3 Separating Columns

```{r, message=TRUE}
# Split event_date_time: sales4
head(sales3$event_date_time)

sales4 <- separate(sales3, event_date_time,
                   c("event_dt", "event_time"), sep = " ")

head(sales4$event_time)

# Split sales_ord_create_dttm: sales5

sales5 <- separate(sales4, sales_ord_create_dttm, c("ord_create_dt", "ord_create_time"), sep = " ")


# Define an issues vector
issues <- c(2516, 3863, 4082, 4183)

# Print values of sales_ord_create_dttm at these indices
sales3$sales_ord_create_dttm[issues]

# Print a well-behaved value of sales_ord_create_dttm
sales3$sales_ord_create_dttm[2517]



```

```{r}
# Find columns of sales5 containing "dt": date_cols
date_cols <- str_detect(names(sales5), "dt")

# Load lubridate
library(lubridate)

# Coerce date columns into Date objects
sales5[, date_cols] <- lapply(sales5[, date_cols], ymd)


# Create logical vectors indicating missing values 
missing <- lapply(sales5[, date_cols], is.na)

# Create a numerical vector that counts missing values: num_missing
num_missing <- sapply(missing, sum)

# Print num_missing
num_missing


# Combine the venue_city and venue_state columns
sales6 <- unite(data = sales5, col = venue_city_state, venue_city, venue_state, sep = ", ", remove = TRUE)


# View the head of sales6
head(sales6)
```


# 2. MBTA Ridership Data

The Massachusetts Bay Transportation Authority ("MBTA" or just "the T" for short) manages America's oldest subway, as well as Greater Boston's commuter rail, ferry, and bus systems.

## 2.1 read excel 
```{r}

# Import mbta.xlsx and skip first row: mbta
mbta <- read_excel(path = "data/mbta.xlsx", skip = 1)

# View the structure of mbta
str(mbta)

# View the first 6 rows of mbta
head(mbta, n = 7)

# View a summary of mbta
summary(mbta)
```

It is strange how the rows and columns are organized. 


## 2.2 gather 
```{r}
# Remove rows 1, 7, and 11 of mbta: mbta2
mbta2 <- mbta[-c(1,7, 11), ]

# Remove the first column of mbta2: mbta3
mbta3 <- mbta2[, -1]
head(mbta3)

# Gather columns of mbta3: mbta4
mbta4 <- gather(mbta3, key = "month", value = "thou_riders", -mode)

# View the head of mbta4
head(mbta4)

# Coerce thou_riders to numeric
mbta4$thou_riders <- as.numeric(mbta4$thou_riders)

```


## 2.3 Separate

Get variables into columns 

```{r}
# Spread the contents of mbta4: mbta5
mbta5 <- spread(mbta4, key = "mode", value = "thou_riders")

# View the head of mbta5
head(mbta5)

# Split month column into month and year: mbta6
mbta6 <- separate(mbta5, month, c("year", "month"),  sep ="-")

# View the head of mbta6
head(mbta6)

```

## 2.4 Explore Data Analysis

```{r}
# View a summary of mbta6
summary(mbta6)

# Generate a histogram of Boat column
hist(mbta6$Boat)

# Look at all T ridership over time (don't change)
# ggplot(mbta_all, aes(x = month, y = thou_riders, col = mode)) + geom_point() + 
#   scale_x_discrete(name = "Month", breaks = c(200701, 200801, 200901, 201001, 201101)) +  
#   scale_y_continuous(name = "Avg Weekday Ridership (thousands)")
```


# 3. World Food Facts

This chapter will import and clean some data about food products from around the world.  The sugar content of a sample of foods from around the world. 

## 3.1 Importing the data
```{r}
# Import food.csv as a data frame
food <- fread("data/food.csv", data.table = FALSE)   # fread() can quickly import CSV files, no matter how large

# View summary of food
summary(food)

# Veiw head of food
head(food)

# view structure of food
str(food)

```

This dataset is so large and has so many variables that even *str*, *head*, and *summary* outputs seemed pretty intimidating!

## 3.2 Inspecting variables
```{r}
# View a glimpse of food
glimpse(food)

# View column names of food
names(food)
```


## 3.3 Removing duplicate info and finding columns

* col 1-9: what and when information was added

* col 10:17, 22:27: meta information about food

* col 18:21, 28:34: where it came from 

* col 35:52: what it's make of

* col 53:54: nutrition grades

* col 55:63: some unclear

* col 64:159: some nutritional information



```{r}
# Create vector of column indices: nutrition
nutrition <- str_detect(names(food), pattern = ".*100g$") 

# View a summary of nutrition columns
summary(food[, nutrition])

# Find indices of sugar NA values: missing
missing <- is.na(food$sugars_100g)

# Replace NA values with 0 
food$sugars_100g[missing] <- 0

# remove negative and 0 
food2 <- food[food$sugars_100g >0, ]

# Create a histogram
hist(food2$sugars_100g, breaks = 100)

# Find entries containing "plasti": plastic
plastic <- str_detect(food$packaging, pattern = "plasti")

# print the sum of plastic
sum(plastic)
```





















# 4. School Attendance Data
Use all of the tools you've learned to import and clean a gnarly dataset containing information on average school attendance in the US. The attendance data from public schools in the US were organized by school level and state, during the 2007-2008 academic year. The data contain information on average daily attendance (ADA) as a percentage of total enrollment, school day length, and school year length. 

## 4.1 Read and inspect data
```{r}
# Load the gdata package
library(gdata)

# Import the spreadsheet: att
att <- read.xls("data/attendance.xls", perl = "c:/Perl64/bin/perl.exe")

# Print the column names
names(att)

# Print the first 6 rows
head(att, n = 6)

# print the last 6 rows
tail(att, n = 6)

# print the structure
str(att)
```

Data are really messy. The column names are mostly missing, irrelevant notes are at the end of the data frame, and the numeric data were imported as factors. 

## 4.2 Removing unnecessary rows, columns

```{r}
# Create remove
remove <- c(3, 56:59)
remove

# Create att2
att2 <- att[-remove, ]

# create remove (unnecessary columns)
un_col <- seq(3, 17, by = 2)

# create att3 (removing unnecessary columns)
att3 <- att2[, -un_col]
```

## 4.3 Splitting the data into: elementary, secondary, and all schools
```{r}
# subset just elementary schools: att_elem
att_elem <- att3[, c(1, 6, 7)]

# subset just secondary schools: att_sec
att_sec <- att3[, c(1, 8:9)]

# subset all schools: att4
att4 <- att3[, 1:5]
```

## 4.4 Replacing the names
```{r}
# Define cnames vector
cnames <- c("state", "avg_attend_pct", "avg_hr_per_day", "avg_day_per_yr", "avg_hr_per_yr")

# Assign column names of att4
colnames(att4) <- cnames

# Remove first two rows of att4
att5 <- att4[-c(1:2), ]

# view the anmes of att5
names(att5)
```


## 4.5 Cleaning up extra characters
```{r}
# Remove all periods in state column
att5$state <- str_replace_all(att5$state, pattern = "\\.", replacement = "")

# Remove white space around state names
att5$state <- str_trim(att5$state)

# View the head of att5
head(att5)
```

## 4.6 Type conversions
```{r}
att5 <- att5 %>%
  mutate_at(.vars = vars(-state), funs(as.numeric))

head(att5)
```

