---
title: "Top 50 R Interview Questions and Answers"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

Ref: Bharani, _"Top 50 R Interview Questions You Must Prepare in 2018"_. https://www.edureka.co/blog/interview-questions/r-interview-questions/. 

Note: Some questions and answers have revised by Jeff Li on 12/20/2018

# (I) Load required libraries
```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

# 0: What is R? 

A programming languange, can be used as statistical analysis, data visulization, data manipulation, predictive modeling, forecase analysis, etc.


# 1: What are the different data strutures in R? 
```{r}
# 1.1 Vector: A vector is a sequence of data elements of the same basic type
a <- 1:10  # integer 
b <- letters[1:10] # character
c <- c(TRUE, FALSE, TRUE) 
map_lgl(.x = list(a, b, c), is.vector)

# 1.2 List: contain elements of different types: numbers, strings, vectors or another list inside it
my_list <- list(num = 1, string = "I love R", vector = 3:5, list(a = 3:7, b = c("good", "better", "best")))
is.list(my_list)

# 1.3 Matrix: two dimentional data structure, must be the same type
A <- matrix(1:4, nrow = 2)
is.matrix(A)

B <- matrix(letters[1:6], nrow = 3)
B
is.matrix(B)

# 1.4 Dataframe: more generic than a matrix, different columns can have different data types
DF <- data.frame(Name = c("John", "Tom", "Pat"), 
                 age = c(25, 40, 10))
is.data.frame(DF)
```


# 2. How can you loada .csv file in R? 

There are many options to read .csv files. My top two choises are: read_cvs() and read.csv() with specify the path of the file. 


# 3. What are the different components of grammar of graphics? 

Grammar of graphics have following common components: 

* Data layer

* Aesthetics layer

* Geometry layer

* Facet layer

* Co-ordinate layer

* Themes layer


# 4. What is Rmarkdown? What is the use of it? 
RMarkdwon is a reporting tool provided by R. We can create high quality report of our code using RMarkdown. The common output formats include: HTML, PDF, Word. 


# 5. How do you install a package in R? 

install.packages("xxxx")


# 6. What are the step to build and evaluate a linear regression model in R? 




