---
title: "Statistical Quality Control Book - Montgomery"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee



## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(pdftools)
```



# 6. control Charts for Variables 

The $\bar{x}$ and R control charts are widely used to monitor the mean and variability of variables. 


## 6.1 Introduction

When dealing with a quality characteristic that is a variable, it is usually necessay to monitor both the mean and variabilit. 

```{r}
# Define upper- and lower- specification limits
upper_specification_limit <- 13
lower_specification_limit <- 7

# simulated data 
set.seed(123)
prod_x <- data.frame(x = rnorm(n = 10000, mean = 10, sd = 1))

# plot
prod_x %>%
  ggplot(aes(x = x)) +
  geom_histogram(bins = 100) + 
  geom_histogram(aes(x = rnorm(n = 10000, mean = 12, sd = 1), alpha = 0.7), bins = 100) +
   geom_histogram(aes(x = rnorm(n = 10000, mean = 10, sd = 2), alpha = 0.7), bins = 100) +
  geom_vline(xintercept = c(upper_specification_limit, lower_specification_limit, 10, 12))

```


## 6.2 Control Charts for X_bar and R 

$\bar{x}$


```{r}
sim <- NULL

for (i in seq(1:1e9)) {
  sim[i] = diff(range(rnorm(n = 5), mean = 0, sd = 1))
}

mean(sim)

```
