---
title: "Programming with purr "
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---



**Course Description**

"Lists can be difficult to both understand and manipulate, but they can pack a ton of information and are very powerful. In this course, you will learn to easily extract, summarize, and manipulate lists and how to export the data to your desired object, be it another list, a vector, or even something else! Throughout the course, you will work with the purrr package and a variety of datasets from the repurrrsive package, including data from Star Wars and Wes Anderson films and data collected about GitHub users and GitHub repos. Following this course, your list skills will be purrrfect!" 



Ref: Auriel Fournier, Foundations of Functional Programming with purrr. https://www.datacamp.com/courses/foundations-of-functional-programming-with-purrr. Accessed on 2018. 


Note: Some course materials and data have been digested and adapted for my teaching. 



# (I) Load Required Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# (b) Load libraries
library(tidyverse)
library(repurrrsive)
```


# 1. Simplifying Iteration and Lists With purrr

## 1.1 The power of iteration


*map(object, function):* 

* object --> vector or list

* function -->  any function in R that takes the input offered by the object


```{r}
# Initialize list
all_files <- list()

files <- list.files("data/", pattern = ".*.csv$", full.names = TRUE)

# For loop to read files into a list
for (i in seq_along(files)) {
  all_files[[i]] <- read_csv(file = files[[i]])
}

# Output size of list object
length(all_files)
```

## 1.2 Iteration with purrr

```{r}
# use map to iterate
all_files_purrr <- map(files, read_csv)

# output size of list object
length(all_files_purrr)

```


## 1.3 More iteration with for loops
```{r}

# change each character element to a number
list_of_df <- list(a = c("1", "2", "3", "4"), 
                   b = c("5", "6", "7", "8"))

class(list_of_df[[1]])


# use for loop: change all characters to numberic
for (i in seq_along(list_of_df)) {
  list_of_df[[i]] = as.numeric(list_of_df[[i]])
}


# easier way to use map
map(.x = list_of_df, .f = as.numeric)

map(.x = list_of_df, .f = function(x) as.numeric(x)*2)
```


## 1.4 Subsetting lists 

```{r}
# Load repurrrsive package, to get access to the wesanderson dataset
#install.packages("repurrrsive")
library(repurrrsive)

# Load wesanderson dataset
data(wesanderson)

# Get structure of first element in wesanderson
str(wesanderson)

# Get structure of GrandBudapest element in wesanderson
str(wesanderson$GrandBudapest)


# Third element of the first wesanderson vector
wesanderson[[1]][3]

# Fourth element of the GrandBudapest wesanderson vector
wesanderson$GrandBudapest[4]

```

## 1.4 map_dbl, map_lgl, map_chr

*map(list, ~function(.x))*: use the argument .x to denote where teh list element goes inside the function. When we want to use .x to show where the element goes in the function, we need to put a ~ in front of the function in front of the function in the second argument of *map()*. 


```{r}
# Map over list to get the length of each element
class(wesanderson)
summary(wesanderson)

map(.x = wesanderson, ~length(.x))

# Create a numcolors column and fill with length of each wesanderson element
data.frame(numcolors = map_dbl(wesanderson, ~length(.x))) # the dbl means a double or a number that can have a decimal. 
```


# 2. More Complex Iterations

purrr is much more than a for loop; it works well with pipes, we can use it to run models and simulate data, and make nested loops!

## 2.1 Map with pipe 

```{r}
# Use pipes to check for names in sw_films
sw_films %>%
  names()

# Set names so each element of the list is named for the film title
sw_films_named <- sw_films %>% 
  set_names(map_chr(sw_films, "title"))


# Check to see if the names worked/are correct
names(sw_films_named)
```

## 2.2 Pipe in map()

We can also use pipes on the inside of map() function to help us iterate a pipeline of tasks over a list of inputs. 

Using pipes inside of *map()* makes iterating over multiple funcitons easy. 

```{r}
# Create a list of values from 1 through 10
numlist <- list(1:10)

# Iterate over the numlist 
map(numlist, ~.x %>% sqrt() %>% sin()) 

```


## 2.3 simulate data using map 
```{r}
# create a list of means
list_of_mean <- list(5, 2, 300, 15)

# Create a list of norm random value
list_of_df <- map(.x = list_of_mean, 
                  ~data.frame(a = rnorm(mean = .x,
                                        n = 200, 
                                        sd = 5/2)))

map(list_of_df, colMeans)
map(list_of_df, summary)

list_of_df %>%
  map(summary)
```



```{r simulating_data_with_purrr}
# List of sites north, east, and west
sites <- list("north", "east", "west")

# Create a list of dataframes, each with a years, a, and b column 
list_of_df <-  map(sites, 
                   ~data.frame(sites = .x,
                               a = rnorm(mean = 5, n = 200, sd = 5/2),
                               b = rnorm(mean = 200, n = 200, sd = 15)))

list_of_df
```


## 2.4 Run linear models

The following code shows how easy to run multiple models adn summarize their results. 

```{r Run_a_linear_model}
# Map over the models to look at the relationship of a vs b
str(list_of_df)

list_of_df %>%
    map(~ lm(a ~ b, data =.)) %>%
    map(summary)

```

```{r}
mtcars %>%
  group_by(am) %>%
  nest() %>%
  mutate(model = map(data, ~lm(mpg ~ wt, data = .x))) 

```

## 2.5  map_chr(), map_lgl(), map_int(), and map_dbl() 

* map(): outputs a list

* map_chr(): character vector

* map_lgl(): logical vector

* map_int(): integer vector

* map_dbl(): double vector


```{r map_chr}
# Pull out the director element of sw_films in a list and character vector
map(sw_films, ~.x[["director"]])
map_chr(sw_films, ~.x[["director"]])


# Compare outputs when checking if director is George Lucas
map(sw_films, ~.x[["director"]] == "George Lucas")
map_lgl(sw_films, ~.x[["director"]] == "George Lucas")


# Pull out episode_id element as list
map(sw_films, ~.x[["episode_id"]])

# Pull out episode_id element as double vector
map_dbl(sw_films, ~.x[["episode_id"]])

# Pull out episode_id element as integer vector
map_int(sw_films, ~.x[["episode_id"]])

```


## 2.3. map2() and pmap()

### 2.3.1 More complex interactions 

```{r}
list_of_mean <- list(5, 2, 300, 15)
list_of_sd  <- list(0.5, 0.01, 20, 1)

sim_data <- map2(.x = list_of_mean, 
                 .y = list_of_sd, 
                 ~data.frame( a = rnorm(mean = .x, 
                                        n = 200, 
                                        sd = .y), 
                              b = rnorm(mean = 200, 
                                        n = 200, 
                                        sd = 15)))

summary(sim_data)
str(sim_data)
```

```{r, pmap}
# create list of inputs
input_list <- list(
  means = list(5, 2), 
  sd  = list(0.5, 0.01), 
  n = list(200, 50)
)

str(input_list)

# using pmap()

sim_data <- pmap(input_list, 
                 function(means, sd, n) {
                   data.frame(a = rnorm(mean = means, 
                                        sd = sd, 
                                        n = n))} )

str(sim_data)

```

### 2.3.2 Simulating data with multiple inputs using map2()

```{r}
# List of 1 through 3
means <- list(1, 2, 3)

# Create sites list
sites <- list("north", "west", "east")

# Map over two arguments: sites and mu
list_of_files_map2 <- map2(.x = sites, 
                           .y = means,  
                           ~data.frame(sites = .x,
                                       a = rnorm(mean = .y, 
                                                 n = 200, 
                                                 sd = (5/2))))

str(list_of_files_map2)
```

### 2.3.3 Simulating data 3+ inputs with pmap()

```{r pmap}

# Create a master list, a list of lists
sites <- list("north", "west", "east")
means1 <- list(1, 2, 3)
means2 <- list(0.5, 1, 1.5)
sigma1 <- list(1, 2, 3)
sigma2 <- list(0.5, 1, 1.5)


pmapinputs <- list(sites = sites,  
                   means = means1, 
                   sigma = sigma1, 
                   means2 = means2, 
                   sigma2 = sigma2)

# Map over the master list
list_of_files_pmap <- pmap(pmapinputs, 
                           function(sites, means1, sigma1, means2, sigma2) 
                             data.frame(sites = sites,
                                        a = rnorm(mean = means1, n = 200, sd = sigma1),
                                        b = rnorm(mean = means2, n = 200, sd = sigma2)))
                           
str(list_of_files_pmap)


```


# 3. Troubleshooting lists with purrr

How to purrr safely()? 

If you map() over a list, and one of the elements does not have the right data type, you will not get the output you expect. 

If you have a very large list, figuring out where things went wrong, and what exactly went wrong can be hard. That is where safely() comes in; it shows you both your results and where the errors occurred in your map() call.

```{r}
# no error
list(5, 10) %>%
  map(function(x) x*10 )

# error
a <- list("unknown", 10) %>%
  map(function(x) x*10)

# using safely
a <- list("unknown", 10) %>%
  map(safely(function(x) x * 10, 
             otherwise = NA_real_))

a

# reordering to easy spot errors
a %>%
  transpose()
```

## 3.1 Using safely() in map()

```{r}

# Map safely over log
a <- list(-10, 1, 10, 0) %>%
      map(safely(log, otherwise = NA_real_)) %>%
  # Transpose the result
      transpose() 

# Print the list
a

# Print the result element in the list
a[["result"]]

# Print the error element in the list
a[["error"]]

```
Now we have the power to start debugging your lists, and you can do it with simple element subsetting. 


## 3.2 convert data to numeric with purrr

Use *purrr* for data wrangling to help clean numeric data in lists

```{r}
# Load sw_people data
data(sw_people)
# str(sw_people)

# Map over sw_people and pull out the height element
height_cm <- map(sw_people, "height") %>%
  map(function(x) {
    ifelse(x == "unknown", NA, as.numeric(x))
  })

str(height_cm)

```


## 3.3 Finding the problem areas

When you are working with a small list, it might not seem like a lot of work to go through things manually and figure out what element has an issue. But if you have a list with hundreds or thousands of elements, you want to automate that process.

Now you'll look at a situation with a larger list, where you can see how the error message can be useful to check through the entire list for issues.

```{r}
# Map over sw_people and pull out the height element
height_ft <- map(sw_people, "height") %>% 
  map(as.numeric) %>%
  map(safely(function(x) {
    x * 0.0328084}, 
    quiet = FALSE)) %>%
  transpose()

# Print the list, the result element, and the error element
height_ft

height_ft[["result"]]
height_ft[["error"]]
```


## 3.4 possible()

```{r}
a <- list(-10, "unknown", 10) %>%
  map(possibly(function(x)
    x * 10, 
    otherwise = NA_real_))

a
```


## 3.5 Replace safely() with possibly() 

Once you have figured out how to solve an issue with safely(), (e.g., output an NA in place of an error), swap out safely() with possibly(). possibly() will run through your code and implement your desired changes without printing out the error messages.

You'll now map() over log() again, but you will use possibly() instead of safely() since you already know how to resolve your errors.

```{r}
# Take the log of each element in the list
a <- list(-10, 1, 10, 0) %>%
  map(possibly(function(x) {
    log(x)
  }, NA_real_))


```

## 3.6 Convert values with possibly()
Let's say you need to convert the Star Wars character heights in sw_people from centimeters to feet. You already know that some of the heights have missing data, so you will use possibly() to convert missing values into NA. Then you will multiply each of the existing values by 0.0328084 to convert them from centimeters into feet.

To get a feel for your data, print out height_cm in the console to check out the heights in centimeters.

```{r}
# map over height-cm data
height_cm <- map(sw_people, "height") %>% map(as.numeric)
head(height_cm)


# Create a pipe workflow that returns double vectors
height_cm %>%
  map_dbl(possibly(function(x) {
    as.numeric(x) * 0.0328084
  }, NA_real_))
```

## 3.7 purrr is a wal() in the park

why walk()?

```{r}
short_list <- list(-10, 1, 10)
short_list

# clean way
walk(short_list, print)

```

## 3.8  Comparing walk() vs no walk() outputs

Printing out lists with map() shows a lot of bracketed text in the console, which can be useful for understanding their structure, but this information is usually not important for communicating with your end users. If you need to print, using walk() prints out lists in a more compact and human-readable way, without all those brackets. walk() is also great for printing out plots without printing anything to the console.

Here, you'll be using the people_by_film dataset, which dataset derived from sw_films that has the url of each character and the film they appear in.
```{r}
# 
```


Now you will try one more use of walk(), specifically creating plots using walk()

```{r}
# Load the gap_split dat
data("gap_split")
# str(gap_split)

# Map over the first 10 elements of gap_split
plots <- map2(.x = gap_split[1:10], 
              .y = names(gap_split[1:10]), 
              ~ ggplot(data = .x, aes(x = year, y = lifeExp)) +
                geom_line() + 
                labs(title = .y)
              )

# Object name, then function name
walk(plots, print)
```




# 4. Problem solving with purrr

## 4.1 Using purrr in your workflow 

```{r}
data("sw_films")
str(sw_films)
names(sw_films)

sw_films <- sw_films %>%
  set_names(map_chr(.x = sw_films, "title"))

names(sw_films)

# Setting names while asking questions
sw_films[[1]][["episode_id"]]

map_chr(sw_films, ~.x[["episode_id"]]) %>%
  set_names(map_chr(sw_films, "title")) %>%
  sort()

```


### 4.1.1 Name review 

```{r}
# Load the data from the repurrsive package
data(gh_users)
str(gh_users)

# chekc if data have names
names(gh_users)

# Map over name element of list
gh_users[[1]][["name"]]

map(gh_users, ~.x[["name"]])

```

### 4.1.2 Setting names
```{r}
set_names(1:4, c("a", "b", "c", "d"))
# If the second argument is ommitted a vector is named with itself
set_names(letters[1:5])

# Name gh_users with the names of the users
gh_users_named <- gh_users %>%
  set_names(map_chr(gh_users, "name"))

# check gh_repos
# str(gh_repos)
class(gh_repos_names)
summary(gh_repos_names)

# Name gh_repos with the names of the repo owner
gh_repos_names <- gh_repos %>%
  map_chr(~.[[1]]$owner$login) %>%
  set_names(gh_repos, .)
```

## 4.1.3 Asking questions from a list

We can use functions to ask any question of the data in just a few lines of code.

```{r}
# Determine who joined githbu first
data("gh_users")
head(gh_users)
map_chr(gh_users, ~.x[["created_at"]]) %>%
  set_names(map_chr(gh_users, "name")) %>%
  sort()

# Determine user versus organization
gh_users[[1]][["type"]]
map_lgl(gh_users, ~.x[["type"]] == "User")

# Determine who has the most publica repositories
map_int(gh_users, ~.x[["public_repos"]]) %>%
  set_names(map_chr(gh_users, "name")) %>%
  sort()

```

## 4.2 More complex list naming 

sometimes each element of a list is another list. This can make setting the names difficult and requires to use multiple map() functions. 

```{r}
# each element is a different GitHub repository
set_names(gh_repos[[1]], map(gh_repos[[1]], ~.x$name))

gh_repos <- gh_repos %>%
  map(~set_names(.x, map(.x, ~.x$name)))

# check to make sure list has the right names
names(gh_repos)
```


## 4.3 Questions about gh_repos

* Which repository is the largest? 

```{r}
# Map over gh_repos to generate numeric output
map(gh_repos, 
    ~map_dbl(.x, ~.x[["size"]])) %>%
  # grab the largest element
  map(~max(.x))
      
```

## 4.4 Graphs in purrr

ggplot() requires dataframe input. 

```{r}
# Create a dataframe with four columns

gh_users[[1]][c("login", "name", "followers", "public_repos")]

map_df(.x = gh_users, 
       .f = `[`, 
       c("login", "name", "followers", "public_repos")) %>%
       # Plot followers by public_repos
    ggplot(., 
         aes(x = followers, y = public_repos)) + 
        # Create scatter plots
        geom_point()
```

## 4.5 purrr and histograms

Now you're going to put together everything you've learned, starting with two different lists, which will be turned into a faceted histogram. You're going to work again with the Stars Wars data from the sw_films and sw_people datasets to answer a question:

What is the distribution of heights of characters in each of the Star Wars films?
Different movies take place on different sets of planets, so you might expect to see different distributions of heights from the characters. Your first task is to transform the two datasets into dataframes since ggplot() requires a dataframe input. Then you will join them together, and plot the result, a histogram with a different facet, or subplot, for each film.

```{r}
str(sw_films)

# Create a dataframe with the "title" of each film, and the "characters" from each film in the sw_films dataset.

sw_films[[1]]$title
sw_films[[1]]$characters

film_by_character <- tibble(filmtitle = map_chr(sw_films, "title")) %>%
  mutate(filmtitle, characters = map(sw_films, "characters")) %>%
  unnest()
  
# Create a dataframe with the "height", "mass", "name", and "url" elements from sw_people.
sw_people[[1]][c("height", "mass", "url")]

sw_charactes <- map_df(.x = sw_people, 
                         .f = `[`, 
                         c("height", "mass", "url") )

# Join the two dataframes together using the "characters" and "url" keys.
character_data <- inner_join(film_by_character, 
                             sw_charactes, 
                             by = c("characters" = "url")) %>%
  # make sure the columns are numbers
  mutate(height = as.numeric(height), 
         mass = as.numeric(mass))

# Plot the heights, faceted by film tilte
ggplot(character_data, aes(x = height)) +
  geom_histogram(stat = "count") + 
  facet_wrap(~filmtitle)
```

