---
title: "Exploratory Data Analysis"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


**Course Description**:

Learn how to use graphical and numerical techniques to uncover the structure of data represented as a table or a database. 

* Suggest interesting relationships

* Find unusual observations


Ref: Andrew Bray. https://www.datacamp.com/courses/exploratory-data-analysis, 2018.


*Note: I modified this course for my company's data training, so the materials are not exactly the same as those on the datacamp.* 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (I) Load Required Libraries
```{r, message = FALSE}
library(tidyverse)
```



# 1. Exploring Categorical Data

## 1.1 Contingency Table 

A contingency table is a useful way to represent the total counts of observations that fall into each combination of the levels of categorical variables. 

```{r}
# (a) read data
comics <- read_csv("data/comics.csv") %>%
  mutate_if(is.character, factor)

# Print the first rows of the data
comics

# Check levels of align
levels(comics$align)

# Check the levels of gender
levels(comics$gender)

# Create a 2-way contingency table
tab <- table(comics$align, comics$gender)
tab
```


## 1.2 Dropping levels

As shown in the above table, the contingency table shows the level of "Reformed Criminals" in "align" has just 3 counts. It will be helpful to drop such levels (easy way). 

```{r}
# Remove align level
comics_filtered <- comics %>%
  filter(align != "Reformed Criminals") %>%
  droplevels()

# See the result
table(comics_filtered$align, comics_filtered$gender)
```


## 1.3 Side-by-Side Barcharts

Side-by-side barchars are often more useful to represent data graphically. 

```{r}

# Create side-by-side barchart of gender by alignment
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "dodge", alpha = 0.7)

# Create side-by-side barchart of alignment by gender
ggplot(comics, aes(x = gender, fill = align)) + 
  geom_bar(position = "dodge", alpha = 0.7) +
  theme(axis.text.x = element_text(angle = 90))
```

As shown above two figures, in general, gender and alignment have an association. More male characters than female characters are in the dataset. 


## 1.4 Conditional proportions

```{r}

# print fewer digits
options(scipen = 999, digits = 3)

# Joint propotions
prop.table(tab)

sum(prop.table(tab)) # = 1


# conditional on columns
prop.table(tab, 2) 
prop.table(tab, 2) %>%
  colSums() 

# conditional on rows
prop.table(tab, 1)
prop.table(tab, 1) %>%
  rowSums()

```


## 1.5 counts or proportions

```{r}
# Plot of gender by align
ggplot(comics, aes(x = align, fill = gender)) +
  geom_bar()
  
# Plot proportion of gender, conditional on align
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "fill") +       # tell the bars to fill the entire height
  ylab("proportion")
```

## 1.6 Marginal barchart and conditional barchart

Marginal barchart: A barchar for just single variable. 

```{r}
# Change the order of the levels in align
comics$align <- factor(comics$align, 
                       levels = c("Bad", "Neutral", "Good"))

# Create plot of align
ggplot(comics, aes(x = align)) + 
  geom_bar()
```



Let's facet the plot of alignment based on gender

```{r}
# Plot of alignment broken down by gender
ggplot(comics, aes(x = align)) + 
  geom_bar() +
  facet_wrap(~ gender, nrow = 1) +
  theme(axis.text.x = element_text(angle = 45))
```

## 1.7 Piechart

I am not a fun of piecharts, because they can be difficult to compare similar areas. But you will see lots of piecharts in movements' reports.

```{r}
# (a) read data
pies <- read.table("data/pies.txt") %>%
  select(flavor = V1) %>%
  mutate(flavor = factor(flavor, levels = c("apple", "key-lime", "boston-creme", "blueberry", "cherry", "pumpkin", "strawberry"))) %>%
  as_tibble()

dim(pies)
names(pies)
class(pies)


# Create barchart of flavor
p1 <- ggplot(pies, aes(x = flavor)) + 
  geom_bar(fill = "chartreuse") + 
  theme(axis.text.x = element_text(angle = 90))

p1 

# Create piechar of flavor

ggplot(pies, aes(x = flavor)) + 
  geom_bar() +
  coord_polar()
```


# 2. Exploring Numerical Data

## 2.1 Structure of Data 

```{r}
# (a) read data
cars <- read_csv("data/cars04.csv")
dim(cars)
names(cars)
summary(cars)
```

## 2.2 Histogram 
```{r}
# Create faceted histogram
ggplot(cars, aes(x = city_mpg)) +
  geom_histogram() +
  facet_wrap(~ suv)
```


## 2.3 Boxplots and Density Plots

```{r}
# Possible levesl of 
table(cars$ncyl)

# Filter cars with 4, 6, 8 cylinders
common_cyl <- filter(cars, ncyl %in% c(4,6,8))

# Create box plots of city mpg by ncyl
ggplot(common_cyl, aes(x = as.factor(ncyl), y = city_mpg)) +
  geom_boxplot()

# Create overlaid density plots for same data
ggplot(common_cyl, aes(x = city_mpg, fill = as.factor(ncyl))) +
  geom_density(alpha = .3)
```

As shown in the density plot, (1) The highest mileage cars have 4 cylinders; (2) The typical 4 cylinder car gets better mileage than the typical 6 cylinder car, which gets better mileage than the typical 8 cylinder car; (3) Most of the 4 cylinder cars get better mileage than even the most efficient 8 cylinder cars; (4) the variability in mileage of 4 cylinder cars is much greater than that of 8 cylinder cars. 



## 2.4 Marginal and Conditional Histograms

```{r}
# Create hist of horsepwr
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  ggtitle("Distribution of Horsepower")

# Create hist of horsepwr for affordable cars
cars %>% 
  filter(msrp < 25000) %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  xlim(c(90, 550)) +
  ggtitle("Distribution of Horsepower for cars under $25,000")
```

Cars with around 200 horsepower are more common than cars with around 200 horsepower.  



## 2.5 Binwidths

The binwidth determines how smooth the distribution will appear: the smaller the binwidth, the more jagged the distribution becomes. 


```{r}
# Create hist of horsepwr with binwidth of 3
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram(binwidth = 3) +
  ggtitle("Distribution of Horsepower with binwidth =3")

# Create hist of horsepwr with binwidth of 30
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram(binwidth = 30) +
  ggtitle("Distribution of Horsepower with binwidth =30")

# Create hist of horsepwr with binwidth of 60
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram(binwidth = 60) +
  ggtitle("Distribution of Horsepower with binwidth =60")

```

We can spot a second model around 300 horsepower. 

With binwidth = 3, we can count cars with 200 and 300 horsepower. 


d## 2.6 Box plots 

*Boxplots* shows the center and spread of a distribution, and also provides a graphical means to detect outliers. 

```{r}
# Construct box plot of msrp
cars %>%
  ggplot(aes(x = 1, y = msrp)) +
  geom_boxplot()

# Exclude outliers from data
cars_no_out <- cars %>%
  filter(msrp < 100000)

# Construct box plot of msrp using the reduced dataset
cars_no_out %>%
  ggplot(aes(x = 1, y = msrp)) +
  geom_boxplot()


# Create plot of city_mpg
cars %>%
  ggplot(aes(x = 1, y = city_mpg)) +
  geom_boxplot()

cars %>%
  ggplot(aes(x = city_mpg)) +
  geom_density()

# Create plot of width
cars %>% 
  ggplot(aes(x = 1, y = width)) +
  geom_boxplot()

cars %>% 
  ggplot(aes(x =  width)) +
  geom_density()

```


## 2.7 Visualization in higher dimensions

```{r}
# Facet hists using hwy mileage and ncyl
common_cyl %>%
  ggplot(aes(x = hwy_mpg)) +
  geom_histogram() +
  facet_grid(ncyl ~ suv, labeller = label_both) +
  ggtitle("Distribution of Highway mileage by cylinder and suv ")

common_cyl %>%
  ggplot(aes(x = hwy_mpg)) +
  geom_density() +
  facet_grid(ncyl ~ suv, labeller = label_both) +
  ggtitle("Distribution of Highway mileage by cylinder and suv ")
```

As shown in the figures, across both SUVs and non-SUVs, mileage tends to decrease as the number of cylinder increases. 



# 3. Numerical Summaries 

## 3.1 Calculate Center Measures

```{r}
# (a) load data
gap2007 <- gapminder::gapminder %>% filter(year == 2007)

dim(gap2007)
summary(gap2007)


# Compute groupwise mean and median lifeExp
gap2007 %>%
  group_by(continent) %>%
  summarize(mean(lifeExp),
            median(lifeExp))

# Generate box plots of lifeExp for each continent
gap2007 %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```


## 3.2 Measure of variability 

Before we decide on the measure for spread, we should study the shape of the distribution.  If we're unsure whether the data have symmetric or skewed distributions, it's a good idea to consider a robust measure like IQR in addition to the usual measures of variance or standard deviation.


```{r}
# Compute groupwise measures of spread
gap2007 %>%
  group_by(continent) %>%
  summarize(sd(lifeExp),
            IQR(lifeExp),
            n())

# Generate overlaid density plots
gap2007 %>%
  ggplot(aes(x = lifeExp, fill = continent)) +
  geom_density(alpha = 0.3)
```

Choose measures for center and spread. We need to decide the most appropriate measures to describe the center and spread. 

```{r}
# Compute stats for lifeExp in Americas
gap2007 %>%
  filter(continent == "Americas") %>%
  summarize(mean(lifeExp),
            sd(lifeExp))

# Compute stats for population
gap2007 %>%
  summarize(median(pop),
            IQR(pop))
```

Median and IQR measure the central and spread, respectively. Most importantly, they are robust to outliers and non-normal data. 



## 3.3 Shape and Transformations

**Describle the shape:**

* Modality: Unimodal, bimodal, multimodal, uniform. 

* Skew: right-skewed, left-skewed, normal


**Transformations:**

Transformations can be helpful to visualize highly skewed distributions. 

```{r}
# Create density plot of old variable
gap2007 %>%
  ggplot(aes(x = pop)) +
  geom_density()

# Transform the skewed pop variable
gap2007 <- gap2007 %>%
  mutate(log_pop = log(pop))

# Create density plot of new variable
gap2007 %>%
  ggplot(aes(x = log_pop)) +
  geom_density()
```


## 3.4 Outliers

Characteristics of a distribution: 

* Center

* Variability

* Shape

* Outliers

```{r}



# Filter for Asia, add column indicating outliers
gap_asia <- gap2007 %>%
  filter(continent == "Asia") %>%
  mutate(is_outlier = lifeExp < 50)

gap_asia %>%
  #filter(!is_outlier) %>%
  ggplot(aes(x = 1, y = lifeExp)) +
  geom_boxplot()

# Remove outliers, create box plot of lifeExp
gap_asia %>%
  filter(!is_outlier) %>%
  ggplot(aes(x = 1, y = lifeExp)) +
  geom_boxplot()
```


# 4. Case Study

## 4.1 Is Spam correlated with the length of an email? 


```{r}
# install.packages("openintro")
library(openintro) # Data sets and supplemental function from "OpenIntro" Textbooks

# (a) dataset
email <- email %>%
  mutate(spam = factor(spam, levels = c(0, 1), labels = c("not-spam", "spam")))

summary(email)
head(email)

table(email$spam)

# (b)  boxplot 
email %>%
  ggplot(aes(x = spam, y = num_char)) +
  geom_boxplot()

# (c) density
email %>%
  ggplot(aes(x = num_char, fill = spam)) +
  geom_density(alpha = 0.5)



# Compute summary statistics
email %>%
  group_by(spam) %>%
  summarise(median(num_char), 
  IQR(num_char))

# Create plot
email %>%
  mutate(log_num_char = log(num_char)) %>%
  ggplot(aes(x = spam, y = log_num_char)) +
  geom_boxplot()

```

The median length of not-spam emails is greater than that of spam emails, and the IQR of email length of not-spam is greater than that of spam. 


## 4.2 Spam and **!!!** 

An obvious indicator of spam: exclamation marks (*!!!*). 

```{r}
# boxplot of raw values
email %>%
  ggplot(aes(x = spam, y = exclaim_mess)) +
  geom_boxplot()

email %>%
ggplot(aes(x = exclaim_mess)) +
geom_histogram() + 
facet_wrap(~spam)


# Compute center and spread for exclaim_mess by spam
email %>%
  group_by(spam) %>%
  summarise(median(exclaim_mess), 
            IQR(exclaim_mess))



# Create plot for spam and exclaim_mess
email %>%
mutate(log_exclaim_mess = log(exclaim_mess + 0.01)) %>%   # log(0) = -Inf. Get around this by adding a small number (like 0.01)
ggplot(aes(x = log_exclaim_mess)) +
geom_histogram() + 
facet_wrap(~spam)

```

The most common value of *exclaim_mess* in both not-spam and spam is zero (a $\log(0 + 0.01) = - 4.6$). Even after a log transformation, the distribution of exlaim_mess in both not-spam and spam emails is still right-skewed.  The typical number of exclamations in the non-spam group appears to be slightly higher than that in the spam group. 


**Zero inflation strategies**:

* Analyze the two components separately

* Collapse into two-level categorical variable


## 4.3 Collapsing Levels

Working with heavy skewed data is a challenge, one way to simplify is to collapse low counts into one. (has image vs no image)

```{r}
table(email$image)


# Create plot of proportion of spam by image
email %>%
  mutate(has_image = image > 0) %>%
  ggplot(aes(x = has_image, fill = spam)) +
  geom_bar(position = "fill")

```

## 4.3 Data Integrity 

Attachment and image might be double counted. 

```{r}
sum(email$image > email$attach)
```

Aha, *image* is never greater than *attach*, images must be counted as attachments in the dataset. 


## 4.4 More questions

For emails containing the word "dollar", does the typical spam email contain a greater number of occurrences of the word than the typical non-spam email? Create a summary statistic that answers this question.

```{r}
email %>%
  filter(dollar > 0) %>%
  group_by(spam) %>%
  summarize(median(dollar))
```

If you encounter an email with greater than 10 occurrences of the word "dollar", is it more likely to be spam or not-spam? Create a barchart that answers this question.

```{r}
email %>%
  filter(dollar > 10) %>%
  ggplot(aes(x = spam)) +
  geom_bar()
```

##  4.5 email$number

```{r}
table(email$number)

# Reorder levels
email$number <- factor(email$number, levels = c("none", "small", "big"))

# Construct plot of number
ggplot(email, aes(x = number)) + 
  geom_bar() + 
  facet_wrap(~spam)
 

```


