---
title: "Manipulating Time Series Data in R with xts & zoo"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee



**Course Description**

"Time series are all around us, from server logs to high frequency financial data. Managing and manipulating ordered observations is central to all time series analysis. The xts and zoo packages provide a set of powerful tools to make this task fast and mistake free. In this course, you will learn everything from the basics of xts to advanced tips and tricks for working with time series data in R." 


Ref: Ryan, Jeffrey. 2018. "Manipulating Time Series Data in R with xts & zoo", https://www.datacamp.com/courses/manipulating-time-series-data-in-r-with-xts-zoo, 2018.


Note: Some course materials have been revised for internal R training. 

## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(zoo)

library(xts)

library(tidyverse)


```

# 1. Introduction to eXtensible Time Series, using xts and zoo for time series

"xts and zoo are just two of the many different types of objects that exist in R. This chapter will introduce the basic objects in xts and zoo and their components, and offers examples of how to construct and examine the data."

What is xts? 

* eXtensible Time Series

* An extended zoo object

* Matrix + Index

* Observations + Times

```{r}
# an Xts example 
# xts = matrix + index

x <- matrix(1:4, ncol = 2, nrow = 2)
x

idx <- as.Date(c("2015-01-01", "2015-02-01"))
idx

X <- xts(x = x, order.by = idx)
class(X)
X

```

Deconstructing xts

```{r}
coredata(X, fmt = FALSE)
index(X)
```

## 1.1 More than a matrix

```{r}
# View the structure of previos created X
str(X)

# Extract the 2nd observation of the 2nd column
X[2, 2]

# Extract the 2nd obsrvation of the 2nd column 
core <- coredata(X, fmt = FALSE)
core[2, 2]
```

```{r}
# Create the object data using 5 random numbers
data <- rnorm(n = 5)

# Create dates as Date class object starting from 2016-01-01
dates <- seq(as.Date("2016-01-01"), length = 5, by = "days")

# Use xts() to create smith
smith <- xts(x = data, order.by = dates)

# Create bday (1899-05-08) using a POSIXct data class object
bday <- as.POSIXct("1899-05-08")


# Create hayek and add a new attribute call born 
hayek <- xts(x = data, order.by = dates, born = bday)

hayek
str(hayek)
```

## 1.2 Deconstructing xts

```{r}
# Extract the core data of hayek
hayek_core <- coredata(hayek)

# View the class of hayek_core
class(hayek_core)

# Extract the index of hayek
hayek_index <- index(hayek)

# View the class of hayek_index
class(hayek_index)
```

## 1.3 Time based indices 


```{r}
# Create dates
dates <- as.Date("2016-01-02") + 0:4

# Create ts_a
ts_a <- xts(x = 1:5, order.by = dates)
ts_a

# create ts_b
ts_b <- xts(x = 1:5, order.by = as.POSIXct(dates))
ts_b

# Extract the rows of ts_a using the index of ts_b
ts_a[index(ts_b)]

# Extract the rows of ts_b using the index of ts_a
ts_b[index(ts_a)]

```

## 1.4 Converting xts objects


```{r}
# Convert austres to an xts object called au
head(austres)
class(austres)
austres

au <- as.xts(austres)
head(au)

# Then convert your xts object (au) into a matrix am
am <- as.matrix(au)

# Inspect the head of am
head(am)

# Convert the original austres into a matrix am2
am2 <- as.matrix(austres)

# Inspect the head of am2
head(am2)

```

## 1.5 Importing data

```{r}
# tmp_file
tmp_file <- "http://s3.amazonaws.com/assets.datacamp.com/production/course_1127/datasets/tmp_file.csv"
dat <- read.csv(tmp_file)
dat


# Convert dat into xts
xts(dat, order.by = as.Date(rownames(dat), "%m/%d/%Y"))


# Read tmp_file using read.zoo
dat_zoo <- read.zoo(tmp_file, index.column = 0, sep =",", format = "%m/%d/%Y" )
dat_zoo

# Convert dat_zoo to xts
dat_xts <- as.xts(dat_zoo)

```

## 1.6 Exporting xts objects

```{r}
# Convert sunspots to xts using as.xts()
sunspots_xts <- as.xts(sunspots)
head(sunspots_xts)

# Get the temporary file name
tmp <- tempfile()
tmp

# Write the xts object using zoo to tmp
write.zoo(sunspots_xts, sep = ",", file = tmp)

# Read teh tmp file. FUN = as.yearmon converts strings such as Jan 1749 into a proper time class
sun <- read.zoo(tmp, sep = ",", FUN = as.yearmon)
head(sun)

# Convert sun into xts. Save this as sun_xts
sun_xts <- as.xts(sun)
head(sun_xts)
```



# 2. First Order of Business - Basic Manipulations

"Now that you can create basic xts objects, it's time to see how powerful they can be. This chapter will cover the basics of one of the most useful features of xts: time based subsetting. From there you'll explore additional ways to extract data using time phrases, and conclude with how to do basic operations like adding and subtracting your xts objects."

## 2.1 Querying for dates

One & two sided intervals
```{r}
# Load fund data
data(edhec, package = "PerformanceAnalytics")

# one sided interval
head(edhec["2007-01", 1:3])

# two sided intervals
head(edhec["2007-01/2007-03", 1:3])

# Truncated dates
head(edhec["200701/03", 1:3])

# Time support
# iday["20160808T2213"]

# Repeating intraday inervals

# iday["T05:30/T06:30"]
```

```{r, eval=FALSE}
# Select all of 2016 from x
x_2016 <- x["2016"]

# Select January 1, 2016 to March 22, 2016
jan_march <- x["2016/2016-03-22"]

# Verify that jan_march contains 82 rows
82 == length(jan_march)
```
Subsetting  a range is useful way to get a snapshot of a larger time series object. 


## 2.2 Extracting recurring intraday intervals

```{r, eval=FALSE}
# Extract all data from irreg between 8AM and 10AM
morn_2010 <- irreg["T08:00/T10:00"]

# Extract the observations in morn_2010 for January 13th, 2010
morn_2010["2010-01-13"]
```


## 2.3 Row selection with time objects, Update and replace elements

Subsetting using a vector of dates is a good way when working with xts objects

```{r, eval=FALSE}
# S# Replace the values in x contained in the dates vector with NA
x[dates] <- NA

# Replace all values in x for dates starting June 9, 2016 with 0
x["2016-06-09/"] <- 0

# Verify that the value in x for June 11, 2016 is now indeed 0
x["2016-06-11"]
    
ubset x using the vector dates
dates <- as.Date(c("2016-06-04", "2016-06-8"))
x[dates]

# Subset x using dates as POSIXct
x[as.POSIXct(dates)]
```

```{r, eval=FALSE}
# Replace the values in x contained in the dates vector with NA
x[dates] <- NA

# Replace all values in x for dates starting June 9, 2016 with 0
x["2016-06-09/"] <- 0

# Verify that the value in x for June 11, 2016 is now indeed 0
x["2016-06-11"]
    

```

## 2.4 Find the first or last period of time

```{r}
# first() and last()
first(edhec[, "Funds of Funds"], "4 months")

last(edhec[, "Funds of Funds"], "1 year")
```

```{r, eval=FALSE}
# Create lastweek using the last 1 week of temps
lastweek <- last(temps, "1 week")

# Print the last 2 observations in lastweek
last(lastweek, n = 2)

# Extract all but the first two days of lastweek
first(lastweek, "-2 days")

# Extract the first three days of the second week of temps
first(last(first(temps, "2 weeks"), "1 weeks"), "3 days")
```

Relative subsetting with first() and last() can be a valuable way to look at the most recent data in your xts object. 


## 2.5 Matrix arithmetic - add, subtract, multiply, and divide in time

When we perform any binary operation using two _xts_ objects, these objects are first aligned using the intersection of the indexes. 

```{r}
a <- c(1, 1, 1)
a_index <- as.Date(c("2015-01-24", "2015-01-25", "2015-01-26"))
a <- xts(a, order.by = a_index) 

b <- xts(2, order.by = as.Date("2015-01-24"))
  
# Add a and b
a+b

# Add a with the numeric value of b
a + as.numeric(b)
```

Adding two xts objects returns only the dates common to both. Adding a numeric to an xts object is a bit more intuitive. 

## 2.6 Math with non-overlapping indexes

```{r}
# Add a to b, and fill all missing rows of b with 0
a + merge(b, index(a), fill = 0)

# Add a to b and fill NAs with the last observation
a + merge(b, index(a), fill = na.locf)
```



# 3. Merging and modifying time series

"One of the most important parts of working with time series data involves creating derived time series. To do this effectively, it is critical to keep track of dates and times. In this chapter you will look at how xts handles merging new columns and rows into existing data, how to deal with the inevitable missing observations in time series, and how to shift your series in time."

## 3.1 Combining xts by column with merge

```{r}
# create  a and b
a <- xts(x = c(-1.2,       0.27,   1.08, -2.3), order.by = as.Date(c("2016-06-05",               "2016-06-08", "2016-06-09", "2016-06-13")))
b <- xts(x = c(0.43, 0.51, -0.57, -0.54      ), order.by = as.Date(c("2016-06-05", "2016-06-06", "2016-06-08", "2016-06-09"              )))
b

# perform an inner join of a and b
merge(a, b, join = "inner")

# perform a left -join of a and b, fill missing values with 0
merge(a, b, join = "left", fill = 0)

```

## 3.2 Filling missing values using last or previous observation, interpolating using na.approx()

* na.locf(): last observation carried forward. 

* fromLast = TRUE: apply next observation carried backward 

```{r}
# create a temps
temps <- xts(x = c(1, 2, NA, 4, 5), order.by = as.Date(c("2018-12-01", "2018-12-02", "2018-12-03", "2018-12-04", "2018-12-05")))
temps

# Fill missing values in temps using the last observation
temps_last <- na.locf(temps)
temps_last

# Fill missing values in temps using the next observation
temps_next <- na.locf(temps, fromLast = TRUE)
temps_next

# na.approx()
na.approx(temps)
```

## 3.3 Combine a leading and lagging time series

```{r}
# create x 
x <- xts(x = 5:9, order.by = (as.Date("2018-12-06") + 0:4) ) 
x

# Create a leading object called lead_x
lead_x <- lag(x, k = -1)
lead_x

# create a lagging object called lag_x
lag_x <- lag(x, k = 1)   # xts follws the literature. 
lag_x

# Merge into z
z <- merge(lead_x, x, lag_x)
z

```

## 3.4 Calculate a difference of a series using diff()

```{r, eval=FALSE}

# Calculate the first difference of AirPass and assign to diff_by_hand
diff_by_hand <- AirPass - lag(AirPass, k = 1)

# Use merge to compare the first parts of diff_by_hand and diff(AirPass)
merge(head(diff_by_hand), head(diff(AirPass)))

# Calculate the first order 12 month difference of AirPass
diff(AirPass, lag = 12, differences = 1)

```


# 4. Apply and aggregate by time

"Now the fun begins! A very common usage pattern for time series is to calculate values for disjoint periods of time or aggregate values from a higher frequency to a lower frequency. For most series, you'll often want to see the weekly mean of a price or measurement. You may even find yourself looking at data that has different frequencies and you need to normalize to the lowest frequency. This chapter is where it all happens. Hang tight, and lets get going!"

## 4.1 Find interval by time in xts

```{r, eval=FALSE}
# Locate the weeks
endpoints(temps, on = "weeks")

# Locate every two weeks
endpoints(temps, on = "weeks", k = 2)
```

Locating endpoints can help speed the process of aggregating time series data. 

## 4.2 Apply a function by time periods

Apply a function to the values between endpoints. 

```{r, eval=FALSE}
# Calculate the weekly endpoints
ep <- endpoints(temps, on = "weeks")

# Now calculate the weekly mean and display the results
period.apply(temps[, "Temp.Mean"], INDEX = ep, FUN = mean)
```

## 4.3 Using lapply() and split to apply functions on intervals
```{r, eval=FALSE}
# Split temps by week
temps_weekly <- split(temps, f = "weeks")

# Create a list of weekly means, temps_avg, and print this list
temps_avg <- lapply(X = temps_weekly, FUN = mean)
temps_avg
```

## 4.3 Selection by endpoints vs. split-lapply-rbind

Find the last observation in each week
```{r}
# Use the proper combination of split, lapply and rbind
temps_1 <- do.call(rbind, lapply(split(temps, "weeks"), function(w) last(w, n = "1 day")))

# Create last_day_of_weeks using endpoints()
last_day_of_weeks <- endpoints(temps, on = "weeks")

# Subset temps using last_day_of_weeks 
temps_2 <- temps[last_day_of_weeks]

```


## 4.4 Converting periodicity

OHLC: Open, High, Low, and Close

Summary: Starting, maximum, minimum and ending value

```{r}
# Load fund data
data(edhec, package = "PerformanceAnalytics")

# Aggregate OHLC
to.period(edhec["1997/2001", 1], period = "years", name = "EDHEC")

to.period(edhec["1997/2001", 1], "years", name = "EDHEC", indexAt = "firstof")

# OHLC = FALSE
to.period(edhec[, 1], period = "years", name = "EDHEC", OHLC = FALSE)

# Extract directly 
edhec[endpoints(edhec, "years"), 1]

```


## 4.5 Convert univariate series to OHLC data
```{r}
# read data
usd_eur <- read.csv("data/USDEUR.csv")
# Convert dat into xts
usd_eur <- xts(usd_eur, order.by = as.Date(rownames(usd_eur), "%Y-%m-%d"))

class(usd_eur)

# convert usd_eur to weekly and assign to usd_eur_weekly
usd_eur_weekly <- to.period(x = usd_eur, period = "weeks")
head(usd_eur_weekly)

# Convert usd_eur to monthly and assign to usd_eur_monthly
usd_eur_monthly <- to.period(x = usd_eur, period = "months")
head(usd_eur_monthly)

# Convert usd_eur to yearly univariate and assign to usd_yearly
usd_eur_yearly <- to.period(x = usd_eur, period = "years", OHLC = FALSE)
head(usd_eur_yearly)
```

Aggregating over time and converting from univariate to OHLC (and vice-versa) are useful skills for time series analysis, especially with financial data.


## 4.6 Convert a series to a lower frequency

*to.period()* can also convert data to lower regularized frequency. 

```{r}
# Load fund data
data(edhec, package = "PerformanceAnalytics")
eq_mkt <- edhec
head(eq_mkt)

# Convert eq_mkt to quarterly OHLC
mkt_quarterly <- to.period(x = eq_mkt, period = "quarters")
head(mkt_quarterly)


# Convert eq_mkt to quarterly using shortcut function
mkt_quarterly2 <- to.quarterly(eq_mkt, name = "edhec_equity", indexAt = "firstof")
head(mkt_quarterly2)
```


## 4.7 Rolling Functions
```{r}
# Discrete rolling windows
x <- xts(c(1, 2, 3), as.Date("2016-01-01") + 0:2)
x
cbind(x, cumsum(x))

# Discrete rolling windows
edhec.yrs <- split(edhec[, 1], f = "years")

edhec.yrs <- lapply(edhec.yrs, cumsum)

edheck.ytd <- do.call(rbind, edhec.yrs)

cbind(edheck.ytd, edhec[, 1])["2007-10/2008-03"]


# continous rolling windows
rollapply(edhec["200701/08", 1], width = 3, FUN = mean)
```

## 4.8 Calculate basic rolling value of series by month

```{r}
# Split edhec into years
edhec_years <- split(edhec , f = "years")

# Use lapply to calculate the cumsum for each year in edhec_years
edhec_ytd <- lapply(edhec_years, FUN = cumsum)

# Use do.call to rbind the results
edhec_xts <- do.call(rbind, edhec_ytd)
edhec_xts["1997-10/1998-03"]

```

## 4.9 Calculate the rolling standard deviation of a time series
```{r}
# Use rollapply to calculate the rolling 3 period sd of eq_mkt
eq_sd <- rollapply(eq_mkt, width = 3, FUN = sd)

head(eq_sd)
```


# 5. Extra Features of xts

"Now that you are comfortable with most of the core features, its time to explore some of the lesser known (but powerful!) aspects of working with xts. In this final chapter you will use the internals of the index to find repeating intervals, see how xts provides intuitive time zone support, and experiment with ways to explore your data by time - including identifying frequency and coverage in time. Let's finish this course!"

```{r}
Z <- xts(c(1:3), Sys.Date()+0:2)
Z

indexClass(Z)
indexTZ(Z)

indexFormat(Z) <- "%B %d, %Y"
Z

# set time zone
Sys.setenv(TZ = "America/Los_Angelas")

# know your time zones
# help("OlsonNames")

tclass(Z)
```

## 5.1 Class attributes - tclass, tzone, and tformat

```{r, eval=FALSE}
# View the first three indexes of temps
index(temps)[1:3]

# Get the index class of temps
indexClass(temps)

# Get the timezone of temps
indexTZ(temps)

# Change the format of the time display
indexFormat(temps) <- "%b-%d-%Y"

# View the new format
head(temps)
```

These commands allow you to quickly and easily modify the internal characteristics of your xts object.

## 5.2 Time Zones 
```{r, eval=FALSE}
# Construct times_xts with tzone set to America/Chicago
times_xts <- xts(1:10, order.by = times, tzone = "America/Chicago")

# Change the time zone of times_xts to Asia/Hong_Kong
tzone(times_xts) <- "Asia/Hong_Kong"
  
# Extract the current time zone of times_xts
indexTZ(times_xts)
```

## 5.3 Periods, periodicity, and timestamps
```{r}
# Periodicity
periodicity(edhec)

periodicity(to.yearly(edhec))

# Counting periods
nyears(edhec)
nquarters(edhec)
nmonths(edhec)
nweeks(edhec)
ndays(edhec)
nhours(edhec)

# Broken down time with .index*
index(edhec)

.indexmday(edhec)
.indexyday(edhec)
.indexyear(edhec) + 1900

# Remove observation of duplicate timestampls
make.index.unique(edhec[1:5, 1:5], eeps = 1e-06, drop = TRUE, fromLast = TRUE)
```

## 5.4 Determining periodicity
```{r, eval=FALSE}
# Calculate the periodicity of temps
periodicity(temps)

# Calculate the periodicity of edhec
periodicity(edhec)

# Convert edhec to yearly
edhec_yearly <- to.yearly(edhec)

# Calculate the periodicity of edhec_yearly
periodicity(edhec_yearly)
```

## 5.5 Find the number of periods in your data
```{r}
# Count the months
nmonths(edhec)

# Count the quarters
nquarters(edhec)

# Count the years
nyears(edhec)

```

## 5.6 index commands to subset your time series data

```{r}
# Explore underlying units of temps in two commands: .index() and .indexwday()
.index(temps)
.indexwday(temps)


# Create an index of weekend days using which()
index <- which(.indexwday(temps) == 6 | .indexwday(temps) == 0)

# Select the index
temps[index]


```

## 5.7 Modifying timestamps
```{r}
# Make z have unique timestamps
z_unique <- make.index.unique(z, eps = 1e-4)

# Remove duplicate times in z
z_dup <- make.index.unique(z, drop = TRUE)

# Round observations in z to the next hour
z_round <- align.time(z, n = 3600)


```

