---
title: "Package SixSigma - Six Sigma Tools for Quality Control and Improvement"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

**SixSigma** contains functions and utilities to perform Statistical Analyses in the Six Sigma way.
Through the DMAIC cycle (Define, Measure, Analyze, Improve, Control), you can manage several
Quality Management studies: Gage R&R, Capability Analysis, Control Charts, Loss Function
Analysis, etc.


Authors: Emilio L. Cano, Javier M. Moguerza, Mariano Prieto and Andrés Redchuk


## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(e1071)

library(SixSigma)
```


# 1. climProfile: Compute profile limits

Function to compute prototype profile and confidence bands for a set of profiles (Phase I)

```{r}
# data: 

# contains 50 profiles corresponding to the density measurments of 50 0.5 in-thick engineered woodboard, measured in 500 locations

head(ss.data.wby)
glimpse(ss.data.wby) # 500 rows (locations) and 50 columns (woodboard)

# Contains the 500 locations at which the density of a 0.5 in-thick engineered woodboard is measured
head(ss.data.wbx)
glimpse(ss.data.wbx)

# Profile
plotProfiles(profiles = ss.data.wby, 
             x = ss.data.wbx)


# Compute profiles limits
wby.phase1 <- ss.data.wby[, 1:35]
wb.limits <- climProfiles(profiles = wby.phase1, 
                          x = ss.data.wbx, 
                          smoothprof = FALSE, 
                          smoothlim = FALSE)


plotProfiles(profiles = wby.phase1, 
             x = ss.data.wbx, 
             cLimits = wb.limits)

```

# 2. Get out-of-control profiles - outProfiles()

Returns a list with information about the out-of-control profiles given a set of profiles and some
control limits

```{r}
wby.phase1 <- ss.data.wby[, 1:35]
wb.limits <- climProfiles(profiles = wby.phase1, 
                          x = ss.data.wbx, 
                          smoothprof = TRUE, 
                          smoothlim = TRUE)

wby.phase2 <- ss.data.wby[, 36:50]

wb.out.phase2 <- outProfiles(profiles = wby.phase2, 
                             x = ss.data.wbx, 
                             cLimits = wb.limits, 
                             tol = 0.8)
wb.out.phase2

plotProfiles(wby.phase2, 
             x = ss.data.wbx, 
             cLimits = wb.limits,
             outControl = wb.out.phase2$idOut, 
             onlyout = TRUE)
```

# 3. Profiles control plot - plotControlProfiles()

Plots the proportion of times that each profile remains out of the confidence bands

```{r}
wby.phase1 <- ss.data.wby[, 1:35]

wb.limits <- climProfiles(profiles = wby.phase1,
                          x = ss.data.wbx,
                          smoothprof = TRUE,
                          smoothlim = TRUE)

wby.phase2 <- ss.data.wby[, 36:50]

wb.out.phase2 <- outProfiles(profiles = wby.phase2,
                             x = ss.data.wbx,
                             cLimits = wb.limits,
                             tol = 0.8)

plotControlProfiles(wb.out.phase2$pOut, 
                    tol = 0.8)
```

# 4. Regularise set of profiles - smoothProfiles()

This function takes a set of profiles and regularise them by means of a SVM


```{r}
wby.smooth <- smoothProfiles(profiles = ss.data.wby,
                             x = ss.data.wbx)

plotProfiles(profiles = wby.smooth,
             x = ss.data.wbx)
```


# 5. ss.ca.yield
Computes the Yield, First Time Yield, Rolled Throughput Yield and Defects per Million Opportunities
of a process.

```{r}
ss.ca.yield(defects = c(3, 5, 12), 
            rework = c(1, 2, 4), 
            opportunities = 1915)

```

# 6. Data for the batteries example
This is a simulated data set of 18 measurements of the voltage of batteries using different voltmeters.

```{r}
head(ss.data.batteries)
glimpse(ss.data.batteries)

# summary
summary(ss.data.batteries)

# line + point plot 
ss.data.batteries %>%
  ggplot(aes(x = as.numeric(run), y = voltage, color = battery)) + 
  geom_point() + 
  geom_line() +
  labs(x = "Run") +
  facet_wrap(~ voltmeter )

# boxplot 
plot(voltage ~ voltmeter, data = ss.data.batteries)
```


# 7. Gage R & R (Measurement System Assessment) 

*ss.rr*: Performs Gage R&R analysis for the assessment of the measurement systems of a process. Related to the measure phase of the DMAIC strategy of Six Sigma

**Usage**: 

Performs an R&R study for the measured variable, taking into account part and appraiser factors. It ouputs the sources of Variability, and six graphs: bar chart with the sources of Variability, plots by appraise, part and interaction and x-bar adn R control charts. 


```{r}
# data set: Gage R&R data - Example data for measurement phase of the Six Sigma methodology

head(ss.data.rr)
str(ss.data.rr)
summary(ss.data.rr)

# plot
ss.data.rr %>%
  ggplot(aes(x = operator, y = time1)) + 
  geom_boxplot() + 
  stat_boxplot(geom = "errorbar", width = 0.5) +
  geom_jitter() + 
  facet_wrap(~ prototype) +
  labs(title = "Time 1")

ss.data.rr %>%
  ggplot(aes(x = operator, y = time2)) + 
  geom_boxplot() + 
  stat_boxplot(geom = "errorbar", width = 0.5) +
  geom_jitter() + 
  facet_wrap(~ prototype) +
  labs(title = "Time 2")

# grouped summary
ss.data.rr %>%
  group_by(prototype, operator)  %>%
  summarize(time1_mean = mean(time1), 
            time1_sd = sd(time1), 
            time2_mean = mean(time2), 
            time2_sd = sd(time2))


## Gage R&R 
ss.rr(var = time1, 
      part = prototype, 
      appr = operator, 
      data = ss.data.rr, 
      sub = "Six Sigma Helicopter Project", 
      alphaLim = 0.05, 
      errorTerm = "interaction", 
      lsl = 0.7, 
      usl = 1.8)
```


## 7.2 Edeniq data: Gage R&R
```{r}

library(edeniqRpackage)

GLE_WATERT_Liq_Starch %>%
  # filter(mz_score < 4 ) %>%
  filter(batch >= 12616) %>%
  ggplot(aes(x = factor(batch), y = liq_starch_pct, color = file_name)) +
 # geom_violin(trim = FALSE) +
  geom_jitter(shape = 16, position = position_jitter(0.2), alpha = 0.7) +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1))



temp <-  GLE_WATERT_Liq_Starch %>%
 # filter(file_name == "012219 Glacial Lakes Total CEF LIQ Master.xlsx" )
  #filter(mz_score < 1.76 ) %>%
  filter(batch >= 12616) %>%
  mutate(parts = factor(batch), 
         operators = factor(file_name, labels = c("op_time1", "op_time2")), 
         run = factor(rep)) %>%
  select(parts, operators,  run, var = liq_starch_pct)

## Gage R&R 
ss.rr(var = var, 
      part = parts, 
      appr = operators, 
      data = temp, 
      sub = "Six Sigma Edeniq Project", 
      alphaLim = 0.05, 
      errorTerm = "interaction")

lm_model <- lm(var ~ batch + operator_time, data = temp)
anova(lm_model)
```

