---
title: "Joining Data in R with dplyr"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

**Course Description**
"This course builds on what you learned in Data Manipulation in R with dplyr by showing you how to combine data sets with dplyr's two table verbs. In the real world, data comes split across many data sets, but dplyr's core functions are designed to work with single tables of data. In this course, you'll learn the best ways to combine data sets into single tables. You'll learn how to augment columns from one data set with columns from another with mutating joins, how to filter one data set against another with filtering joins, and how to sift through data sets with set operations. Along the way, you'll discover the best practices for building data sets and troubleshooting joins with dplyr. Afterwards, you’ll be well on your way to data manipulation mastery!"


Instructor: Garrett Grolemund  (Data Scientist at RStudion)

Ref: Grolemund, G. https://www.datacamp.com/courses/joining-data-in-r-with-dplyr. Accessed on 11/20/2018




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (I) Load Required Libraries
```{r, message = FALSE}
library(dplyr)
library(tidyverse)

```

# 1. Mutating joins

* `left_join()`: the basic join function in *dplyr*

* `rignt_joint()`: 

* `inner_joint()`

* `full_jont()`


Note: left_joint(x, y, by = "name") equals to right_joint(y, x, by = "name")

check a dataset (A) is equal to a dataset (B): 

* `setequal(A, B)`


```{r}
# (a) create a dataframe of songs

songs <- tibble(song = c("Come Together", "Dream On", "Hello, Goodbye", "It's Not Unusual"), 
                album = c("Abbey Road", "Aerosmith", "Magical Mystery Tour", "Along Came Jonnes"), 
                first = c("John", "Steven", "Paul", "Tom"), 
                last = c("Lennon", "Tyler", "McCartney", "Jones"))

# (b) Create the dataframe of artists
artists <- tribble(~first, ~last, ~instrument,
                    "Jimmy",   "Buffett",  "Guitar",    
                    "George", "Harrison ",  "Guitar",    
                    "Mick  ", "Jagger",  "Vocals",    
                    "Tom   ", "Jones",  "Vocals",    
                    "Davy  ", "Jones",  "Vocals",    
                    "John  ", "Lennon",  "Guitar",    
                    "Paul  ", "McCartney",  "Bass  ",    
                    "Jimmy ", "Page",  "Guitar",    
                    "Joe   ", "Perry",  "Guitar",    
                    "Elvis ", "Presley  ",  "Vocals",    
                    "Keith ", "Richards ",  "Guitar",    
                    "Paul  ", "Simon    ",  "Guitar",    
                    "Ringo ", "Starr    ",  "Drums ",    
                    "Joe   ", "Walsh    ",  "Guitar",    
                    "Brian ", "Wilson   ",  "Vocals",    
                    "Nancy ", "Wilson   ",  "Vocals") 

# removing whitespace
artists <- data.frame(lapply(artists, trimws), stringsAsFactors = FALSE)


# (c) create albums 
albums <- tribble(~album, ~band, ~year, 
                  "A Hard Day's Night       ",  "The Beatles       ",   1964,
                  "Magical Mystery Tour     ",  "The Beatles       ",   1967,
                  "Beggar's Banquet         ",  "The Rolling Stones",   1968,
                  "Abbey Road               ",  "The Beatles       ",   1969,
                  "Led Zeppelin IV          ",  "Led Zeppelin      ",   1971,
                  "The Dark Side of the Moon",  "Pink Floyd        ",   1973,
                  "Aerosmith                ",  "Aerosmith         ",   1973,
                  "Rumours                  ",  "Fleetwood Mac     ",   1977,
                  "Hotel California         ",  "Eagles            ",   1982) 

# remove whitespace
albums <- data.frame(lapply(albums, trimws), stringsAsFactors = FALSE)

# (d) Create "bands"
bands <- tribble(~first, ~last, ~band, 
                  "John     ",  "Bonham   ", "Led Zeppelin      ",
                  "John Paul",  "Jones    ", "Led Zeppelin      ",
                  "Jimmy    ",  "Page     ", "Led Zeppelin      ",
                  "Robert   ",  "Plant    ", "Led Zeppelin      ",
                  "George   ",  "Harrison ", "The Beatles       ",
                  "John     ",  "Lennon   ", "The Beatles       ",
                  "Paul     ",  "McCartney", "The Beatles       ",
                  "Ringo    ",  "Starr    ", "The Beatles       ",
                  "Jimmy    ",  "Buffett  ", "The Coral Reefers ",
                  "Mick     ",  "Jagger   ", "The Rolling Stones",
                  "Keith    ",  "Richards ", "The Rolling Stones",
                  "Charlie  ",  "Watts    ", "The Rolling Stones",
                  "Ronnie   ",  "Wood     ", "The Rolling Stones")

# remove whitespace
bands <- data.frame(lapply(bands, trimws), stringsAsFactors = FALSE)

```


## 1.1 Basic join
```{r}
bands %>%
  left_join(artists, by = c("first", "last"))


songs %>%
  inner_join(albums, by = "album")


artists %>%
  full_join(bands, by = c("first", "last"))
```


# 2. Filtering joins and set operations

## 2.1 Filtering joins 

* `semi_joint()`: A concise way to filter data from the first dataset based on information in a second dataset

* `anti_join()`: use an anti-join to see which rows will not be matched to a second dataset by a join

```{r}
# find artists with songs
artists %>%
  semi_join(songs, by = c("first", "last"))

albums %>% 
  # Collect the albums made by a band
  semi_join(bands, by = "band") %>% 
  # Count the albums made by a band
  nrow()

# Return rows of artists that don't have bands info
artists %>% 
  anti_join(bands, by = c("first", "last"))


```

## 2.2 Set operations

* `union()`: combine two datasets without duplicating any values. It will remove duplicate rows. 

* `intersect()`: very much like `semi_join()* if datasets contain the exact same variables

* `setdiff()`: 


## 2.3 Comparing datasets

* `setequal(x, y)` 



# 3. Assembling data

## 3.1 `bind_rows()`

```{r}
# list files inside a zip file 
master <- as.character(unzip("data/jimi_hendrix.zip", list = TRUE)$Name)

# read files from the zip files
discography <- read_csv(unz("data/jimi_hendrix.zip", master[2]))

jimi <- list(`Are You Experienced` = read_csv(unz("data/jimi_hendrix.zip",  master[3])), 
                 `Axis: Bold As Love` = read_csv(unz("data/jimi_hendrix.zip",  master[4])),
                 `Electric Ladyland` = read_csv(unz("data/jimi_hendrix.zip",  master[5])))
 
jimi %>% 
  # Bind jimi into a single data frame
  bind_rows(.id = "album") %>% 
  # Make a complete data frame
  left_join(discography, by = "album")
 
```

## 3.2 `bind_cols()`

```{r}
# list files inside a zip file 
master <- as.character(unzip("data/hank_williams.zip", list = TRUE)$Name)

# read files from the zip files without unzip
hank_charts <- read_csv(unzip("data/hank_williams.zip", master[1]), col_types = list(col_character(), col_integer()))


hank_years <- read_csv(unzip("data/hank_williams.zip", master[2]), col_types = list(col_integer(), col_character()))


# bind columns

hank_years %>% 
  # Reorder hank_years alphabetically by song title
  arrange(song) %>% 
  # Select just the year column
  select(year) %>% 
  # Bind the year column
  bind_cols(hank_charts) %>% 
  # Arrange the finished dataset
  arrange(year, song)
```

If two datasets whose rows do not align, it is meaningless data if we bind them together. 


## 3.3 data_frame(), as_data_frame()

```{r}
# create three vectors for the following excericises
hank_year <- hank_years$year
hank_song <- hank_years$song
hank_peak <- hank_charts$peak


# Make combined data frame using data_frame()
data_frame(year = hank_year, 
           song = hank_song, 
           peak = hank_peak ) %>% 
  # Extract songs where peak equals 1
  filter(peak == 1)
```



# 4. Advanced joining

# 5. Case study 

Lahman's Baseball Database. The Lahman's baseball database contains 26 tables. 

```{r, eval = FALSE}
library(Lahman)

LahmanData %>% `$`(file)

# Generate a concise description of the data files in the Lahman package

Lahman_dataframe <- vcdExtra::datasets(package="Lahman") %>%
 separate(col = dim, into = c("obs", "features_no"), sep = "x") %>% arrange(Item) %>% `$`(Item)

# list variable names of data frames in Lahman
message("need to extract variable names of data frames in Lahmna")

# Examine lahmanNames
lahmanNames

# Find variables in common
reduce(lahmanNames, intersect)

```

No common variables in all 26 datasets.  Let's try to find the most common vairable names. 


```{r, eval = FALSE}
lahmanNames %>%  
  # Bind the data frames in lahmanNames
  bind_rows(.id = "dataframe") %>%
  # Group the result by var
  group_by(var) %>%
  # Tally the number of appearances
  summarize(n = n()) %>%
  # Filter the data
  filter(n>1) %>% 
  # Arrange the results
  arrange(desc(n))
```

*yearID, playerID, lgID, and teamId are the most common variable names. 

Which datasets use *playerID*? 

```{r, eval = FALSE}
lahmanNames %>% 
  # Bind the data frames
  bind_rows(.id = "dataframe") %>%
  # Filter the results
  filter(var == "playerID") %>% 
  # Extract the dataframe variable
  `$`(dataframe)
```

The *playerID* variable appears in 19 of the 26 datasets in Lahman. 


**Who are the players**

Generate a concise list of player IDs and names. 

```{r}
head(Master)

players <- Master %>% 
  # Return one row for each distinct player
  distinct(playerID, nameFirst, nameLast)

summary(players)

```

How many players are completely missing salary information? 

```{r}
players %>% 
  # Find all players who do not appear in Salaries
  anti_join(Salaries, by = "playerID") %>%
  # Count them
  count()

```

We are missing the salaries for 13,958 players. No wonder baseball players strikes so often!

Is it possible that these players somehow did not play (and hence did not earn a salary)?

```{r}
players %>% 
  anti_join(Salaries, by = "playerID") %>% 
  # How many unsalaried players appear in Appearances?
  semi_join(Appearances, by = "playerID") %>% 
  count()
```

There are 13,765 players who played a game, but are missing salary information. 193 palyers neither played a game nor have a recorded salary. 

Perhaps the unsalaried players only played one or two games, and hence did not earn a full salary. 

```{r}
players %>% 
  # Find all players who do not appear in Salaries
  anti_join(Salaries, by = "playerID") %>% 
  # Join them to Appearances
  left_join(Appearances, by = "playerID") %>% 
  # Calculate total_games for each player
  group_by(playerID) %>%
  summarize(total_games = sum(G_all, na.rm = TRUE)) %>%
  # Arrange in descending order by total_games
  arrange(desc(total_games))
```

Many of these players appeared in thousands of games, which rejects our hypothesis that they did not appear in enough games ot earn a salary. 

Is it possible that the unsalaried players did not actually play in the games that they appeared in? 

One way to check would be to determine if the players had an at-bat (i.e. batted) in the games that they appeared in.

```{r}
players %>%
  # Find unsalaried players
  anti_join(Salaries, by = "playerID") %>% 
  # Join Batting to the unsalaried players
  left_join(Batting, by = "playerID") %>% 
  # Group by player
  group_by(playerID) %>% 
  # Sum at-bats for each player
  summarize(total_at_bat = sum(AB, na.rm = TRUE)) %>% 
  # Arrange in descending order
  arrange(desc(total_at_bat))
```

The unpaid players definitely participated in the games.


**Hall of fame nominations**

The Hall of Fame is a collection of distinguished baseball players selected by a committee of baseball experts. As with any hall of fame, more players are nominated for membership than are actually admitted.


How many players have been nominated for the Hall of Fame? 
```{r}
# Find the distinct players that appear in HallOfFame
nominated <- HallOfFame %>% 
  distinct(playerID)

nominated %>% 
  # Count the number of players in nominated
  count()

nominated_full <- nominated %>% 
  # Join to Master
  left_join(Master, by = "playerID") %>% 
  # Return playerID, nameFirst, nameLast
  select(playerID, nameFirst, nameLast)
```

1,260 players were nominated for the hall of fame. Let's now see how many players were admitted to the hall of fame to examine how selective the voting process is. 

```{r}
# Find distinct players in HallOfFame with inducted == "Y"
inducted <- HallOfFame %>% 
  filter(inducted == "Y") %>% 
  distinct(playerID)

inducted %>% 
  # Count the number of players in inducted
  count()

inducted_full <- inducted %>% 
  # Join to Master
  left_join(Master, by = "playerID") %>% 
  # Return playerID, nameFirst, nameLast
  select(playerID, nameFirst, nameLast)
```

317 players have been inducted into the Hall of Fame out of 1,260 nominees. 

```{r}
# Tally the number of awards in AwardsPlayers by playerID
nAwards <- AwardsPlayers %>%
  group_by(playerID) %>%
  tally()

nAwards %>% 
  ungroup() %>%
  # Filter to just the players in inducted 
  semi_join(inducted, by = "playerID") %>% 
  # Calculate the mean number of awards per player
  summarize(avg_n = mean(n))

nAwards %>% 
  ungroup() %>%
  # Filter to just the players in nominated 
  semi_join(nominated, by = "playerID") %>% 
  # Filter to players NOT in inducted 
  anti_join(inducted, by = "playerID") %>% 
  # Calculate the mean number of awards per player
  summarize(avg_n = mean(n, na.rm = TRUE)) 
```

Does the maximum salary earned by inductees tend to be greater than the maximum salary earned by nominees who were not inducted?

```{r}
# Find the players who are in nominated, but not inducted
notInducted <- nominated %>% 
  anti_join(inducted, by = "playerID")

Salaries %>% 
  # Find the players who are in notInducted
  semi_join(notInducted, by = "playerID") %>% 
  # Calculate the max salary by player
 group_by(playerID) %>% 
  summarize(max_salary = max(salary)) %>% 
  # Calculate the average of the max salaries
  summarize(avg_salary = mean(max_salary))

# Repeat for players who were inducted
Salaries %>% 
  semi_join(inducted, by = "playerID") %>% 
  group_by(playerID) %>% 
  summarize(max_salary = max(salary)) %>% 
  summarize(avg_salary = mean(max_salary))

```


**Retirement**

One of the rules of the Hall of Fame is that players cannot be nominated until five years after they retire. Is this reflected in our data?

```{r}
Appearances %>% 
  # Filter Appearances against nominated
  semi_join(nominated, by = "playerID") %>% 
  # Find last year played by player
  group_by(playerID) %>% 
  summarize(last_year = max(yearID)) %>% 
  # Join to full HallOfFame
  left_join(HallOfFame, by = "playerID") %>% 
  # Filter for unusual observations
  filter( last_year >= yearID)
```

It looks like quite a few players have been nominated before they retired. It is much less practice in recent years. 

