---
title: "Importing Data in R Part 2"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

# (I) Load Required Libraries

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(jsonlite)
library(rvest)
library(scales)
library(lubridate)

```



# 1. Downloading Files and Using API Clients

**Objectives**: 

* Learn how to download and read in static files

* develop skills to use *APIs* when pre-existing clients are available


Working with Web Data in R include: 

*  Downloading files and using specialized packages to get data from web

* *httr* package to query *APIs* using *GET()* and *POST()*

* *JSON* and *XML*: data formats commonly returned

* *CSS* to navigate and extract data from web packages

## 1.1  importing data from a URL

```{r}

# (a) importing data from a URL

# (a1) This .CSV file contains the monthly Renewable Identification Number (RIN) generation data 
# for the Renewable Fuel Standard (RFS)
RINS_201810 <- read_csv('https://www.epa.gov/sites/production/files/2018-10/generationbreakout.csv')

# (a2) This .CSV file contains the Renewable Identification Number (RIN) availability to date for the Renewable Fuel Standard (RFS).
RINS_available_201810 <- read_csv("https://www.epa.gov/sites/production/files/2018-10/availablerins.csv") %>% 
  mutate(`Fuel (D Code)` = factor(`Fuel (D Code)`), 
         RIN_Year = make_date(year = `RIN Year`, month = 1, day = 30))

# (b) quick summary 
summary(RINS_201810)
summary(RINS_available_201810)

# (c) check the imported data
RINS_201810 %>%
  mutate(RINs_code = as.factor(`Fuel Code`), 
         `Total RINs by Fuel Code (million)` = `Total RINs by Fuel Code`/10^6) %>%
  ggplot(aes(x = `RIN Year`, y = `Total RINs by Fuel Code (million)`, color = RINs_code)) +
  geom_point() + 
  geom_line() + 
  scale_y_continuous(label = scales::comma) +
  facet_wrap(~RINs_code, ncol = 1, scales = "free")
 

# RINS_available_201810 %>%
#   #filter(Assignment == "Assigned") %>%
#   mutate(`Total Available (million)` = `Total Available`/10^6) %>%
#   ggplot(aes(x = RIN_Year, y = `Total Available (million)`, color = `Fuel (D Code)`)) +
#   geom_point() + 
#   geom_line() + 
#   scale_y_continuous(label = scales::comma) +
#   facet_wrap(~`Fuel (D Code)` + Assignment, ncol = 2, scales = "free")          

# (d) Saving formatted files to disk

saveRDS(object = RINS_201810, file = "data/RINS_201810.RDS")
saveRDS(object = RINS_available_201810, file = "data/RINS_available_201810.RDS")

readRDS("data/RINS_201810.RDS")

```

## 1.2 Downloading data from a URL

```{r}
download.file(
  url ='https://www.epa.gov/sites/production/files/2018-10/generationbreakout.csv', 
  destfile = "data/RINS_201810.csv"
)

download.file(
  url ="https://www.epa.gov/sites/production/files/2018-10/availablerins.csv", 
  destfile = "data/RINS_available_201810.csv"
)
```

## 1.3 APIs

**APIs**: Application programming interfaces

* *websites*, but for machines

* can be used to expose data automatically

* Lets us make queries for specific bits of that data


API Clients: native (in R!) interfaces to APIs, hides API complexity, let us read data in as R objects

Using API Clients: always use a client if we can; find them by googling `CRAN [name of website]' 


```{r}
# a client to Wikipedia's API
# Load pageviews
library(pageviews)

# Get the pageviews for "Hadley Wickham"
RINs_pageviews <- article_pageviews(project = "en.wikipedia","Renewable Identification Number")

# Examine the resulting object
str(RINs_pageviews)


##----------Using access tokens-------------
# Load birdnik
#install.packages("birdnik")
library(birdnik)

# Get the word frequency for "vector", using api_key to access it
api_key <- "d8ed66f01da01b0c6a0070d7c1503801993a39c126fbc3382"
vector_frequency <- word_frequency(api_key, "vector")


```



# 2. Using *httr* to interact with APIs directly 


# 3. Handling JSON and XML


# 4. Web Scrapping with XPATHs



# 5. CSS Web Scraping and Final Case Study
