---
title: "Monte Carlo Package in R "
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"A happy life is one spent in learning, earning, and yearning". - Lillian Gish


https://cran.r-project.org/web/packages/MonteCarlo/vignettes/MonteCarlo-Vignette.html



# (I) Load required libraries
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(gmodels)
library(MVN)

library(MonteCarlo)
```


There are only two main functions in the package:

* MonteCarlo() runs a simulation study for a user defined parameter grid. It handles the generation of loops over these parameter grid and parallelizes the computation on a user specified number of CPUs.

* MakeTable() creates LaTeX tables from the output of MonteCarlo(). It stacks high dimensional output arrays into tables with a user specified ordering of rows and columns.


# 1. A First Example

```{r}
# A standard t-test : t = x-hat / sigma-hat

# define function that generate data and applies the method of interest
ttest <- function(n, loc, scale) {
   # generate sample
   sample <- rnorm(n, loc, scale)
   # calculate test statistic
   stat <- mean(sample)/(sd(sample)/sqrt(n))
   # get test decision: 
   decision <- abs(stat) > 1.96
   # return result
   return(list("decision" = decision))
   
}


# define parameter grid:
n_grid <- c(50, 100, 250, 500)
loc_grid <- seq(0, 1, 0.2)
scale_grid <- c(1, 2)

# collect parameter grids in list:
param_list <- list("n" = n_grid, "loc" = loc_grid, "scale" = scale_grid)

# Run simulation:
MC_result <- MonteCarlo(func = ttest, nrep = 1000, param_list = param_list)

# summary 
summary(MC_result)

# generate table
MakeTable(output = MC_result, rows = "n", cols = c("loc", "scale"), digits = 2, include_meta = FALSE)

```

# 2. MakeFrame

```{r}
# define function that generate data and applies the method of interest
ttest <- function(n, loc, scale) {
   # generate sample
   sample <- rnorm(n = n, mean = loc, sd = scale)
   # calculate test statistic
   stat <- sqrt(n) * mean(sample)/sd(sample)
   # return result
   return(list("stat" = stat))
   
}


# define parameter grid:
n_grid <- c(50, 100, 250, 500)  # 4 levels
loc_grid <- seq(0, 1, 0.2)      # 6 levels
scale_grid <- c(1, 2)           # 2 levels

# collect parameter grids in list:
param_list <- list("n" = n_grid, "loc" = loc_grid, "scale" = scale_grid)

# Run simulation:
MC_result <- MonteCarlo(func = ttest, nrep = 1000, param_list = param_list)

# Make a dataframe
df <- MakeFrame(MC_result) %>% as_tibble()
head(df)

dim(df)   

# plot
df %>%
   filter(loc == 0.2 & scale == 1) %>%
   ggplot(aes(x = stat, col = factor(n))) +
   geom_density() 

```



# 3. EPA Cellulose Conversion

```{r}
# define function that generate data and applies the method of interest
ccf <- function(n) {
   # generate sample
   comp1 <- rnorm(n = n, mean = 76, sd = 1.52)
   comp2 <- rnorm(n = n, mean = 70, sd = 1.4)
   comp3 <- rnorm(n = n, mean = 13, sd = 1.3)
   comp4 <- rnorm(n = n, mean = 9, sd = 0.9)
   
   # calculate test statistic
   stat <- mean( (comp1 - comp2) - (comp3 - comp4) )
   # return result
   return(list("stat" = stat))
   
}

# define parameter grid:
n_grid <- c(1, 10, 20, 50, 100)  # 4 levels



# collect parameter grids in list:
param_list <- list("n" = n_grid)

# Run simulation:
ccf_result <- MonteCarlo(func = ccf, nrep = 1000000, param_list = param_list)

# Make a dataframe
df <- MakeFrame(ccf_result) %>% as_tibble()
head(df)

dim(df)   


# Density plot 
ggplot(data = df, aes(x = stat)) +
  geom_density(adjust = 2) + 
  facet_wrap(~ n, nrow = 1) +
  geom_vline(xintercept = c(0,2, 6), linetype = "dotted", col = "red") +
  scale_x_continuous(limits = c(-5, 10))


# plot
df %>%
  mutate(n= as.factor(n)) %>%
   ggplot(aes(x = stat, fill = n)) +
   geom_histogram()  +
  facet_wrap(~ n, nrow = 1, scales = "free_x") +
  geom_vline(xintercept = c(0,2, 6), linetype = "dotted", col = "red", size = 0.8) 
  

df %>%
   filter(n == 10) %>%
   ggplot(aes(x = stat)) +
   geom_histogram() 

df %>%
   filter(n == 20) %>%
   ggplot(aes(x = stat)) +
   geom_histogram() 

df %>%
   filter(n == 100) %>%
   ggplot(aes(x = stat)) +
   geom_histogram() 


# boxplot
df %>%
   ggplot(aes(x = factor(n), y = stat)) + 
   #geom_boxplot() + 
   geom_violin() + 
   geom_hline(yintercept = c(1, 2, 3), linetype = "dotted", col = "red")

df %>%
   filter(n == 100) %>%
   mutate(out_of_range = between(stat, 1.6, 2.4)) %>%
   summarize(mean = mean(out_of_range))


min(df$stat)
max(df$stat)
```

