---
title: "Data Visualization with ggplot2 (Part2)"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


Ref: Scavetta, R. Data Visualization with ggplot2 (Part2). https://www.datacamp.com/courses/data-visualization-with-ggplot2-2, 2018.


**Course Description**

* Statistics

* Coordinates and Facets

* Best Practices

* Case Study 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (I) Load Required Libraries
```{r, message = FALSE}
library(tidyverse)
```

# 1. Statistics with Geoms

**Objectives:**

* Learn how to use ggplot2 for graphic data analysis


Statistics Layer:

* Call from within a geom_

* call independently: stat_

stat_bin(): related with geom_histogram(), geom_bar(), and geom_freqpoly(). 


stat_smooth(): realted with  geom_smooth()



## 1.1 Smoothing 


```{r}
dim(mtcars)

# ggplot2 is already loaded

# Explore the mtcars data frame with str()
str(mtcars)

# A scatter plot with LOESS smooth
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() + 
  geom_smooth(method = "loess")


# A scatter plot with an ordinary Least Squares linear model
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() + 
  geom_smooth(method = "lm")


# The previous plot, without CI ribbon
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)


# The previous plot, without points
ggplot(mtcars, aes(x = wt, y = mpg)) +
geom_smooth(method = "lm", se = FALSE)

```


```{r}

# 1 - Define cyl as a factor variable
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

# 2 - Plot 1, plus another stat_smooth() containing a nested aes()
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() + 
  stat_smooth(method = "lm", se = FALSE) +
  stat_smooth(aes(group = 1), method = "lm", se = FALSE) 

```


## 1.2 Modifying stat_smooth (1)

```{r}
# Plot 1: change the LOESS span
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  # Add span below
  geom_smooth(se = FALSE, span = 0.7)

# Plot 2: Set the second stat_smooth() to use LOESS with a span of 0.7
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  # Change method and add span below
  stat_smooth(method = "loess", aes(group = 1),
              se = FALSE, col = "black", span = 0.7)

# Plot 3: Set col to "All", inside the aes layer of stat_smooth()
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  stat_smooth(method = "loess",
              # Add col inside aes()
              aes(group = 1, col = "All"),
              # Remove the col argument below
              se = FALSE, span = 0.7)

# Plot 4: Add scale_color_manual to change the colors
myColors <- c(RColorBrewer::brewer.pal(3, "Dark2"), "black")

ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE, span = 0.7) +
  stat_smooth(method = "loess", 
              aes(group = 1, col="All"), 
              se = FALSE, span = 0.7) +
  # Add correct arguments to scale_color_manual
  scale_color_manual("Cylinders", values = myColors)
```

The default span for LOESS is 0.9. A lower span will result in a better fit with more detail. 


## 1.3 Modifying stat_smooth (2)

```{r}
Vocab <- read.table("data/Vocab.txt") %>%  as.tibble()

# Plot 1: Jittered scatter plot, add a linear model (lm) smooth
ggplot(Vocab, aes(x = education, y = vocabulary)) +
  geom_jitter(alpha = 0.2) +
  stat_smooth(method = "lm", se = FALSE) # smooth

# Plot 2: points, colored by year
ggplot(Vocab, aes(x = education, y = vocabulary, col = year)) +
  geom_jitter(alpha = 0.2) 

# Plot 3: lm, colored by year
ggplot(Vocab, aes(x = education, y = vocabulary, col = factor(year))) +
  stat_smooth(method = "lm", se = FALSE) # smooth
  
# Plot 4: Set a color brewer palette
ggplot(Vocab, aes(x = education, y = vocabulary, col = factor(year))) +
  stat_smooth(method = "lm", se = FALSE) +  # smooth
  scale_color_brewer()  # colors

# Plot 5: Add the group aes, specify alpha and size
ggplot(Vocab, aes(x = education, y = vocabulary, col = year, group = factor(year))) +
  stat_smooth(method = "lm", se = FALSE, alpha = 0.6, size = 2) +
  scale_color_gradientn(colors = RColorBrewer::brewer.pal(9, "YlOrRd"))
```


## 1.4 Quantiles 

```{r}
# Use stat_quantile instead of stat_smooth
ggplot(Vocab, aes(x = education, y = vocabulary, col = year, group = factor(year))) +
  stat_quantile(alpha = 0.6, size = 2) +
  scale_color_gradientn(colors = RColorBrewer::brewer.pal(9,"YlOrRd"))

# Set quantile to 0.5
ggplot(Vocab, aes(x = education, y = vocabulary, col = year, group = factor(year))) +
  stat_quantile(alpha = 0.6, size = 2, quantiles = 0.5) +
  scale_color_gradientn(colors = RColorBrewer::brewer.pal(9,"YlOrRd"))


```


Quantile regression is a great tool for getting a more detailed overview of a large dataset.


## 1.5 stat_sum()

```{r}
# Plot 1: Jittering only
p <- ggplot(Vocab, aes(x = education, y = vocabulary)) +
  geom_jitter(alpha = 0.2)

# Plot 2: Add stat_sum
p +
  stat_sum() # sum statistic

# Plot 3: Set size range
p +
  stat_sum() + # sum statistic
  scale_size(range = c(1, 10)) # set size scale

```

## 1.6 stat_summary()

### 1.6.1 Preparations

Preparations: In this exercise we'll consider the preparations. That means we'll make sure the data is in the right format and that all the positions that we might use in our plots are defined. Lastly, we'll set the base layer for our plot. ggplot2 is already loaded, so you can get started straight away!

```{r}
# Display structure of mtcars
str(mtcars)

# Convert cyl and am to factors
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$am <-as.factor(mtcars$am)

# Define positions
posn.d <- position_dodge(width = 0.1)
posn.jd <-position_jitterdodge(jitter.width = 0.1, dodge.width = 0.2)
posn.j <- position_jitter(width = 0.2)

# Base layers
wt.cyl.am <- ggplot(data = mtcars, aes( x = cyl, y = wt, col = am, fill = am, group = am)) 
```

### 1.6.2 Plotting variations 


```{r}
# wt.cyl.am, posn.d, posn.jd and posn.j are available

# Plot 1: Jittered, dodged scatter plot with transparent points
wt.cyl.am +
  geom_point(position = posn.jd, alpha = 0.6)

# Plot 2: Mean and SD - the easy way

# Calculating statistics in Hmisc 
Hmisc::smean.sdl(mtcars$wt, mult = 1)
# Calculating statistics in ggplot2 
mean_sdl(mtcars$wt, mult = 1) 

# plot 
wt.cyl.am +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), position = posn.d)



# Plot 3: Mean and 95% CI - the easy way
ERR <- qt(p = 0.975, df = length(mtcars$wt)-1 ) * sd(mtcars$wt/sqrt(length(mtcars$wt)))
ERR

mean(mtcars$wt) + ERR * c(-1, 1) # very cool to -/+

# 95% ci calculation in Hmisc
Hmisc::smean.cl.normal(mtcars$wt)

# 95% ci calculatin in ggplot2
mean_cl_normal(mtcars$wt)

# plot
wt.cyl.am +
  stat_summary(fun.data = mean_cl_normal, position = posn.d)


# Plot 4: Mean and SD - with T-tipped error bars - fill in ___
wt.cyl.am +
  stat_summary(geom = "point", fun.y = mean,
               position = posn.d) +
  stat_summary(geom = "errorbar", fun.data = mean_sdl,
               position = posn.d, fun.args = list(mult = 1), width = 0.1)
```

 We can always specify our own function to the **fun.data** argument as long as the variable names match the aesthetics that you will need for the geom layer.


### 1.6.3 Custom Functions

In the video we saw that the only difference between ggplot2::mean_sdl() and Hmisc::smean.sdl() is the naming convention. In order to use the results of a function directly in ggplot2 we need to ensure that the names of the variables match the aesthetics needed for our respective geoms.

Here we'll create two new functions in order to create the plot shown in the viewer. One function will measure the full range of the dataset and the other will measure the interquartile range.



```{r}
## (a) generate xx 

xx <- seq(1, 100, 1)


# Function to save range for use in ggplot
gg_range <- function(x) {
  # Change x below to return the instructed values
  data.frame(ymin = min(x), # Min
             ymax = max(x)) # Max
}

gg_range(xx)
# Required output
#   ymin ymax
# 1    1  100

# Function to Custom function
med_IQR <- function(x) {
  # Change x below to return the instructed values
  data.frame(y = median(x), # Median
             ymin = quantile(x)[2], # 1st quartile
             ymax = quantile(x)[4])  # 3rd quartile
}

med_IQR(xx)
# Required output
#        y  ymin  ymax
# 25% 50.5 25.75 75.25


# The base ggplot command; 
wt.cyl.am <- ggplot(mtcars, aes(x = cyl, y = wt, col = am, fill = am, group = am))

# Add three stat_summary calls to wt.cyl.am
wt.cyl.am +
  stat_summary(geom = "linerange", fun.data = med_IQR,  
               position = posn.d, size = 3) +
  stat_summary(geom = "linerange", fun.data = gg_range,
               position = posn.d, size = 3,
               alpha = 0.4) +
  stat_summary(geom = "point", fun.y = median,
               position = posn.d, size = 3,
               col = "black", shape = "X")



```


# 2. Coordinates and Facets

## 2.1 Zooming in 


```{r}
# Basic ggplot() command, coded for you
p <- ggplot(mtcars, aes(x = wt, y = hp, col = am)) + geom_point() + geom_smooth()
p

# Add scale_x_continuous()
p + scale_x_continuous(limits = c(3, 6), expand = c(0, 0))

# Add coord_cartesian(): the proper way to zoom in
p + coord_cartesian(xlim = c(3, 6))

```

## 2.2 Aspect Ratio


```{r}
# Complete basic scatter plot function
base.plot <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, col = Species)) +
               geom_jitter() +
               geom_smooth(method = "lm", se = FALSE)

# Plot base.plot: default aspect ratio
base.plot 

# Fix aspect ratio (1:1) of base.plot
base.plot + coord_equal()
```

## 2.3 Pie Charts

```{r}
# Create a stacked bar plot: wide.bar
wide.bar <- ggplot(mtcars, aes(x = 1, fill = cyl)) +
              geom_bar()
wide.bar 

# Convert wide.bar to pie chart
wide.bar +
  coord_polar(theta = "y")


# Create stacked bar plot: thin.bar
thin.bar <- ggplot(mtcars, aes(x = 1, fill = cyl)) +
              geom_bar(width = 0.1) +
              scale_x_continuous(limits = c(0.5, 1.5))
thin.bar

# Convert thin.bar to "ring" type pie chart
thin.bar + 
  coord_polar(theta = "y")
```



## 3.4 Facet

```{r}
# Basic scatter plot
p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()

p

# 1 - Separate rows according to transmission type, am
p +
  facet_grid(am ~ .)

# 2 - Separate columns according to cylinders, cyl
p +
  facet_grid(. ~ cyl)

# 3 - Separate by both columns and rows 
p +
  facet_grid(am ~ cyl)

```


## 3.5 Handle many variables

```{r}
# Code to create the cyl_am col and myCol vector
mtcars$cyl_am <- paste(mtcars$cyl, mtcars$am, sep = "_")

library(RColorBrewer)

myCol <- rbind(brewer.pal(9, "Blues")[c(3,6,8)],
               brewer.pal(9, "Reds")[c(3,6,8)])

# Map cyl_am onto col
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl_am)) +
  geom_point() +
  # Add a manual colour scale
  scale_color_manual(values = myCol)

  
# Grid facet on gear vs. vs

ggplot(mtcars, aes(x = wt, y = mpg, col = cyl_am)) +
  geom_point() +
  # Add a manual colour scale
  scale_color_manual(values = myCol) +
  facet_grid(gear ~ vs)



# Also map disp to size
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl_am, size = disp)) +
  geom_point() +
  # Add a manual colour scale
  scale_color_manual(values = myCol) +
  facet_grid(gear ~ vs)



```


## 3.6 Dropping levels

```{r}

# Mammalian sleep dataset 

mamsleep <- read_csv("data/mamsleep.csv")

head(mamsleep)

# Basic scatter plot
p <- ggplot(mamsleep, aes(x = time, y = name, col = sleep)) +
  geom_point()
  
# Execute to display plot
p

# Facet rows accoding to vore
p +
  facet_grid(vore ~ .)

# Specify scale and space arguments to free up rows
p +
  facet_grid(vore ~ ., scale = "free_y", space = "free_y")


```



# 3. Themes 

**Objectives:**

* Learn to make theme pretty

* Enable us to make publication quality plots in R


What is **Themes Layer**?

* All the non-data ink

* Visual elements not part of data

* Three types: text element_text(), line element_line(), and rectangel element_rect(). 


## 3.1 Starting point: 

```{r}
z <- mtcars %>% 
  mutate(cyl = factor(cyl)) %>%
  ggplot(aes(x = wt, y = mpg, color = cyl)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ cyl, scales = "free") +
  scale_x_continuous(limits = c(0, 6)) + 
  scale_y_continuous(limits = c(10, 35)) +
  labs(x = "Weight (lb/1000)", y = "Miles/(US) gallon") +
  scale_color_manual(values = c("light blue", "blue", "dark blue")) 

# starting piont
z
```


## 3.2 Rectangles 

```{r}
# Plot 1: Change the plot background fill to myPink
myPink <- "#FEE0D2"

z +
  theme(plot.background = element_rect((fill = myPink)))

# Plot 2: Adjust the border to be a black line of size 3
z +
    theme(plot.background = element_rect(fill = myPink, color = "black", size = 3))

# Theme to remove all rectangles
no_panels <- theme(rect = element_blank())

# Plot 3: Combine custom themes
z2 <- z +
  no_panels +
  theme(plot.background = element_rect(fill = myPink, color = "black", size = 3))

z2
  
```

Background color has dramatically impact on plot's visual. 


## 3.3 Lines: 

To change the appearance of lines use the `element_line()` function. 

```{r}
z3 <- z2 +
  theme(panel.grid = element_blank(),   # removing excess lines to reduce non-data ink. 
        axis.line = element_line(color = "red"),
        axis.ticks = element_line(color = "red")) 

z3
 
```

## 3.4 Text

```{r}
myRed <- "#99000D"

z4 <- z3 +
  theme(strip.text = element_text(size = 16, color = myRed), 
        axis.title = element_text(color = myRed, hjust = 0, face = "italic"), 
        axis.text = element_text(color = "black"))

z4  
```


## 3.5 Legends 

```{r}

# Move legend by position
z4 +
  theme(legend.position = c(0.85, 0.85))


# Change direction
z4 +
  theme(legend.direction= "horizontal")


# Change location by name
z4 +
  theme(legend.position = "bottom")

# Remove legend entirely
z5 <- z4 +
  theme(legend.position = "none")

z5

```

## 3.6 Positions

```{r}
library(grid)

# Increase spacing between facets
z5 + 
  theme(panel.spacing.x = unit(2, "cm"))


# Adjust the plot margin
z5 + 
  theme(
    # add more spacing between the different facets
    panel.spacing.x = unit(2, "cm"), 
    # add space for top, right, bottom, and left
    plot.margin = unit(c(1, 2, 1, 1), "cm")) 

```


## 3.7 Recyling Themes 

* theme_set()

*theme_update()


## 3.8 ggthemes

ggplot2 has many themes: 

* theme_bw()

* theme_classic()

* theme_gray()


**How to use:** 

* theme_set(theme_bw())  - apply to all following plots

* + theme_bw() - apply on an individual plot


```{r}

library(ggthemes)

# Original plot
z5

# Load ggthemes
library(ggthemes)

# Apply theme_tufte(), plot additional modifications
custom_theme <- theme_tufte() +
  theme(legend.position = c(0.9, 0.9),
        legend.title = element_text(face = "italic",
        size = 12), 
        axis.title = element_text(face = "bold", size = 14))

# Draw the customized plot
z5 + custom_theme
 
# Use theme set to set custom theme as default
theme_set(custom_theme)

# Plot z2 again
z5


```


# 4. Best Practices

Three mostly discouraged plots in the data community: 

* heat maps

* pie charts

* dynamite plots - bar plots with error bars 



```{r}

# Base layers
m <- ggplot(mtcars, aes(x = cyl, y = wt))

# Draw dynamite plot
m +
  # any function can be specified in fun.data or fun.y. We can also specify any geom. 
  stat_summary(fun.y = mean, geom = "bar", fill = "skyblue") +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "errorbar", width = 0.1)


```


Draw a dynamite plot about the weight of the cars per cylinder type, add a distinction between transmission type, *am*, for the dynamite plots. 

```{r}
# Base layers
m <- ggplot(mtcars, aes(x = cyl,y = wt, col = factor(am), fill = factor(am)))

# Plot 1: Draw dynamite plot
m +
  stat_summary(fun.y = mean, geom = "bar") +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "errorbar", width = 0.1)

# Plot 2: Set position dodge in each stat function
m +
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "errorbar", width = 0.1, position = "dodge")

# Set your dodge posn manually
posn.d <- position_dodge(0.9)

# Plot 3: Redraw dynamite plot
m +
  stat_summary(fun.y = mean, geom = "bar", position = posn.d) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "errorbar", width = 0.1, position = posn.d)
  
```

If it is appropriate to use bar plots, then it would also be nice to give an impression of the number of values in each group. 

```{r}
library(tidyverse)
mtcars.cyl <- mtcars %>%
  group_by(cyl) %>%
  summarize(wt.avg = mean(wt), 
          sd = sd(wt), 
          n = n()) %>%
  ungroup() %>%
  mutate( prop = n/sum(n))

mtcars.cyl

# base layers
m <- ggplot(mtcars.cyl, aes(x = cyl, y = wt.avg))

# plot 1: draw bar plot with geom_bar
m + geom_bar(stat = "identity", fill = "skyblue")

# plot2: Draw bar plot with geom_col
m + geom_col(fill = "skyblue")

# plot 3: geom_col with variable widths
m + geom_col(fill = "skyblue", width = mtcars.cyl$prop)

# Plot 4: Add error bars 

m + geom_col(fill = "skyblue", width = mtcars.cyl$prop) +
  geom_errorbar(aes(ymin = wt.avg - sd, ymax = wt.avg + sd), width = 0.1)

```

**Pie Charts** 

```{r}
# bar chart
ggplot(mtcars, aes( x = cyl, fill = factor(am))) + 
  geom_bar(position = "fill")

head(mtcars)

```


# 5. Case Study

"In this case study, we’ll explore the large, publicly available California Health Interview Survey dataset from 2009. We’ll go step-by-step through the development of a new plotting method - a mosaic plot - combining statistics and flexible visuals. At the end, we’ll generalize our new plotting method to use on a variety of datasets we’ve seen throughout the first two courses."

California Health Information Survey Descriptive Statistics: 

* Largest state health survey in US

* Wide variety of variables

* Personal health adn economic measurements

## 5.1 Histogram 

```{r, eval = FALSE}
# Explore the dataset with summary and str
summary(adult)
str(adult)


# Age histogram
ggplot(adult, aes(x = SRAGE_P)) + 
geom_histogram()


# BMI value histogram
ggplot(adult, aes(x = BMI_P)) +
geom_histogram()


# Age colored by BMI, binwidth = 1
ggplot(adult, aes(x = SRAGE_P, fill = factor(RBMI))) + 
geom_histogram(binwidth = 1)

```

It looks like everyone 85 and above has been categorized as 85 years old. 

## 5.2 Data Cleaning 
```{r, eval=FALSE}
# Keep adults younger than or equal to 84
adult <- adult[adult$SRAGE_P <= 84, ] 

# Keep adults with BMI at least 16 and less than 52
adult <- adult[adult$BMI_P >= 16 & adult$BMI_P < 52, ]

# Relabel the race variable
adult$RACEHPR2 <- factor(adult$RACEHPR2, labels = c("Latino", "Asian", "African American", "White"))

# Relabel the BMI categories variable
adult$RBMI <- factor(adult$RBMI, labels = c("Under-weight", "Normal-weight", "Over-weight", "Obese"))
```

## 5.3 Multiple Histograms

```{r, eval = FALSE}
# The dataset adult is available

# The color scale used in the plot
BMI_fill <- scale_fill_brewer("BMI Category", palette = "Reds")

# Theme to fix category display in faceted plot
fix_strips <- theme(strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.1, size = 14),
                    strip.background = element_blank(),
                    legend.position = "none")

# Histogram, add BMI_fill and customizations
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) + 
  geom_histogram(binwidth = 1) +
  fix_strips + 
  BMI_fill +
  facet_grid(RBMI~.) + 
  theme_classic()

```


```{r, eval=FALSE}
# Plot 1 - Count histogram
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(binwidth = 1) +
  BMI_fill

# Plot 2 - Density histogram
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) + 
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  BMI_fill

# Plot 3 - Faceted count histogram
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(binwidth = 1) +
  BMI_fill + 
  facet_grid(RBMI~.)



# Plot 4 - Faceted density histogram
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) + 
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  BMI_fill +
  facet_grid(RBMI~.)



# Plot 5 - Density histogram with position = "fill"
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) + 
  geom_histogram(aes(y = ..density..), position = "fill", binwidth = 1) +
  BMI_fill



# Plot 6 - The accurate histogram
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) + 
  geom_histogram(aes(y = ..count../sum(..count..)), position = "fill", binwidth = 1) +
  BMI_fill


```


