---
title: "Visualizing Time Series Data in R"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


**Course Description**

"As the saying goes, “A chart is worth a thousand words”. This is why visualization is the most used and powerful way to get a better understanding of your data. After this course you will have a very good overview of R time series visualization capabilities and you will be able to better decide which model to choose for subsequent analysis. You will be able to also convey the message you want to deliver in an efficient and beautiful way."

Ref: Amsellem, Arnaud (2018) "Visualizing Time Series Data in R", https://www.datacamp.com/courses/visualizing-time-series-data-in-r, 2019.


Note: Some course materials and data have been revised for training by Jeff Li. 

# (I) Load required libraries
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(gmodels)

```

# 1. R Time Series Visualization Tools

*plot()* function 

```{r, eval=FALSE}
# Plot two charts on same graphical window
par(mfrow = c(2, 1))
plot(data$yahoo, main = "yahoo")
plot(data$microsoft, main = "microsoft")



# Replot with reduced margin and character sizes
par(mfrow = c(2, 1), mex = 0.6, cex = 0.8)
plot(data$yahoo, main = "yahoo")
plot(data$microsoft, main = "microsoft")

```


```{r, eval=FALSE}
plot(my_ts, main = "My Stocks")
lines(my_ts2, col = "red")
axis(side = 4, at = pretty(my_ts2))
legend(x = "bottomright", legend = c("Stock X", "Stock Y"), 
       col = c("black", "red"), lty = c(1, 1))

abline(v = as.Date("2016-04-14"))
abline(h = 1)
```


```{r, eval=FALSE}
library(PerformanceAnalytics)
period <- c("2014-01/2015-06")
chart.TimeSeries(my_ts, period.areas = period, main = "my_ts")

```


```{r, eval=FALSE}
# Plot the "microsoft" series
plot(data$microsoft, main = "Stock prices since 2015")

# Add the "dow_chemical" series in red
lines(data$dow_chemical, col = "red")

# Add a Y axis on the right side of the chart
axis(side = 4, at = pretty(data$dow_chemical))

# Add a legend in the bottom right corner
legend(x = "bottomright", 
       legend = c("microsoft", "dow_chemical"), 
       col = c("black", "red"),  
       lty = c(1, 1))
 
```



```{r, eval=FALSE}
# Plot the "citigroup" time series
plot(data$citigroup, main = "Citigroup")

# Create vert_line to identify January 4th, 2016 in citigroup
vert_line <- which(index(data$citigroup) == as.Date("2016-01-04"))

# Add a red vertical line using vert_line
abline(v = .index(data$citigroup)[vert_line], col = "red")

# Create hori_line to identify average price of citigroup
hori_line <- mean(data$citigroup)

# Add a blue horizontal line using hori_line
abline(h = hori_line, col = "blue")
```


```{r, eval=FALSE}
# Create period to hold the 3 months of 2015
period <- c("2015-01/2015-03")

# Highlight the first three months of 2015 
chart.TimeSeries(data$citigroup, period.areas = period)

# Highlight the first three months of 2015 in light grey
chart.TimeSeries(data$citigroup, period.areas = period, period.color = "lightgrey")
```


```{r, eval=FALSE}
# Plot the microsoft series
plot(data$microsoft, main = "Dividend date and amount")

# Add the citigroup series
lines(data$citigroup, col = "orange", lwd = 2 )

# Add a new y axis for the citigroup series
axis(side = 4, col = "orange", at = pretty(data$citigroup))
```

```{r, eval=FALSE}
# Same plot as the previous exercise
plot(data$microsoft, main = "Dividend date and amount")
lines(data$citigroup, col = "orange", lwd = 2)
axis(side = 4, at = pretty(data$citigroup), col = "orange")

# Create the two legend strings
micro <- paste0("Microsoft div. of ", "$0.39"," on ", "15 Nov. 2016")
citi <- paste0("Citigroup div. of ", "$0.16"," on ", "13 Nov. 2016")

# Create the legend in the bottom right corner
legend(x = "bottomright", legend = c(micro, citi), col = c("black", "orange"), lty = c(1, 1))

```



# 2. Univariate Time Series

Univariate plots are designed to learn as much as possible about the distribution, central tendency and spread of the data. 


## 2.1 Representing a univariate time series

The very first step in any time series analyses is to address if the time time series have the right mathematical properties to apply the standard statistical framework. If not, you much transform the time series first. 

```{r, eval=FALSE}
# Plot Apple's stock price 
plot(data, main = "Apple stock price")

# Create a time series called rtn
rtn <- ROC(data)

# Plot Apple daily price and daily returns 
par(mfrow = c(1, 2))
plot(data, main = "Apple stock price")
plot(rtn, main = "Apple daily returns")


```


Histogram with density lines

```{r, eval=FALSE}
# Create a histogram of Apple stock returns
hist(rtn, main = "Apple stock return distribution", probability = TRUE)


# Redraw a thicker, red density line
lines(density(rtn), col = "red", lwd = 2)
```


Box and Whisker plot 

```{r, eval=FALSE}
# Draw box and whisker plot for the Apple returns
boxplot(rtn, horizontal = TRUE, main = "Apple returns")

# Draw a box and whisker plot of a normal distribution
boxplot(rnorm(1000), horizontal = TRUE)

# Redraw both plots on the same graphical window
par(mfrow = c(2, 1))
boxplot(rtn, horizontal = TRUE)
boxplot(rnorm(1000), horizontal = TRUE) 

```

Autocorrelation: helps to understand time-lagged relationship in your data. 

```{r, eval=FALSE}
# Draw autocorrelation plot
acf(rtn, main = "Apple return autocorrelation")

# Redraw with a maximum lag of 10
acf(rtn, main = "Apple return autocorrelation", lag.max = 10)
```


q-q plot: 

```{r}
# Create q-q plot
qqnorm(rtn, main = "Apple return QQ-plot")

# Add a red line showing normality
qqline(rtn, col = "red")
```

A comprehensive time series diagnostic: 

```{r}
# Set up 2x2 graphical window
par(mfrow = c(2, 2))

# Draw histogram and add red density line
hist(rtn, probability = TRUE)
lines(density(rtn), col = "red")

# Draw box and whisker plot
boxplot(rtn)

# Draw autocorrelogram
acf(rtn)

# Draw q-q plot and add a red line for normality
qqnorm(rtn)
qqline(rtn, col = "red")

```


# 3. Multivariate Time Series

Learn how to identify pattern in the distribution, central tendency and spread over pairs or groups of data. 



# 4. Case study: visually selecting a stock that improves your existing portfolio

Let's put everything you learned so far in practice! Imagine you already own a portfolio of stocks and you have some spare cash to invest, how can you wisely select a new stock to invest your additional cash? Analyzing the statistical properties of individual stocks vs. an existing portfolio is a good way of approaching the problem.

```{r}
dir("data/")

data <-  read_csv("data/data_4_1.csv")


# plot the portfolio values
plot(data$value, main = "Portfolio Value")

# Plot the portfolio return
plot(data$return, main = "Portfolio Return")

# Plot a histogram of portfolio return
hist(data$return, probability = TRUE)

# Add a density line 
lines(density(data$return), col = "red", lwd = 2)
```


```{r}
data <- read_csv("data/data_4_3.csv")

# Plot the four stocks on the same graphical window
par(mfrow = c(2, 2), mex = 0.8, cex = 0.8)

plot(data$GS)
plot(data$KO)
plot(data$DIS)
plot(data$CAT)
```

