---
title: "Designing and Analyzing Clinical Trials in R"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee



**Course Description**

"Clinical trials are scientific experiments that are conducted to assess whether treatments are effective and safe. They are used by a variety of organizations, including pharmaceutical companies for drug development. Biostatisticians play a key role in ensuring the success of a clinical trial. In this course you will gain an overview of the important principles and a practical introduction to commonly used statistical analyses. This course would be valuable for data analysts, medical students, clinicians, medical researchers and others interested in learning about the design and analysis of clinical trials." 


Ref: Alfred, Tamuno. 2018. Designing and Analyzing Clinical Trials in R. https://www.datacamp.com/courses/designing-and-analyzing-clinical-trials-in-r. 


Note: Some course materials have been revised for internal R training. 

## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(samplesize)
library(gsDesign)

```

# 1. Principles



# 2. Trial Designs


# 3. Sample Size and Power

By the end of this chapter you will be able to calculate the numbers of patients needed for a clinical trial under a range of scenarios.

## 3.1 Sample size for comparing means
```{r}
#Generate the sample size for delta of 1, with SD of 3 and 80% power.
ss1 <- power.t.test(delta = 1, sd = 3, power = 0.8)
ss1

#Round up and display the numbers needed per group
ceiling(ss1$n)

#Use the sample size from above to show that it provides 80% power
power.t.test(n = 143, delta = 1, sd = 3)
```

## 3.2 Sample Size and Treatment Difference
```{r}
# Generate a vector containing values between 0.5 and 2.0, incrementing by 0.25
delta <- seq(from = 0.5, to = 2.0, by = 0.25)
npergp <- NULL # number per group

# Specify the standard deviation and power
for (i in 1:length(delta)) {
  npergp[i] <- ceiling(power.t.test(delta = delta[i], sd = 3, power = 0.80)$n)
}

# Create a data frame for the deltas and sample sizes
sample_sizes <- data.frame(delta, npergp)

# Plot the patients per group against the treatment differences
ggplot(sample_sizes, aes(x = delta, y = npergp)) +
  geom_line() +
  geom_point() +
  labs(title = "Sample Size Scenarios", x = "Treatment Difference", y = "Patients per Group")
```


## 3.3 Sample Size for Comparing Proportions

Use the power.prop.test() to calculate the sample size needed for a trial with a recovery percentage of 40% and 60% in the placebo and active treatment groups, respectively, and 80% power.

```{r}
# Use the power.prop.test to generate sample sizes for the proportions
power.prop.test(p1 = 0.4, p2 = 0.6, power = 0.8)

# Find the minimum detectable percentage ofr the above using 150 patients per group
power.prop.test(p1 = 0.4, power = 0.8, n = 150)$p2 * 100
```
It the placebo recovery percentage is 40%, then with 97 patients per group, we have 80% power to detect a treatment difference of 20%, i.e. 60% in the active arm. With 150 patients per arm, we can detect a smaller difference, ~16%, i.e. ~56% in the active arm. 


## 3.4 Sample Size for Unequal Groups

Calculate the sample size required for a trial where patients will be randomized to active treatment and placebo in a 2:1 ratio for a two-sided t-test. 

```{r}
# Use the *samplesize* library 
library(samplesize)

# Use 90% power, delta 1.5, standard deviation of 2.5, fraction of 0.5
unequalgps <- n.ttest(power = 0.9, alpha = 0.05, mean.diff = 1.5, sd1 = 2.5, sd2 = 2.5, k = 0.5, design = "unpaired", fraction = "unbalanced")

unequalgps
```

## 3.5 Sample Size for One-sided Tests
```{r}
# Generate sample sizes comparing the proportions using a two-sided test
two_sided <- power.prop.test(p1 = 0.1, p2 = 0.3, power = 0.80, alternative = "two.sided")
two_sided
ceiling(two_sided$n)

# Repeat using a one-sided test
one_sided <- power.prop.test(p1 = 0.1, p2 = 0.3, power = 0.80, alternative = "one.sided")
one_sided
ceiling(one_sided$n)

# Display the reduction per group
ceiling(two_sided$n) - ceiling(one_sided$n)

```

One sided test needs fewer patients. 



## 3.6 Stopping Rules

To avoid prolonging a study, a trial may be stopped early if there is strong evidence for efficacy at an interim analysis. Here we will derive the p-values needed to stop a trial early under the Pocock and O’Brien-Fleming spending functions.

```{r}
# Use the gsDesing function to generate the p-values for four analyses under the Pocock rule
Pocock <- gsDesign(k = 4, test.type = 2, sfu = "Pocock")
Pocock
2*(1-pnorm(Pocock$upper$bound))


# Repeat for the O'Brein & Fleming Rule
OF <- gsDesign(k = 4, test.type = 2, sfu = "OF")
OF
2 * (1 - pnorm(OF$upper$bound))
```

## 3.7 Sample Size Adjustments for Interim Analyses

A trial was originally planned to have no interim analyses. A sample size calculation estimated 500 patients were needed for 90% power at 5% siginificant level.

Here we will derive the new sample size requirements if three interim analyses are planned with the potential to stop early under the Pocock and O’Brien-Fleming spending functions. 

```{r}
# Use the gsDesign function to generate the sample sizes at each stage under the Pocock rule
Pocock.ss <- gsDesign(k = 4, test.type = 2, sfu = "Pocock", n.fix = 500, beta = 0.1)
ceiling(Pocock.ss$n.I)

# Repeat for the O'Brein-Feming rule
OF.ss <- gsDesign(k = 4, test.type = 2, sfu = "OF", n.fix = 500, beta = 0.1)
ceiling(OF.ss$n.I)
```
Under the Pcocock rule, we would need to increase the total sample size to 592 and it would be 512 if we followed the O'Brein & Fleming. 
