---
title: "String Manipulation in R with stringr"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


**Course Description**

"Character strings can turn up in all stages of a data science project. You might have to clean messy string input before analysis, extract data that is embedded in text or automatically turn numeric results into a sentence to include in a report. Perhaps the strings themselves are the data of interest, and you need to detect and match patterns within them. This course will help you master these tasks by teaching you how to pull strings apart, put them back together and use stringr to detect, extract, match and split strings using regular expressions, a powerful way to express patterns." 


Ref: Wickham, Charlotte. https://www.datacamp.com/courses/string-manipulation-in-r-with-stringr, 2018.

Note: Some materials have been adapted for company's internal training. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (I) Load Required Libraries
```{r, message = FALSE}
library(tidyverse)
library(purrr)
```


# 1. String basics 

* How to enter strings

* How to control numbers as strings

* Combine strings into sentences and tables


## 1.1 Quotes

Some guidelines to use double quotes ( " ) and single ( ' ). 

* Prefer double quotes ( " )

* Use single ( ' ) if string has  " inside

* Use " if string has " and ' with escape any double quotes inside the string ( \")


```{r}
  
# Line 1 -- double ( " ) 

line1 <- "The table was a larger one, but the three were all crowded together at one corner of it:"
line1

# Line 2 - sing ('), string has ( " ) inside

line2 <- ' "No room! No room!" they cried out when they saw Alice coming.'
line2

# Line 3 - use double ( "), string has both (") and ('), need to use escape (\)

line3 <- "\"There's plenty of room!\" said Alice indignantly, and she sat down in a large arm-chair at one end of the table."
line3


```


## 1.2 R print() vs writeLines()

* print() - displays strings as you might enter them 

* writeLines() - represent the real string, a great way to check the string you entered really does represent the string you wanted. 

```{r}
# putting lines in a vector
lines <- c(line1, line2, line3)

# print lines 
lines

# use writeLines()

writeLines(lines) 

# Write  lines with a space separator

writeLines(lines, sep = " ")

# what's going on 
writeLines("hello\n\U1F30D")

```


## 1.3 Escape sequences 

```{r}

# Should display: To have a \ you need \\
writeLines("To have a \\ you need \\\\")

# Should display: 
# This is a really 
# really really 
# long string
writeLines("This is a really\nreally really\nlong string")

# Use writeLines() with 
# "\u0928\u092e\u0938\u094d\u0924\u0947 \u0926\u0941\u0928\u093f\u092f\u093e"
writeLines("\u0928\u092e\u0938\u094d\u0924\u0947 \u0926\u0941\u0928\u093f\u092f\u093e")


```


## 1.4 format() with numbers

* format() - displayed in fixed (scientific = FAlSE)

* format() - scientific (scientific = TRUE)

The behavior of format() can be pretty confusing. 

```{r}
# Some vectors of numbers
percent_change  <- c(4, -1.91, 3.00, -5.002)
income <-  c(72.19, 1030.18, 10291.93, 1189192.18)
p_values <- c(0.12, 0.98, 0.0000191, 0.00000000002)

# Format c(0.0011, 0.011, 1) with digits = 1
format(c(0.0011, 0.011, 1), digits = 1)

# Format c(1.0011, 2.011, 1) with digits = 1
format(c(1.0011, 2.011, 1), digits = 1)

# Format percent_change to one place after the decimal point
format(percent_change, digits = 2)

# Format income to whole numbers
format(income, digits = 2)

# Format p_values in fixed format
format(p_values, scientific = FALSE)
```

## 1.5 formatC() 

Arguments for formatC()

* "f" for fixed

* "e" for scientific

* "g" for fixed unless scientific saves space


the *digits* arguments: the number of digits after the decimal point, which is more predictable than *format()*

```{r}
# From the format() exercise
x <- c(0.0011, 0.011, 1)
y <- c(1.0011, 2.011, 1)

# formatC() on x with format = "f", digits = 1
formatC(x, format = "f", digits = 1)

# formatC() on y with format = "f", digits = 1
formatC(y, format = "f", digits = 1)

# Format percent_change to one place after the decimal point
formatC(percent_change, format = "f", digits = 1)

# percent_change with flag = "+"
formatC(percent_change, format = "f", digits = 1, flag = "+")

# Format p_values using format = "g" and digits = 2
formatC(p_values, format = "g", digits = 2)
```


## 1.6 annotation of numbers

```{r}
# Add $ to pretty_income
income <- c(72, 1030, 10292, 1189192)
pretty_income <- format(income, digits = 2, big.mark = ",")

paste("$", pretty_income, sep = "")
 
# Add % to pretty_percent
pretty_percent <- c("+4.0", "-1.9", "+3.0", "-5.0")
paste(pretty_percent, "%", sep = "")

# Create vector with elements like 2010: +4.0%`
years <- c(2010, 2011, 2012, 2013)
year_percent <- paste(years, ": ", pretty_percent, "%", sep = "")

# Collapse all years into single string
paste(years, ": ", pretty_percent, "%", sep = "", collapse = ", ")
```


## 1.7 A very simple table 

```{r}
# Define the names vector
income_names <- c("Year 0", "Year 1", "Year 2", "Project Lifetime")

# Create pretty_income
pretty_income <- format(income, digits = 2, big.mark = ",")

# Create dollar_income
dollar_income <- paste("$", pretty_income, sep = "")

# Create formatted_names
formatted_names <- format(income_names, justify = "right")

# Create rows
rows <- paste(formatted_names, dollar_income, sep = "   ")

# Write rows
writeLines(rows)
```

## 1.8 Order pizza with three toppings
```{r}
# Randomly sample 3 toppings
my_toppings <- c("anchoview", "bacon", "cheese")

# Print my_toppings
print(my_toppings)

# Paste "and " to last element: my_toppings_and
my_toppings_and <- paste(c("", "", "and "), my_toppings, sep = "")

# Collapse with comma space: these_toppings
these_toppings <- paste(my_toppings_and, collapse = ", ")

# Add rest of sentence: my_order
my_order <- paste("I want to order a pizza with ", these_toppings, ".",  sep = "")

# Order pizza with writeLines()
writeLines(my_order)
```



# 2. Introduction to *stringr*

## 2.1 str_c()  

str_c: works like paste(), takes vectors of strings as input along with *sep* and  *collapse* arguments

how to handle missing values? 

* paste(): turns missing values into the string "NA"

* str_c(): propagates missing values. combining any strings with a missing value will result in another missing value. 

```{r}
library(stringr)

my_toppings <- c("cheese", NA, NA)
my_toppings_and <- paste(c("", "", "and "), my_toppings, sep = "")

# Print my_toppings_and
print(my_toppings_and)

# Use str_c() instead of paste(): my_toppings_str
my_toppings_str <- str_c(c("", "", "and "), my_toppings)

# Print my_toppings_str
print(my_toppings_str)

# paste() my_toppings_and with collapse = ", "
paste(c("", "", "and "), my_toppings, collapse =", ")

# str_c() my_toppings_str with collapse = ", "
str_c(c("", "", "and "), my_toppings, collapse = ", ")

```


## 2.2 String length 

str_length() - takes a vector of strings as input and returns the number of characters in each strings. It is better than *nchar()*. 


```{r}
library(stringr)
library(babynames)
library(dplyr)
library(ggplot2)

# Extracting vectors for boys' and girls' names
babynames_2014 <- filter(babynames, year == 2014)
boy_names <- filter(babynames_2014, sex == "M")$name
girl_names <- filter(babynames_2014, sex == "F")$name

# Take a look at a few boy_names
head(boy_names)

# Find the length of all boy_names
boy_length <- str_length(boy_names)

# Take a look at a few lengths
head(boy_length)

# Find the length of all girl_names
girl_length <- str_length(girl_names)

# Find the difference in mean length
mean(girl_length) - mean(boy_length)

# Confirm str_length() works with factors
head(str_length(factor(boy_names)))


# 
names_length <- bind_rows(data.frame(names_length = boy_length), 
          data.frame(names_length = girl_length), 
          .id = "sex")

names_length %>%
  ggplot(aes(x = names_length, fill = factor(sex, labels = c("boy", "girl")))) +
  geom_density(alpha = 0.3 )

names_length %>%
  ggplot(aes(x = names_length, fill = as.factor(sex))) +
  geom_histogram(alpha = 0.3, position = "dodge", binwidth = 0.2)

```


## 2.3 Extracting substrings

str_sub(string, start = 1L, end = -1L) 

str_sub(string, start = 1L, end = -1L, omit_na = FALSE) <- value


```{r}
# Extract first letter from boy_names
boy_first_letter <- str_sub(boy_names, 1, 1)

# Tabulate occurrences of boy_first_letter
table(boy_first_letter)
  
# Extract the last letter in boy_names, then tabulate
boy_last_letter <- str_sub(boy_names, -1, -1)
table(boy_last_letter)

# Extract the first letter in girl_names, then tabulate
girl_first_letter <- str_sub(girl_names, 1, 1)
table(girl_first_letter)

# Extract the last letter in girl_names, then tabulate
girl_last_letter <- str_sub(girl_names, -1, -1)
table(girl_last_letter)
  
```


## 2.4 Detecting matches

*str_detect()* detects if a string contains a *pattern*. 

```{r}

# a sample

my_string <- c("zz", "my_zz", "no no", "a really good")

str_detect(my_string, pattern = fixed("zz"))


# Look for pattern "zz" in boy_names
contains_zz <- str_detect(boy_names, pattern = fixed("zz"))

# Examine str() of contains_zz
str(contains_zz)

# How many names contain "zz"?
sum(contains_zz)

# Which names contain "zz"?
boy_names[contains_zz]

# Which rows in boy_df have names that contain "zz"?

babynames %>%
  filter(sex == "M" ) %>%
  filter(str_detect(name, pattern = fixed("zz")))

```

## 2.5 Subsetting strings based on match

```{r}
# find boy_names that contain "zz"

head(boy_names)
str_subset(boy_names, pattern = fixed("zz"))

# Find girl_names that contain "zz"
str_subset(girl_names, pattern = fixed("zz"))

# Find girl_names that contain "U"
starts_U <- str_subset(girl_names, pattern = fixed("U"))
starts_U

# Find girl_names that contain "U" and "z"
str_subset(starts_U, pattern = fixed("z"))

```

## 2.6 Counting matches

*str_count()*: how many times does the pattern occur in each string? 


```{r}
# Count occurrences of "a" in girl_names
number_as <- str_count(girl_names, pattern = fixed("a"))

# Count occurrences of "A" in girl_names
number_As <- str_count(girl_names, pattern = fixed("A"))

# Histograms of number_as and number_As
hist(number_as)
hist(number_As)
 

# Find total "a" + "A"
total_as <- number_as + number_As

# girl_names with more than 4 a's
girl_names[total_as > 4]

```

## 2.7 Parsing strings into variables


```{r}
# Some date data
date_ranges <- c("23.01.2017 - 29.01.2017", "30.01.2017 - 06.02.2017")

# Split dates using " - "
split_dates <- str_split(date_ranges, pattern = fixed(" - "))
split_dates

```

## 2.8 Some simple text statistics 

```{r}
# Split lines into words
words <- str_split(lines, pattern = fixed(" "))

# Number of words per line
lapply(words, length)

# a better choice 
map_int(words, length)
  
# Number of characters in each word
word_lengths <- lapply(words, str_length)

word_lengths <- map(words, str_length)  # a better option
word_lengths

  
# Average word length per line
lapply(word_lengths, mean)

map_dbl(word_lengths, mean) # a better function 

```


## 2.9 Replacing to tidy strings

*str_replace()*

*str_replace_all()*


```{r}
# Some IDs
ids <- c("ID#: 192", "ID#: 118", "ID#: 001")

# Replace "ID#: " with ""
id_nums <- str_replace(ids, "ID#: ", "")
id_nums

# Turn id_nums into numbers
id_ints <- as.numeric(id_nums)
id_ints
```

## 2.10 DNA Sequences

```{r}
# Read rds data
genes <- read_rds("data/dna.rds")

head(genes)

summary(genes)

# Find the number of nucleotides in each sequence
str_length(genes)


# Find the number of A, T, C, and G's occur in each sequence
ATCG_num <- data.frame( A = str_count(genes, pattern = fixed("A")),
            T = str_count(genes, pattern = fixed("T")), 
            C = str_count(genes, pattern = fixed("C")), 
            G = str_count(genes, pattern = fixed("G")))

row.names(ATCG_num) <- c("YPO0001", "asnC", "asnA")

ATCG_num

# Return the sequences that contain "TTTTTT"
str_subset(genes, pattern = fixed("TTTTTT"))

# Replace all the "A"s in the sequences with a "_"
str_replace_all(genes, pattern = fixed("A"), "_")
```

## 2.11 Final Challenges

```{r}
# Define some full names
names <- c("Diana Prince", "Clark Kent")

# Split into first and last names
names_split <- str_split(names, pattern = " ", simplify = TRUE)

# Extract the first letter in the first name
abb_first <- str_sub(names_split[,1], 1, 1 )

# Combine the first letter ". " and last name
str_c(abb_first, ". ", names_split[, 2])


```


# 3. Regular expressions 

What is *regular expressions*? - A language for describing patterns. For example,  *^.[\d]+* - The start of the string, followed by any single character, followed by one or more digits. 

## 3.1 Matching the start or end of the string

```{r}
library(rebus)
# Some strings to practice with
# Match two characters, where the second is a "t"
str_view(x, pattern = ANY_CHAR %R% "t")x <- c("cat", "coat", "scotland", "tic toc")

# Print END
END

# Run me
str_view(x, pattern = START %R% "c")

# Match two characters, where the second is a "t"
str_view(x, pattern = ANY_CHAR %R% "t")


pattern <- "q" %R% ANY_CHAR

# Find names that have the pattern
names_with_q <- str_subset(boy_names, pattern = pattern)

# How many names were there?
length(names_with_q)
```

## 3.2 Alternation

* or() in *rebus* - specify a set of alternatives

```{r}
# Match Jeffrey or Geoffrey
whole_names <- or("Jeffrey", "Geoffrey")
str_view(boy_names, pattern = whole_names, match = TRUE)

```

## 3.3 Character classes

```{r}
# Create character class containing vowels
vowels <- char_class("aeiouAEIOU")

# Print vowels
vowels

# See vowels in x with str_view()
str_view(x, pattern = vowels)


# Number of vowels in boy_names
num_vowels <- str_count(boy_names, pattern = vowels)

# Proportion of vowels in boy_names
name_length <- str_length(boy_names)

# Calc mean number of vowels
mean(num_vowels)

# Calc mean fraction of vowels per name
mean(num_vowels / name_length)
```


## 3.4 Repetition

* *one_or_more()*

* *zero_or_more()*

* *optional()*  

```{r}
# Vowels from last exercise
vowels <- char_class("aeiouAEIOU")


one_or_more(vowels)
zero_or_more(vowels)
optional(vowels)

x <- c("ow", "ooh", "yeeeah!", "shh")
str_view(x, pattern = one_or_more(vowels))

str_detect(x, pattern = one_or_more(vowels))
str_subset(x, pattern = one_or_more(vowels))
str_subset(x, pattern = optional(vowels))

# See names with only vowels
str_view(boy_names, 
  pattern = exactly(one_or_more(vowels)), 
  match = TRUE)


# Use `negated_char_class()` for everything but vowels
not_vowels <- negated_char_class("aeiouAEIOU")

# See names with no vowels
str_view(boy_names, 
  pattern = exactly(one_or_more(not_vowels)), 
  match = TRUE)

```

## 3.5 Hunting for phone numbers

```{r}
# create a three digit patter

three_digits <- DIGIT %R% DIGIT %R% DIGIT

# test it 
contact <- c("Call me at 555-555-7890", 
             "123 Main ST", 
             "(555) 555 0191 mobile: 123.345.4789")
str_view_all(contact, pattern = three_digits)


# Create a separator pattern
(separator <- char_class("-.() "))
str_view_all(contact, pattern = separator)

str_split(contact, pattern = separator, simplify = TRUE)


# Use these components
(three_digits <- DIGIT %R% DIGIT %R% DIGIT)
(four_digits <- three_digits %R% DIGIT)
(separator <- char_class("-.() "))

# create phone pattern 
# Create phone pattern
phone_pattern <- optional(OPEN_PAREN) %R%
  three_digits %R%
  zero_or_more(separator) %R%
  three_digits %R% 
  zero_or_more(separator) %R%
  four_digits

phone_pattern


# test it 
str_view_all(contact, pattern = phone_pattern)

# Extract phone numbers
str_extract(contact, pattern = phone_pattern)

# Extract ALL phone numbers
str_extract_all(contact, pattern = phone_pattern)

```

## 3.6 Extract age and gender from accident narratives

```{r}
# read narratives data
narratives <- read_rds("data/narratives.rds")
head(narratives)

# Pattern to match one or two digits
(age <- or(DGT, DGT %R% DGT))

age <- DGT %R% optional(DGT)
age

# test it
str_view(narratives, pattern = age)


# Use these patterns
age <- DGT %R% optional(DGT)

age < dgt(1, 2) # use this to match one or two digits
  
unit <- optional(SPC) %R% or("YO", "YR", "MO")
gender <- optional(SPC) %R% or("M", "F")

# Extract age, unit, gender
str_extract(narratives, age %R% unit %R% gender)
```

## 3.7 Parsing age and gender into pieces

Several ways: 

* extract out the piece we need

* replace everything that isn't the piece we need with ""

* split into the pieces


Note: *stringr* has *str_remove*, which works like str_replace() with replacement = "". 

```{r}
# Extract age and make numeric
as.numeric(str_extract(narratives, pattern = age))

# extract age_gender

age_gender <- str_extract(narratives, pattern = age %R% unit %R% gender)

# Replace age and units with ""
genders <- str_remove(age_gender, pattern = age %R% unit)

# Replace extra space
str_remove_all(genders, pattern = one_or_more(SPC))



# Numeric ages, from previous step
ages_numeric <- as.numeric(str_extract(age_gender, age))

# Extract units 
time_units <- str_extract(age_gender, pattern = unit)

# Extract first word character
time_units_clean <- str_extract(time_units, pattern = WRD )

# Turn ages in months to years
ifelse(time_units_clean == "Y", ages_numeric, ages_numeric/12)
```


# 4. More Advanced Matching and Manipulation

* Select parts of a match (a.k.a capturing)

* Referring back to parts of a match (a.k.a back-referencing)



## 4.1 Capturing parts of a pattern 

* *str_match()*: returns a character matrix. First column is the complete match, followed by one column for reach capture group. 

* *str_match_all()*: return a list of character matrices. 


```{r}
# capture parts between @ and . and after . 

email <- capture(one_or_more(WRD)) %R%
  "@" %R% capture(one_or_more(WRD)) %R%
  DOT %R% capture(one_or_more(WRD))

email

# 
hero_contacts <- c("(wolverine@xmen.com)",  "wonderwoman@justiceleague.org", "thor@avengers.com", "my123@yahoo.com")

# check match hasn't changed
str_view(hero_contacts, pattern = email)



# Pull out match and captures
email_parts <- str_match(hero_contacts, pattern = email)
email_parts

# Save host
host <- email_parts[, 3]
host

# save ID
email_id <- email_parts[, 2]
email_id
```


## 4.2 Pulling out parts of a phone number

```{r}
# create a text containing phone numbers
contact <- c("Call me at 919-555-7890", 
             "123 Main St", 
             "(517) 321 0191" , 
             "Phone: 923.555.3256 Mobile: 876.478.4567")

# Add capture() to get digit parts
phone_pattern <- capture(three_digits) %R% zero_or_more(separator) %R% capture(three_digits) %R% zero_or_more(separator) %R% capture(four_digits)

# pull out the parts with str_match()

phone_numbers <- str_match(contact, pattern = phone_pattern)
phone_numbers

str_match_all(contact, pattern = phone_pattern)
 
```


## 4.3 Extracting age and gender again

One step to extract age and gender. 

```{r}
head(narratives)

# build a pattern and add capture()

pattern <- capture(optional(DGT) %R% DGT) %R%
  optional(SPC) %R% capture(or("YO", "YR", "MO")) %R%
  optional(SPC) %R% capture(or("M", "F"))

#  pull out from narratives

str_match(narratives, pattern = pattern) %>% 
  as.tibble() %>%
  mutate(V3 = str_extract(V3, pattern = WRD),
         V2 = ifelse(V3 == "Y", as.numeric(V2), as.numeric(V2)/12), 
         V4 = factor(V4, levels = c("M", "F"), labels = c("Male", "Female"))) %>%
  select(sex = V4, age_year = V2, raw = V1)



# Edit to capture just Y and M in units
pattern2 <- capture(optional(DGT) %R% DGT) %R%  
  optional(SPC) %R% capture(or("Y", "M")) %R% optional(or("O","R")) %R%
  optional(SPC) %R% capture(or("M", "F"))

# Check pattern
str_view(narratives, pattern = pattern2)

# Pull out pieces
str_match(narratives, pattern = pattern2)

```

How powerful it is the combination of *capture()* and *str_math()* ! 


## 4.4 Using back references in patterns

Back references can be useful in matching because they allow you to find repeated patterns or words. Using a back reference requires two things: you need to capture() the part of the pattern you want to reference, and then you refer to it with REF1.

```{r}

# It matches and captures any lower case character, then is followed by the captured character: it detects repeated characters regardless of what character is repeated. 

str_view(c("hello", "sweet", "kitten"), pattern = capture(LOWER) %R% REF1)
```

```{r}
# All to lower case
boy_names <- map_chr(boy_names, tolower)

# names with three repeated letters
repeated_three_times <- capture(LOWER) %R% REF1 %R% REF1

# test it
str_view(boy_names, pattern = repeated_three_times, match = TRUE)


# Names with a pair of repeated letters
pair_of_repeated <- capture(LOWER %R% LOWER) %R% REF1

# Test it
str_view(boy_names, pattern = pair_of_repeated, match = TRUE)

# Names with a pair that reverses
pair_that_reverses <- capture(LOWER) %R% capture(LOWER) %R% REF2 %R% REF1

# Test it
str_view(boy_names, pattern = pair_that_reverses, match = TRUE)


# Four letter palindrome names
four_letter_palindrome <- exactly(capture(LOWER) %R% capture(LOWER) %R% REF2 %R% REF1)

# Test it
str_view(boy_names, pattern = four_letter_palindrome, match = TRUE)

```


## 4.5 Replacing with regular expressions

**str_replace()**: 

* first argument - string a vector of strings to do the replacements in

* second argument - pattern that identifies the parts of strings to replace

* third arguments - replacement the thing to use as a replacement.

```{r}
args(str_replace)
```

```{r}
# View text containing phone numbers
contact

# Replace digits with "X"
str_replace(contact, DGT, "X")

# Replace all digits with "X"
str_replace_all(contact, DGT, "X")

# Replace all digits with different symbol
str_replace_all(contact, DGT, c("X", ".", "*", "_"))
```


## 4.6 Replacing with back references 
```{r}
x <- c("hello", "sweet", "kitten")
str_replace(x, capture(ANY_CHAR), str_c(REF1, REF1))
```

```{r}
# Build pattern to match words ending in "ING"
pattern <- one_or_more(WRD) %R% "ING"
str_view(narratives, pattern)

# Test replacement
str_replace(narratives, capture(pattern), str_c("CARELESSLY", REF1, sep = " "))

# One adverb per narrative
adverbs <- read_rds("data/adverbs.rds")
adverbs_10 <- sample(adverbs, 10)

# Replace "***ing" with "adverb ***ly"
str_replace(narratives, capture(pattern), str_c(adverbs_10, REF1, sep = " "))
```


## 4.7 Matching a specific code point or code groups

```{r}
library(stringi)
x <- c("\u00e8", "\u0065\u0300")
writeLines(x)

str_view(x, "\u00e8")

as.hexmode(utf8ToInt(stringi::stri_trans_nfd("\u00e8")))
as.hexmode(utf8ToInt(stringi::stri_trans_nfc("\u0065\u0300")))
```


```{r}
# Names with builtin accents
(tay_son_builtin <- c(
  "Nguy\u1ec5n Nh\u1ea1c", 
  "Nguy\u1ec5n Hu\u1ec7",
  "Nguy\u1ec5n Quang To\u1ea3n"
))

# Convert to separate accents
tay_son_separate <- stri_trans_nfd(tay_son_builtin)

# Verify that the string prints the same
tay_son_separate

# Match all accents
str_view_all(tay_son_separate, UP_DIACRITIC)
```

## 4.8 Matching a single grapheme

```{r}
x <- c("Adele", "Ad\u00e8le", "Ad\u0065\u0300le")
writeLines

str_view(x, "Ad" %R% ANY_CHAR %R% "le")

str_view(x, "Ad" %R% GRAPHEME %R% "le")
```

```{r}
# tay_son_separate has been pre-defined
tay_son_separate

# View all the characters in tay_son_separate
str_view_all(tay_son_separate, pattern = ANY_CHAR )

# View all the graphemes in tay_son_separate
str_view_all(tay_son_separate, GRAPHEME)
```

```{r}
# Combine the diacritics with their letters
tay_son_builtin <- stri_trans_nfc(tay_son_separate)
tay_son_builtin

# View all the graphemes in tay_son_builtin
str_view_all(tay_son_builtin, GRAPHEME)
```


# 5. Case Study

"The truth is rarely pure and never simple."

Task: Read the play and count the number of lines each character has. 

tool box: 

* readLines() 

* stringi::stri_read_lines() 


## 5.1 Getting the play into R


```{r}
dir("data/")

# Read play in using stri_read_lines()

earnest <- stri_read_lines("data/importance-of-being-earnest.txt")
head(earnest)
tail(earnest)
summary(earnest)


# Detect start and end lines
start <- str_which(earnest, "START OF THE PROJECT")
end <- str_which(earnest, "END OF THE PROJECT")

# Get rid of gutenberg intro text
earnest_sub  <- earnest[(start + 1):(end - 1)]


# Detect first act
lines_start <- str_which(earnest_sub, "FIRST ACT")

# Set up index
intro_line_index <- 1:(lines_start - 1)

# Split play into intro and play
intro_text <- earnest_sub[intro_line_index]
play_text <- earnest_sub[-intro_line_index]

# Set up index
intro_line_index <- 1:(lines_start - 1)


# Take a look at the first 20 lines
writeLines(play_text)




# Get rid of empty strings
empty <- stri_isempty(earnest)
sum(empty)

play_lines <- earnest[!empty]
summary(play_lines)



# Pattern for start, word then .
pattern_1 <- START %R% one_or_more(WRD) %R% DOT

# Test pattern_1
str_view(play_lines, pattern_1, match = TRUE) 
str_view(play_lines, pattern_1, match = FALSE)


# Pattern for start, capital, word then .
pattern_2 <-START %R% ascii_upper() %R% one_or_more(WRD) %R% DOT


# Test pattern_2
str_view(play_lines, pattern_2, match = TRUE)
str_view(play_lines, pattern_2, match = FALSE)

# Get subset of lines that match
lines <- str_subset(play_lines,pattern = pattern_2 )
lines[10:15]
summary(lines)

# Extract match from lines
who <- str_extract(lines, pattern = pattern_2)
sample(who, 5)

# Let's see what we have
unique(who)
table(who)

message("Missed Lady Brackness, picked up lines starting with University, Sunday, States, Rev. ")

```

## 5.2 Identifying the lines take 2

The pattern "starts with a capital letter, has some other characters then a full stop" wasn't specific enough. You ended up matching lines that started with things like University., July., London., and you missed characters like Lady Bracknell and Miss Prism.

Let's take a different approach. You know the characters names from the play introduction. So, try specifically looking for lines that start with their names. You'll find the or1() function from the rebus package helpful. It specifies alternatives but rather than each alternative being an argument like in or(), you can pass in a vector of alternatives.

```{r}
# Create vector of characters
characters <- c("Algernon", "Jack", "Lane", "Cecily", "Gwendolen", "Chasuble", "Merriman", "Lady Bracknell", "Miss Prism")

# Match start, then character name, then .
pattern_3 <- START %R% or1(characters) %R% DOT
pattern_3

# View matches of pattern_3
str_view(play_lines, pattern_3, match = TRUE)
  
# View non-matches of pattern_3
str_view(play_lines, pattern_3, match = FALSE)


# Pull out matches
lines <- str_subset(play_lines, pattern = pattern_3)
lines[3:4]

# Extract match from lines
who <- str_extract(lines, pattern = pattern_3)

# Let's see what we have
unique(who)

# Count lines per character
table(who)


```

Algernon and Jack get the most lines, more than ten times more than Merriman who has the fewest. If you were looking really closely you might have noticed the pattern didn't pick up the line Jack and Algernon [Speaking together.] which you really should be counting as a line for both Jack and Algernon. One solution might be to look for these "Speaking together" lines, parse out the characters, and add to your counts.


## 5.3 Changing case to ease matching

* str_to_lower(x) 

* whole_word() 

```{r}
# read cat-related accidents
catcidents <- read_rds("data/catcidents.rds")

# catcidents has been pre-defined
head(catcidents)

# Construct pattern of DOG in boundaries
whole_dog_pattern <- whole_word("DOG")
whole_dog_pattern

# See matches to word DOG
str_view(catcidents, whole_dog_pattern, match = TRUE)

# Transform catcidents to upper case
catcidents_upper <- str_to_upper(catcidents)

# View matches to word "DOG" again
str_view(catcidents_upper, whole_dog_pattern, match = TRUE)

# Which strings match?
has_dog <- str_detect(catcidents_upper, whole_dog_pattern)
has_dog[3:7]

# Pull out matching strings in original 
catcidents[has_dog]

```


## 5.4 Ignoring case when matching

```{r}
x <- c("Cat", "CAT", "cAt") 
str_view(x, "cat")

str_view(x, regex("cat", ignore_case = TRUE), match = TRUE)

# View matches to "TRIP"
str_view(catcidents, "TRIP", match = TRUE)

# Construct case insensitive pattern
trip_pattern <- regex("TRIP", ignore_case = TRUE)

# View case insensitive matches to "TRIP"
str_view(catcidents, trip_pattern, match = TRUE)


# Get subset of matches
trip <- str_subset(catcidents, trip_pattern)

# Extract matches
str_extract(trip, trip_pattern)
```

## 5.5 Fixing case problems

* str_to_upper()

* str_to_lower()

* str_to_title()

* stri_trans_totitle() - has type = "word" or type = "sentence"

```{r}
library(stringi)

# Get first five catcidents
cat5 <- catcidents[1:5]

# Take a look at original
writeLines(cat5)

# Transform to title case
writeLines(str_to_title(cat5))

# Transform to title case with stringi
writeLines(stri_trans_totitle(cat5, type = "word"))

# Transform to sentence case with stringi
writeLines(stri_trans_totitle(cat5, type = "sentence"))
```

