---
title: "Importing and Managing Financial Data in R"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

**Course Description**

"If you've ever done anything with financial or economic time series, you know the data come in various shapes, sizes, and periodicities. Getting the data into R can be stressful and time-consuming, especially when you need to merge data from several different sources into one data set. This course will cover importing data from local files as well as from internet sources."


Ref: Ulrich, Joshua. "Importing and Managing Financial Data in R". https://www.datacamp.com/courses/importing-and-managing-financial-data-in-r. 2018. 

Note: Some course materials have been revised for internal training by Jeff Li. 


## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# (b) I like to have the package of "tidyverse"
library(tidyverse)
library(quantmod)
library(Quandl)
```


# 1. Introduction and downloading data

A wealth of financial and economic data are available online. Learn how getSymbols() and Quandl() make it easy to access data from a variety of sources.

What is getSymbols()? 

* Provides access to multiple data sources

* Return xts object by default

* Return data like an ordinary function or create an object like load() does in base R

```{r}
# getSymbols() example
aapl <- getSymbols(Symbols = "AAPL", src = "yahoo", auto.assign = FALSE)
head(aapl)

getSymbols(Symbols = "AAPL", src = "yahoo", auto.assign = TRUE)
head(AAPL)
```

## 1.1 Introducing getSymbols()

The getSymbols() function from the quantmod package provides a consistent interface to import data from various sources into your workspace. By default, getSymbols() imports the data as a xts object.

import QQQ data from Yahoo! Finance. QQQ is an exchange-traded fund that tracks the Nasdaq 100 index, and Yahoo! Finance is the default data source for getSymbols().

```{r}
# Import QQQ data from Yahoo! Finance
getSymbols(Symbols = "QQQ", auto.assign = TRUE)

# Look at the structure of the object getSymbols created
str(QQQ)

# Look at the first few rows of QQQ
head(QQQ)

# plot 
plot(QQQ)
```

## 1.2 Data Sources

Import data from __Alpha Vantage__ and __FRED__. __Alpha Vantage__ is a source similar to Yahoo! Finance. __FRED__ is an online database of economic time series data created and maintained by the Federal Reserve Bank of St. Louis.

```{r}
# Import QQQ data from Alpha Vantage
getSymbols(Symbols = "QQQ", 
           src = "av", 
           auto.assign = TRUE, 
           api.key = Sys.getenv("av_API"))

head(QQQ)

# Look at the structure of QQQ
str(QQQ)

# Plot QQQ
plot(QQQ)

# Import GDP data from FRED
getSymbols(Symbols = "GDP", 
           src = "FRED", 
           auto.assign = TRUE)

# Look at the structure of GDP
str(GDP)

# look at the  head of GDP
head(GDP)

# Plot GDP
plot(GDP)
```

## 1.3 Make *getSymbols() return the data it retrieves
```{r}
# Assign SPY data to *spy* using auto.assign argument
spy <- getSymbols(Symbols = "SPY", auto.assign = FALSE)

# Look at the structure of the `spy` object
str(spy)

plot(spy)

# Assign JNJ data to *jnj* using env argument
jnj <- getSymbols(Symbols = "JNJ", env = NULL)

# Look at the structure of the jnj
str(jnj)

plot(jnj)
```

Turning off auto.assign is useful if we want to assign the data to an object ourselves.


## 1.4 Quandl() vs getSymbols()

Defaulst: 

* getSymbols() returns *xts*

* Quandl() returns *data.frame*


```{r}
# Return xts object
dgs10 <- Quandl::Quandl(code = "FRED/DGS10", type = "xts")
head(dgs10)

# set Quandl API key
Quandl.api_key(Sys.getenv("quandl_api_key"))

# Import GDP data from FRED
gdp <- Quandl(code = "FRED/GDP")

# Look at the structure of the object returned by Quandl
str(gdp)

plot(gdp)

# 
quandldata <- quandl.api(path = "datasets/NSE/OIL", http = "GET")

str(quandldata)
```

## 1.5 Return data type
```{r}
# Import GDP data from FRED as xts
gdp_xts <- Quandl(code = "FRED/GDP", type = "xts")

# Look at the structure of gdp_xts
str(gdp_xts)

# Import GDP data from FRED as zoo
gdp_zoo <- Quandl(code = "FRED/GDP", type = "zoo")

# Look at the structure of gdp_zoo
str(gdp_zoo)

head(gdp_zoo)
```


## 1.6 Find Stock Ticker From Yahoo Finance

Search Yahoo Finance for the ticker symbol for Pfizer stock. *getSymbols()* will error if the data are not available for download. 

```{r}
# Create an object containing the Pfizer ticker symbol
symbol <- "PFE"

# Use getSymbols to import the data
getSymbols(Symbols = symbol, src = "yahoo", auto.assign = TRUE)

# Look at the first few rows of data
head(PFE)

# Aemetis, Inc. (AMTX) 
getSymbols(Symbols = "AMTX", src = "yahoo", auto.assign = TRUE)

plot(AMTX$AMTX.Adjusted)
head(AMTX["201811/201812"])
tail(AMTX)
```

## 1.7 Download exchange rate data from Oanda

Oanda.com provides 180 days of historical data. 

```{r}
# see the list of currencies provided by Oanda.com
currencies <- quantmod::oanda.currencies

# Create a currecy_pair object
currency_pair <- "GBP/CAD" # The British Pound and the Canadian Dollar

# Load British Pound to Canadian Dollar exchange rate data
getSymbols(currency_pair, src = "oanda", auto.assign = TRUE)

# Examine object using str()
str(GBPCAD)
plot(GBPCAD)

# USD to RMB
getSymbols("USD/CNY", src = "oanda", auto.assign = TRUE)
tail(USDCNY)

plot(USDCNY)

```

## 1.8 Find and import Unemployment Rate data from FRED

```{r}
# Create a series_name object 
series_name <- "UNRATE" # United States civilian umemployment rate

# Load the data using getSymbols
getSymbols(series_name, src = "FRED", auto.assign = TRUE)
plot(UNRATE)


# Creae a quandl_code object
quandl_code <- "FRED/UNRATE"

# Load the data using Quandl
unemploy_rate <- Quandl(quandl_code, type = "xts")

head(unemploy_rate)
tail(unemploy_rate)

```

**FRED** contains a huge number of economic data series. 


# 2. Extracting and transforming data

You've learned how to import data from online sources, now it's time to see how to extract columns from the imported data. After you've learned how to extract columns from a single object, you will explore how to import, transform, and extract data from multiple instruments.

## 2.1 Extract one column from one instrument

The *quantmod* package provides several helpful functions: 
Op(), Hi(), Lo(), Cl(), Vol(), and Ad().

```{r}
# Amyris company stock 
getSymbols(Symbols = "AMRS", src = "yahoo", auto.assign = TRUE)

plot(AMRS$AMRS.Adjusted)
head(AMRS)
tail(AMRS)

# Extract the close column
AMRS_close <- Cl(AMRS)
head(AMRS_close)
plot(AMRS_close)

AMRS_volume <- Vo(AMRS)
head(AMRS_volume)
plot(AMRS_volume)
```

## 2.2 Extract multiple columns from one instrument
The *quantmod* package provides: OHLC(), HLC(), OHLCV()

```{r}
# Extract the high, low, and close columns of Amyris 
HLC(AMRS) %>% tail()

# Extract open,high, low, close, and volume
OHLCV(AMRS) %>% tail()
```

## 2.3 Use getPrice to extract other columns

```{r, eval=FALSE}
# Download CME data for CL and BZ as an xts object
# set Quandl API key
Quandl.api_key(Sys.getenv("quandl_api_key"))

# Download CME data for CL and BZ as an xts object
oil_data <- Quandl(code = c("CME/CLH2016", "CME/BZH2016"), type = "xts")

# Look at the column names of the oil_data object
colnames(oil_data)

# Extract the Open price for CLH2016
cl_open <- getPrice(oil_data, symbol = "CLH2016", prefer = "Open$")

# Look at January, 2016 using xts' ISO-8601 subsetting
cl_open["2016-01"]
```

## 2.4 Use Quandl to download weekly return data

Common aggregations and transformations: 

* collapse

* transform 



