---
title: "Importing and Managing Financial Data in R"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

**Course Description**

"If you've ever done anything with financial or economic time series, you know the data come in various shapes, sizes, and periodicities. Getting the data into R can be stressful and time-consuming, especially when you need to merge data from several different sources into one data set. This course will cover importing data from local files as well as from internet sources."


Ref: Ulrich, Joshua. "Importing and Managing Financial Data in R". https://www.datacamp.com/courses/importing-and-managing-financial-data-in-r. 2018. 

Note: Some course materials have been revised for internal training by Jeff Li. 


## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# (b) I like to have the package of "tidyverse"
library(tidyverse)
library(quantmod)
```


# 1. Introduction and downloading data

A wealth of financial and economic data are available online. Learn how getSymbols() and Quandl() make it easy to access data from a variety of sources.

What is getSymbols()? 

* Provides access to multiple data sources

* Return xts object by default

* Return data like an ordinary function or create an object like load() does in base R

```{r}
# getSymbols() example
aapl <- getSymbols(Symbols = "AAPL", src = "yahoo", auto.assign = FALSE)
head(aapl)

getSymbols(Symbols = "AAPL", src = "yahoo", auto.assign = TRUE)
head(AAPL)
```

## 1.1 Introducing getSymbols()

The getSymbols() function from the quantmod package provides a consistent interface to import data from various sources into your workspace. By default, getSymbols() imports the data as a xts object.

import QQQ data from Yahoo! Finance. QQQ is an exchange-traded fund that tracks the Nasdaq 100 index, and Yahoo! Finance is the default data source for getSymbols().

```{r}
# Load the quantmod package
library(quantmod)

# Import QQQ data from Yahoo! Finance
getSymbols("QQQ", auto.assign = TRUE)

# Look at the structure of the object getSymbols created
str(QQQ)

# Look at the first few rows of QQQ
head(QQQ)

# plot 
plot(QQQ)
```

## 1.2 Data Sources

Import data from __Alpha Vantage__ and __FRED__. __Alpha Vantage__ is a source similar to Yahoo! Finance. __FRED__ is an online database of economic time series data created and maintained by the Federal Reserve Bank of St. Louis.

```{r}
# Import QQQ data from Alpha Vantage
getSymbols(Symbols = "QQQ", src = "av", auto.assign = TRUE, api.key = Sys.getenv("av_API"))
head(QQQ)

# Look at the structure of QQQ
str(QQQ)

# Plot QQQ
plot(QQQ)

# Import GDP data from FRED
getSymbols(Symbols = "GDP", src = "FRED", auto.assign = TRUE)

# Look at the structure of GDP
str(GDP)

# look at the  head of GDP
head(GDP)

# Plot GDP
plot(GDP)
```

## 1.3 Make *getSymbols() return the data it retrieves
```{r}
# Assign SPY data to *spy* using auto.assign argument
spy <- getSymbols("SPY", auto.assign = FALSE)

# Look at the structure of the `spy` object
str(spy)

plot(spy)

# Assign JNJ data to *jnj* using env argument
jnj <- getSymbols("JNJ", env = NULL)

# Look at the structure of the jnj
str(jnj)

plot(jnj)
```

Turning off auto.assign is useful if we want to assign the data to an object ourselves.