---
title: "Quandl Renewable Fuel Analysis in R "
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---


# (I) Load Required Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# (b) Load libraries
library(Quandl)
library(quantmod)
library(DBI)
library(tidyverse)
library(purrr)
library(caret)
library(rstudioapi)
```


# (II) my Quandl Api Key
```{r}
Quandl.api_key(askForPassword("My Quandl API Key"))
```



# 1. Quandl and Identifying datasets with IDs

## 1.1 Quandl ID

At Quandl, every dataset is identified by a unique id, which can be found on the Quandl website as the *Quandl Code*. 

```{r}

# Assign your first dataset to the variable:
mydata <- Quandl("NSE/OIL")

str(mydata)

# Assign the Prague Stock Exchange to:
PragueStockExchange <- Quandl("PRAGUESE/PX")
head(PragueStockExchange, n = 5)

```

## 1.2 Quandl data format

The Quandl package is able to return data in 4 formats: 

* "raw" - data frame

* "ts" - ts

* "zoon" - zoo

* "xts" - xts

```{r}
# Load the Facebook data with the help of Quandl
Facebook <- Quandl("WIKI/FB", type = "xts")

# Plot the chart with the help of candleChart()
candleChart(Facebook)
```

## 1.3 Searching a Quandl dataset in R

```{r}
# Look up the first 3 results for 'Bitcoin' within the Quandl database:
results <- Quandl.search(query = "BitCoin", silent = FALSE)

# Print out the results
str(results)

# Assign the data set with code B
BitCoin <- Quandl("BCHAIN/TOTBC")

str(BitCoin)



# Look for ethanol data 

Ethanol_fuel <- Quandl.search(query = "Ethanol")

OECD_HIGH_AGLINK_2010_WLD_ET_XP <- Quandl("OECD/HIGH_AGLINK_2010_WLD_ET_XP")

head(OECD_HIGH_AGLINK_2010_WLD_ET_XP)

# Fuel Ethanol consumption 

Fuel_EthOH_Mon_EIA <- Quandl("EIA/STEO_EOTCPUS_M")

```

## 1.4  Working with datatables in Quandl

```{r}
# etrieve all rows from ZACKS/FC for AAPL (Apple) and MSFT (Microsoft). Assign the result to datatable_all
datatable_all <- Quandl.datatable("ZACKS/FC", ticker = c("AAPL", "MSFT"))

# Retrieve all rows for ZACKS/FC for 'AAPL' and 'MSFT'.  Only return the ticker and per_end_date columns
datatable_columns <- Quandl.datatable("ZACKS/FC", ticker = c("AAPL","MSFT"), qopts.columns = c("ticker", "per_end_date"))


```


# 2. Manipulating Quandl Datasets

Use *start_date* and *end_date* arguments


```{r}
# NSE/Oil data

quandldata = Quandl("NSE/OIL", collapse="monthly", start_date="2013-01-01", type="ts")
plot(quandldata[,1])


# the USD/EUR exchange rate 

exchange <- Quandl(code = "CUR/EUR", start_date = "2013-01-01", end_date = "2013-12-01")

plot(exchange)

```


