---
title: "Intermediate R for Finance"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee



**Course Description**

"If you enjoyed the Introduction to R for Finance course, then you will love Intermediate R for Finance. Here, you will first learn the basics about how dates work in R, an important skill for the rest of the course. Your next step will be to explore the world of if statements, loops, and functions. These are powerful ideas that are essential to any financial data scientist's toolkit. Finally, we will spend some time working with the family of apply functions as a vectorized alternative to loops. And of course, all examples will be finance related! Enjoy!" 


Ref: Dirik, Lore. https://www.datacamp.com/courses/intermediate-r-for-finance, 2018.


Note: Some course materials have been revised for internal R training. 

## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```


# 1. Dates 

"Welcome! Before we go deeper into the world of R, it will be nice to have an understanding of how dates and times are created. This chapter will teach you enough to begin working with dates, but only scratches the surface of what you can do with them."

## 1.1 Dates - Lecture 1

* Date - Calendar dates

* POSIX - Portable Operating System Interface - POSIXct and POSIXlt, Date+ time + timezone ("2017-03-14 10:00:21 EDT")

```{r}

# What is the current date?
Sys.Date()

# What is the current date and time?
Sys.time()

# Create the variable today
today <- Sys.Date()


# Confirm the class of today
class(today)

# creating dates
character_date <- "1957-03-04"
class(character_date)

# as.Date
sp500_birthday <- as.Date(character_date)
sp500_birthday
class(sp500_birthday)

```

## 1.2 Create Data
```{r}

# Create crash
crash <- as.Date("2008-09-29")

# Print crash
crash

# crash as a numeric
as.numeric(crash)

# Current time as a numeric
as.numeric(Sys.time())

# Incorrect date format
#as.Date("09/29/2008")

```

```{r}

# Create dates from "2017-02-05" to "2017-02-08" inclusive.
dates <- c("2017-02-05", "2017-02-06", "2017-02-07", "2017-02-08")

# Add names to dates
names(dates) <- c("Sunday", "Monday", "Tuesday", "Wednesday") 

# Subset dates to only return the date for Monday
dates[2]
```

## 1.3 Date formats and extractor functions
```{r}
# ISO 8601 Standard: year-month-day
as.Date("2017-01-28")

# Alternative form: year/month/day
as.Date("2017/01/28")

# Explicitly tell R the format  (%b = Abbreviated month name, %B = Full month name)
as.Date("01/28/2017", format = "%m/%d/%Y")

# Lehman Brothers bankruptcy
as.Date("September 15, 2008", format = "%B %d, %Y")

# Extractor functions
dates <- as.Date(c("2017-01-03", "2017-01-04"))
weekdays(dates)
```

As a reminder, here are the formats:

%Y: 4-digit year (1982)
%y: 2-digit year (82)
%m: 2-digit month (01)
%d: 2-digit day of the month (13)
%A: weekday (Wednesday)
%a: abbreviated weekday (Wed)
%B: month (January)
%b: abbreviated month (Jan)

```{r}


# char_dates
char_dates <- c("1jan17", "2jan17", "3jan17", "4jan17", "5jan17")

# Create dates using as.Date() and the correct format 
dates <- as.Date(char_dates, format ="%d%b%y")

# Use format() to go from "2017-01-04" -> "Jan 04, 17"
format(dates, format = "%b %d, %y")

# Use format() to go from "2017-01-04" -> "01,04,2017"
format(dates, format = "%m, %d, %Y")


```

Only useful to report or export dates. 


## 1.4 Subtraction of dates

```{r}

# Dates
dates <- as.Date(c("2017-01-01", "2017-01-02", "2017-01-03"))

# Create the origin
origin <- as.Date("1970-01-01")

# Use as.numeric() on dates
as.numeric(dates)

# Find the difference between dates and origin
dates - origin

difftime(dates, origin, units = "days")
```

## 1.5 months(), weekdays(), and quarters()

```{r}

# dates
dates <- as.Date(c("2017-01-02", "2017-05-03", "2017-08-04", "2017-10-17"))

# Extract the months
months(dates)

# Extract the quarters
quarters(dates)

# dates2
dates2 <- as.Date(c("2017-01-02", "2017-01-03", "2017-01-04", "2017-01-05"))

# Assign the weekdays() of dates2 as the names()

names(dates2) <- weekdays(dates2)

# Print dates2
dates2
```


# 2. If Statements and Operators

"Imagine you own stock in a company. If the stock goes above a certain price, you might want to sell. If the stock drops below a certain price, you might want to buy it while it's cheap! This kind of thinking can be implemented using operators and if statements. In this chapter, you will learn all about them, and create a program that tells you to buy or sell a stock."


# 3. Loops 

"Loops can be useful for doing the same operation to each element of your data structure. In this chapter you will learn all about repeat, while, and for loops!"

## 3.1 Repeat loops

We must specify when to *break* out of the loop, otherwise it runs for eternity!

```{r}
# don't do infinite loop!
# i <- 0
# repeat{
#   i <- i + 1
#   print(i)
# }

# add a break point
i <- 0
repeat{
  i <- i +1
  print(i)
  
  if(i == 5) {
    break
  }
}

# Repeatedly check a stock price
stock_price <- 52.1

repeat{
  stock_price <- stock_price * runif(1, min = 0.99, max = 1.01)
  print(stock_price)


if(stock_price > 52.5) {
  print("Stock price is above 52.5! Sell!")
  break
  }
}

```

## 3.2 While loop 
```{r}
# example 1
i <- 0
while(i < 3) {
  print(i)
  i <- i + 1
}

# check while true
stock_price <- 52.1

while(stock_price <= 52.5) {
  stock_price <- stock_price * runif(1, 0.99, 1.01)
  print(stock_price)
}
print("Stock price is above 52.5! Sell!")

```

```{r}
# plot with a while loop
debt <- 5000
i <- 0
x_axis <- i
y_axis <- debt

# Initial plot
plot(x_axis, y_axis, xlim = c(0, 10), ylim = c(0, 5000))

# Graph the debt
while(debt > 0) {
  #pay debt 
  debt <- debt - 500
  i <- i + 1
  x_axis <- c(x_axis, i)
  y_axis <- c(y_axis, debt)
  
  # next plot
  plot(x_axis, y_axis, xlim = c(0, 10), ylim = c(0, 5000))
}
```


Add a break point inside *while*: 

```{r}
# debt and cash
debt <- 5000
cash <- 4000

# Pay off your debt...if you can!
while (debt > 0) {
  debt <- debt - 500
  cash <- cash - 500
  print(paste("Debt remaining:", debt, "and Cash remaining:", cash))

   if (cash == 0) {
     print("You ran out of cash!")
     break
   }
}
```

## 3.3 For loops
```{r}
# ex1: for loop
one_to_five <- 1:5
one_to_five

for(number in one_to_five) {
  print(number)
}
```

Loop over a list

```{r}
# create a list
stock_list <- list(stock_name = "Apple", 
                   ticker = "AAPL", 
                   price = 126.5, 
                   good_deal = TRUE)
class(stock_list[[1]])

# check *class*
for (i in stock_list) {
  print(class(i))
}


# loop for data.frame
df <- data.frame(x = 1:3, 
                 y = c("A", "B", "C"), 
                 z = c(TRUE, FALSE, TRUE))

for(i in df) {
  print(class(i))
}
```


Break and next

```{r}
# create a vector
my_vec <- c(1, 20, 3, 4, 10, NA, 78)

# for loop: berak and next
for (value in my_vec) {
  if(is.na(value)) {
    print("Missing value")
    next
  }
  
  if(value > 10) {
    print("Value is greater than 10")
    break
  } else("look OK")
}
```

# 4. Functions

"If data structures like data frames and vectors are how you hold your data, functions are how you tell R what to do with your data. In this chapter, you will learn about using built-in functions, creating your own unique functions, and you will finish off with a brief introduction to packages.

"


```{r}
# tidyquant package
library(tidyquant)

# Pull Apple stock data
apple <- tq_get("AAPL", get = "stock.prices", from = "2007-01-03", to = "2017-06-05")


# take a lokk at what it returned
head(apple)
sto

# plot the stock price over time
plot(apple$date, apple$adjusted, type = "l")

# Calculate daily stock return for the adjusted price
apple <- tq_mutate(data = apple, 
                   ohlc_fun = Ad, 
                   mutate_fun = dailyReturn)

# Sort the return from the least to greatest
sorted_returns <- sort(apple$daily.returns)

# plot them
plot(sorted_returns)
```


# 5. Apply 
"A popular alternative to loops in R are the apply functions. These are often more readable than loops, and are incredibly useful for scaling the data science workflow to perform a complicated calculation on any number of observations. Learn about them here!"


