---
title: "Spatial Data Science"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


ref: Edzer Pebesma, and Roger Bivand. "Spatial Data Science". https://keen-swartz-3146c4.netlify.com/



## (I) Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(spatstat)
library(sf)
```

# 1. Getting Started
## 1.1 A first map

```{r}
# Find the full file names of files in packages of "sf"
nc_32119 <- system.file("gpkg/nc.gpkg", package = "sf") %>%
  # read into R 
  read_sf() %>%
  # Transform into a new coordinate reference system (North Carolina State Plane, with EPSG code 32119)
  # convert geographic coordinates (degrees longitude and latitude) in to "flat"
  st_transform(32119) 


nc_32119 %>%
  # only select one feature
  select(BIR74) %>%
  # plot
  plot(graticule = TRUE, axes = TRUE)


# A ggplot is created we use geom_sf: 
ggplot() +  geom_sf(data = nc_32119) +  aes(fill = BIR74) +
  theme(panel.grid.major = element_line(color = "white")) +
  scale_fill_gradientn(colors = sf.colors(20))



# An interactive, leaflet-type map is obtained by
suppressPackageStartupMessages(library(mapview))

nc_32119 %>% 
  mapview(zcol = "BIR74", legend = TRUE, col.regions = sf.colors)
```


```{r}

```


## 1.2 Reading and Writing 

 Package sf can read from a large number of different data source types, each having its own driver. The following commands show how many vector and raster drivers we have available:
 
```{r}
# vector drivers
st_drivers("vector") %>% nrow()

# raster drivers
st_drivers("raster") %>% nrow() 
```
 
### 1.2.3 reading and writing raster data 

Raster data can be read with function read_stars from package stars

```{r}
library(stars)
tif = system.file("tif/L7_ETMs.tif", package = "stars")
(x = tif %>% read_stars())

plot(x)

```


# 2. Coordinate systems 

