---
title: "Importing Data in R Part 1"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

**Course Description**

"Importing data into R to start your analyses—it should be the easiest step. Unfortunately, this is almost never the case. Data come in all sorts of formats, ranging from CSV and text files and statistical software files to databases and HTML data. Knowing which approach to use is key to getting started with the actual analysis.

In this course, you will get started with learning how to read CSV and text files in R. You will then cover the readr and data.table packages to easily and efficiently import flat file data. After that you will learn how to read XLS files in R using readxl and gdata."


Ref: Schouwenaars, Filip. https://www.datacamp.com/courses/importing-data-in-r-part-1. 2018. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (I) Load Required Libraries
```{r, message = FALSE}
# (a) these packages make creating packages so easy (but not a piece of cake!)
library(XLConnect)

# (b) I like to have the package of "tidyverse"
library(tidyverse)
```

# 1. Importing data from flat files with utils

There are 5 types of files: 

* Flat files

* Data from Excel

* Database

* Web 

* Statistical software

`path <- file.path("~", "datasets", "states.csv")`

`read.csv(path)` 

## 1.1 read.csv  

```{r}
# I already copied several files into the data folder
dir("data/")

# read swimming_pools.csv

pools <- read.csv("data/swimming_pools.csv")
str(pools)

# I prefer to use

pools_2 <- read_csv("data/swimming_pools.csv" )


# stringsAsFactors: in the *utils* package, stringsAsFactors = TRUE

pools <- read.csv("data/swimming_pools.csv", stringsAsFactors = FALSE)
str(pools)

```

## 2.2 read.delim & read.table








# 4. Reproducible Excel work with XLConnect

XLConnect was wrote by Martin Studer.  It works with Excel through R; Bridge between Excel adn R; XLS and XlSX. 


`library("XLConnect")`

`book <- loadWorkbook("cites.xlsx")`

`book` 

`getSheets(book)`

`library(readxl)`

`excel_sheets("cites.xlsx")`

`readWorksheet(book, sheet = "xx", startRow = x, endRow = x, startCol = x, header = FALSE)`

## 4.1 Connect to a workbook 

The first step is to build a "bridge" between Excel and R with `loadWorkbook()`. 

```{r}

# Build connection to urbanpop.xlsx: my_book
my_book <- loadWorkbook("data/urbanpop.xlsx")

# Print out the class of my_book
class(my_book)
```

## 4.2 List and read Excel sheets
```{r}
# List the sheets in my_book
sheets <- getSheets(my_book)
sheets

# Import the second sheet in my_book
readWorksheet(my_book, sheet = "1967-1974")

```

## 4.3 Customize readWorksheet

Without opening up the Excel file, let's get an overview about the excel file

```{r}
all <- lapply(sheets, readWorksheet, object = my_book)

# str(all)
```

We can select *sheet*, *startCol*, and *endCol*. 

```{r}

# Import columns 3, 4, and 5 from second sheet in my_book: urbanpop_sel
urbanpop_sel <- readWorksheet(my_book, sheet = 2, startCol = 3, endCol = 5)

# Import first column from second sheet in my_book: countries
countries <- readWorksheet(my_book, sheet = 2, startCol = 1, endCol = 1)

# cbind() urbanpop_sel and countries together: selection

selection <- cbind(countries, urbanpop_sel)

# I like this method

selection <- bind_cols(countries, urbanpop_sel)

```


## 4.4 Adapting sheets

```{r}
# (a) Create a new data
# pop_2010 <- data.frame(
#   Capital = c("New York", "Berline", "Madrid", "Stockholm"), 
#   Population = c(8191900, 3460725, 3273000, 1372565)
# )
# 
# pop_2010
# 
# # (b) loadWorkbok
# book <- loadWorkbook("data/cities.xlsx", create = TRUE, password = "1234")
# book
# 
# # (c) createSheet
# # createSheet(book, name = "year_2010")
# 
# # (d) Write worksheet
# writeWorksheet(book, pop_2010, sheet = "year_2010")
# 
# # (e) Save
# saveWorkbook(book, file = "data/cities.xlsx")
# 
# # (f) rename sheet
# renameSheet(book, "year_2010", "Y2010")
# 
# saveWorkbook(book, file = "data/cities.xlsx")

# (f) remove sheet
# removeSheet()
#removeWorkbook(book, file = "data/cities.xlsx")
```

Also, we can add more functionality: Styling cells, working with formulas, arranging cells.  


## 4.5 Add worksheet, populate worksheet, reaneme sheets

The *readxl* and *gdata* can import Excel data. *XLConnect* is more powerful, we can edit excel files inside R. 

```{r}
# Build connection to urbanpop.xlsx: my_book
my_book <- loadWorkbook("data/urbanpop.xlsx")

# Add a worksheet to my_book, named "data_summary"
createSheet(my_book, "data_summary")

# Use getSheets() on my_book
sheets <- getSheets(my_book)
sheets

# create data frame: summ

dims <- sapply(sheets[1:3], function(x) dim(readWorksheet(my_book, sheet = x)), USE.NAMES = FALSE)
dims

summ <- data.frame(sheets = sheets[1:3], 
                   nrows = dims[1, ], 
                   ncols = dims[2, ])

summ

# Add data in summs to "data_summary" sheet
writeWorksheet(my_book, summ, sheet = "data_summary")

# save workbook as summary.xlsx
saveWorkbook(my_book, file = "data/summary.xlsx")

# rename "data_summary" sheet to "summary"
my_book <- loadWorkbook("data/summary.xlsx")
renameSheet(my_book, "data_summary", newName = "summary")

# print out sheets of my_book
getSheets(my_book)

# Save workbook to "renamed.xlsx"
saveWorkbook(my_book, file = "data/renamed.xlsx")
```

## 4.6 Remove sheets
```{r}
# build a connection with "renamed.xlsx"
my_book <- loadWorkbook("data/renamed.xlsx")
getSheets(my_book)

# remove sheet
removeSheet(my_book, sheet = "summary")

# Same workbook to "clean.xlsx"
saveWorkbook(my_book, file = "data/clean.xlsx")
```

# 5. The XLConnect Package
Author: Anna Maria Kseiezopolska (April 2018)

The following materials are from XLConnect, and it is a good top-leve introduciton.  It can be opened the user manual by entering the following command: 
```{r}
# vignette("XLConnect")
```

## 5.1  Basic Usage and Examples 

*loadWorkbook()* - loading/creating an Excel workbook

* Load and excel workbook (create if not existing): `loadWorkbook(filename, create = TRUE)`

* Create a worksheet of a chose within a workbook: `createSheet(object, name)`

* Write into sheets within an Excel workbook: `writeWorksheet(object, data, sheet, startRow = 1, startCol = 1, header = TRUE)`

* Create a  name for a specified formula within a workbook: `createName(object, name, formula, overwrite)`

* Write a named range into a workbook: `writeNamedRegion(object, data, name, header)`

* Save a workbook t oa cholse Excel file: `saveWorkbook(object)`

* `writeWorksheetToFile()`: a wrapper function, calling `loadWorkbook()`, `createSheet()`, and `saveWorkbook()` functions subsequently. 

* `writeNamedRegionToFile()`: a wrapper function, calling `loadWorkbook()`, `createSheet()`, `createName()`, `writeNamedRegion()`, and `saveWorkbook()`. 


### 5.1.1 An example: writing a sheet

```{r}
# Four steps  
# First 3 steps are perfomed in -memory, and only at the point when we call *saveWorkbook(), the Excel file is written to disk. 

wb <- loadWorkbook("data/XLConnectExample1.xlsx", create = TRUE)
createSheet(wb, name = "chickSheet")
writeWorksheet(object = wb, data = ChickWeight, sheet = "chickSheet", startRow = 3, startCol = 4)
saveWorkbook(wb)  

```

### 5.1.2 Example: writing a sheet with one call
```{r}
writeWorksheetToFile("data/XLConnectExample2.xlsx", data = ChickWeight, sheet = "chickSheet", startRow = 3, startCol = 4)
```

### 5.1.3 Example: writing a named region

```{r}
wb <- loadWorkbook("data/XLConnectExample3.xlsx", create = TRUE)
createSheet(wb, name = "womenData")
createName(wb, name = "womenName", formula = "womenData!$C$5", overwrite = TRUE)
writeNamedRegion(wb, data = women, name = "womenName" )
saveWorkbook(wb)

```

### 5.1.4 Example: writing a named region with one call
```{r}
# writeNamedRegionToFile("data/XLConnectExample4.xlsx", data = women, name = "womenName", formula = "womenData!$C$5" )
```

## 5.2 Reading Excel files: basic functions

* Read data from a worksheet of an Excel workbook: `readWorksheet(object, sheet, startRow, startCol, endRow, endCol, header = TRUE)`

* Read data from a worksheet of an Excel workbook with one: `readWorksheetFromFile(file, ...)`

* Read data from a named region of an Excel workbook: `readNamedRegion(object, name, header = TRUE)`

* Read data from a named region of an Excel file: `readNamedRegionFromFile(file, name, header = TRUE)`

### 5.2.1 Example: reading from an Excel sheet

```{r}
wb <- loadWorkbook("data/XLConnectExample1.xlsx", create = TRUE)

data <- readWorksheet(wb, sheet = "chickSheet", startRow = 0, endRow = 10, startCol = 0, endCol = 0)

head(data)
```

### 5.2.2 Example: reading from an Excel sheet with one call

```{r}
data <- readWorksheetFromFile("data/XLConnectExample1.xlsx", sheet = "chickSheet", startRow = 0, endRow = 10, startCol = 0, endCol = 0)
# head(data)
```

### 5.2.3 Example: reading from an Excel named region

```{r}
wb <- loadWorkbook("data/XLConnectExample3.xlsx", create = TRUE)
data <- readNamedRegion(wb, name = "womenName") # the Named Region was created before. 
data

# or with one call

data <- readNamedRegionFromFile("data/XLConnectExample3.xlsx", "womenName")
data
```

