---
title: "Multivariate Probability Distributions in R"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


**Course Description**

"When working with data that contains many variables, we are often interested in studying the relationship between these variables using multivariate statistics. In this course, you'll learn ways to analyze these datasets. You will also learn about common multivariate probability distributions, including the multivariate normal, the multivariate-t, and some multivariate skew distributions. You will then be introduced to techniques for representing high dimensional data in fewer dimensions, including principal component analysis (PCA) and multidimensional scaling (MDS)."

Ref: Ray, Surajit (2018) Multivariate Probability Distributions in R, https://www.datacamp.com/courses/, 2018.


Note: Some course materials and data have beem revised for training by Jeff Li. 

# (I) Load required libraries
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(gmodels)
```

# 1. Reading and plotting multivariate data

In this introduction to multivariate data, you will learn how to read and summarize it. You will learn how to summarize multivariate data using descriptive statistics, such as the mean vector, variance-covariance, and correlation matrices. You'll then explore plotting techniques to provide insights into multivariate data.

## 1.1 Reading multivariate data

```{r}
# Read in the wine dataset
wine_url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data"
wine <- read.table(wine_url, sep = ",")

# Print the first four entries
head(wine, n = 4)

# Find the dimensions of the data
dim(wine)

# Check the names of the wine dataset
names(wine)

# Assign new names 
names(wine) <- c('Type', 'Alcohol', 'Malic', 'Ash', 'Alcalinity', 'Magnesium', 'Phenols', 'Flavanoids', 'Nonflavanoids','Proanthocyanins', 'Color', 'Hue', 'Dilution', 'Proline')

# Check the new column names
names(wine)

# Check data type/structure of each variable 
str(wine)

# change the Type variable data type
wine$Type <- factor(wine$Type)
str(wine)
```

## 1.2 Calculating the mean vector
```{r}
# Calculate the mean of the Alcohol, Malic, Ash, and Alcalinity variables
colMeans(wine[, 2:5])

# Calculate the mean of the variables by wine type
by(data = wine[, 2:5], INDICES = wine$Type, FUN = colMeans)
```

## 1.3 Calculating the variance-covariance matrix

Along with the mean, an equally important statistic for a multivariate observation is its variance-covariance matrix. The variance-covariance matrix generalizes the concept of variance to multiple dimensions. 

```{r}
# Calculate the variance-covariance matrix of the variables Alcohol, Malic, Ash, Alcalinity
var_wine <- var(wine[, 2:5])

# Round the matrix values to two decimal places
round(var_wine, digits = 2)
```

## 1.4 Calculating the correlation matrix
A correlation matrix is a generalization of the concept of correlation between two variables. If we are looking at a large number of variables, it can also be useful to visualize the correlation matrix to understand the relationship of the variables on each other. 

```{r}
# calculate the correlation matrix
cor_wine <- cor(wine[, 2:5])

# Round the matrix to two decimal places
round(cor_wine, digits = 2)

# Plot the correlations 
corrplot::corrplot(corr = cor_wine, method = "ellipse")
```

## 1.5 Pairs plot using base graphics and lattice
```{r, message=FALSE}
# Scatter plot matrix using the base R plot function
pairs(wine[, 2:5])
plot(wine[,2:5])

# Scatter plot matrix with lattice
lattice::splom(~wine[,2:5], pch = 16, col = wine$Type)
```

## 1.6 Plotting multivariate data using ggplot
```{r, message=FALSE}
# Produce a matrix of plots for the first four variables
wine_gg <- GGally::ggpairs(data = wine, columns = 2:5, mapping = aes(color = Type))
wine_gg
```

## 1.7 3D plotting techniques

```{r}
# plot the three variables
scatterplot3d::scatterplot3d(x = wine$Alcohol, y =wine$Malic, z = wine$Alcalinity, color = wine$Type)
```


# 2. Multivariate Normal Distribution

Multivariate normal is the most important and widely used multivariate probability distribution. This chapter covers how to generate random samples from a multivariate normal distribution and how to calculate and plot the densities and probabilities under this distribution. 

## 2.1 Samples from multivariate normal distributions

```{r}
# Define mean and variance-covariance matrix
mu_sim <- c(2, -2)
sigma_sim <- matrix(c(9,5, 5, 4), nrow = 2)

# Generate 100 bivariate normal samples
library(mvtnorm)
multnorm_sample <- mvtnorm::rmvnorm(n = 100, mean = mu_sim, sigma = sigma_sim)

# Scatterplot of the bivariate samples
cor(multnorm_sample)
plot(multnorm_sample)

```

