---
title: "Multivariate Probability Distributions in R"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"A happy life is one spent in learning, earning, and yearning". - Lillian Gish


**Course Description**

"When working with data that contains many variables, we are often interested in studying the relationship between these variables using multivariate statistics. In this course, you'll learn ways to analyze these datasets. You will also learn about common multivariate probability distributions, including the multivariate normal, the multivariate-t, and some multivariate skew distributions. You will then be introduced to techniques for representing high dimensional data in fewer dimensions, including principal component analysis (PCA) and multidimensional scaling (MDS)."

Ref: Ray, Surajit (2018) Multivariate Probability Distributions in R, https://www.datacamp.com/courses/, 2018.


Note: Some course materials and data have beem revised for training by Jeff Li. 

# (I) Load required libraries
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(gmodels)
library(MVN)

```

# 1. Reading and plotting multivariate data

In this introduction to multivariate data, you will learn how to read and summarize it. You will learn how to summarize multivariate data using descriptive statistics, such as the mean vector, variance-covariance, and correlation matrices. You'll then explore plotting techniques to provide insights into multivariate data.

## 1.1 Reading multivariate data

```{r}
# Read in the wine dataset
wine_url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data"
wine <- read.table(wine_url, sep = ",")

# Print the first four entries
head(wine, n = 4)

# Find the dimensions of the data
dim(wine)

# Check the names of the wine dataset
names(wine)

# Assign new names 
names(wine) <- c('Type', 'Alcohol', 'Malic', 'Ash', 'Alcalinity', 'Magnesium', 'Phenols', 'Flavanoids', 'Nonflavanoids','Proanthocyanins', 'Color', 'Hue', 'Dilution', 'Proline')

# Check the new column names
names(wine)

# Check data type/structure of each variable 
str(wine)

# change the Type variable data type
wine$Type <- factor(wine$Type)
str(wine)
```

## 1.2 Calculating the mean vector
```{r}
# Calculate the mean of the Alcohol, Malic, Ash, and Alcalinity variables
colMeans(wine[, 2:5])

# Calculate the mean of the variables by wine type
by(data = wine[, 2:5], INDICES = wine$Type, FUN = colMeans)
```

## 1.3 Calculating the variance-covariance matrix

Along with the mean, an equally important statistic for a multivariate observation is its variance-covariance matrix. The variance-covariance matrix generalizes the concept of variance to multiple dimensions. 

```{r}
# Calculate the variance-covariance matrix of the variables Alcohol, Malic, Ash, Alcalinity
var_wine <- var(wine[, 2:5])

# Round the matrix values to two decimal places
round(var_wine, digits = 2)
```

## 1.4 Calculating the correlation matrix
A correlation matrix is a generalization of the concept of correlation between two variables. If we are looking at a large number of variables, it can also be useful to visualize the correlation matrix to understand the relationship of the variables on each other. 

```{r}
# calculate the correlation matrix
cor_wine <- cor(wine[, 2:5])

# Round the matrix to two decimal places
round(cor_wine, digits = 2)

# Plot the correlations 
corrplot::corrplot(corr = cor_wine, method = "ellipse")
```

## 1.5 Pairs plot using base graphics and lattice
```{r, message=FALSE}
# Scatter plot matrix using the base R plot function
pairs(wine[, 2:5])
plot(wine[,2:5])

# Scatter plot matrix with lattice
lattice::splom(~wine[,2:5], pch = 16, col = wine$Type)
```

## 1.6 Plotting multivariate data using ggplot
```{r, message=FALSE}
# Produce a matrix of plots for the first four variables
wine_gg <- GGally::ggpairs(data = wine, columns = 2:5, mapping = aes(color = Type))
wine_gg
```

## 1.7 3D plotting techniques

```{r}
# plot the three variables
scatterplot3d::scatterplot3d(x = wine$Alcohol, y = wine$Malic, z = wine$Alcalinity, color = wine$Type)
```


# 2. Multivariate Normal Distribution

Multivariate normal is the most important and widely used multivariate probability distribution. This chapter covers how to generate random samples from a multivariate normal distribution and how to calculate and plot the densities and probabilities under this distribution. 

## 2.1 Samples from multivariate normal distributions

* Normal: rnorm, dnorm, pnorm, qnorm

* Multivariate Normal: rmvnorm, dmvnorm, pmvnorm, qmvnorm

* t: rt, dt, pt, qt

* multivariate t: rmvt, dmvt, pmvt, qmvt


p for "probability", q for "quantile", d for "density", and r for "random". 


```{r}
# Define mean and variance-covariance matrix
mu_sim <- c(2, -2)
sigma_sim <- matrix(c(9,5, 5, 4), nrow = 2)

# Generate 100 bivariate normal samples
library(mvtnorm)
multnorm_sample <- mvtnorm::rmvnorm(n = 100, mean = mu_sim, sigma = sigma_sim)

# Scatterplot of the bivariate samples
cor(multnorm_sample)
plot(multnorm_sample)

```


Generate 1000 samples from a 3 dimensional norma
```{r}
# define mean, variance and co-variance
mu1 <- c(1, 2, -5)
sigma1 <- matrix(c(1, 1, 0, 
                   1, 2, 0, 
                   0, 0, 5), nrow = 3, ncol = 3)

# generate samples
set.seed(34)
sample_3var <- rmvnorm(n = 1000, mean = mu1, sigma = sigma1)

# plot
cor(sample_3var)
var(sample_3var)

pairs(sample_3var)
```


## 2.2 Calculate the density of a distribution

```{r}
# define mu and sigma
mu1 <- c(1, 2)
sigma1 <- matrix(c(1, 0.5, 0.5, 2), 2)

# calculate density
dmvnorm(x = c(0, 0), mean = mu1, sigma = sigma1)
```

```{r}
# Code for plotting bivariate densities
# Create grid
d <- expand.grid(seq(-3, 6, length.out = 50), 
                 seq(-3, 6, length.out = 50))

# Calculate density on grid
dens1 <- dmvnorm(as.matrix(d), 
                 mean = c( 1, 2), 
                 sigma = matrix(c(1, 0.5,
                                  0.5, 1), nrow = 2))

# Convert to matrix
dens1 <- matrix(dens1, nrow = 50)

# Use perspective plot
persp(dens1,theta = 80, phi = 30, expand = 0.6, shade = 0.2, 
      col = "lightblue", xlab = "x", ylab = "y", zlab = "dens")
```



```{r}
# calcualting the density of multivariate normal
multnorm_dens <- dmvnorm(multnorm_sample, mean = mu_sim, sigma = sigma_sim)

# create scatter plot of density heights
scatterplot3d::scatterplot3d(cbind(multnorm_sample, multnorm_dens), 
              color = "blue", pch = "", type = "h", 
              xlab = "x", ylab = "y", zlab = "density")

# Calculate density over the specified gride
mvals <- expand.grid(seq(-5, 10, length.out = 40), 
                     seq(-8, 4, length.out = 40))

# Calculate density over the specified grid
mvds <- dmvnorm(mvals, mean = mu_sim, sigma = sigma_sim)
matrix_mvds <- matrix(mvds, nrow = 40)

# create a perspective plot
persp(matrix_mvds, theta = 80, phi = 30, expand = 0.6, 
      shade = 0.2, col = "lightblue", xlab = "x", ylab = "y", 
      zlab = "dens")

```


## 2.3 Using pmvnorm function 

In the following example, we will use *pmvnorm()* function to calculate the cumulative distribution for specififed bivaraite norm

The *pmvnorm()* can calculate the cumulative distributions of multivariate normals. 
```{r}
# Valume undere a bivariate standard norml
pmvnorm(lower = c(-1, -1), upper = c(1, 1))

# Volume under specifed mean and variance-covariance matrix
pmvnorm(lower = c(-5, -5), upper = c(5, 5), mean = mu_sim, sigma = sigma_sim)
```

## 2.4 Calculate probability contours using *qmvnorm()*

```{r}
# Probability contours for a standard bivariate normal 
qmvnorm(p = 0.9, tail = "both", sigma = diag(2))

# Probability contours for a bivariate normal
qmvnorm(p = 0.95, tail = "both", mean = mu_sim, sigma = sigma_sim)
```


## 2.5 Checking normality of multivariate data


* Classical statistical techniques that assume univariate/multivariate normaliy: 

   * Multivariate regression
   
   * Discriminant analysis
   
   * Model-based clustering
   
   * Principal component analysis (PCA)
   
   * Multivariate analysis of variance (MANOVA)
   
*qqnorm()* + *qqline()* : for univariate normality tests

Deviation from the line might indicate: heavier tails, skewness, outliers, and clustered data. 

*uniPlot()* for all variables

```{r}
qqnorm(iris[, 1])
qqline(iris[, 1])

# qqnorm of all variable
library(MVN)
MVN::uniPlot(iris_raw[, 1:4])

setosa = iris[1:50, 1:4] # Iris data only for setosa and four variables
uniPlot(setosa, type = "qqplot")
```






