---
title: "Linear Algebra For Data Science in R"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

Reference: Eager, E. Linear Algebra For Data Science in R, www.datacamp.com, Paid class: 10/18/2018 - 10/20/2018. 

This is my course note, I have modified some materials for my teaching and practicing. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (I) Load Required Libraries
```{r, message = FALSE}
library(tidyverse)
```

# 1. Introduction to Linear Algebra

## 1.1 Create vector in R 
```{r}
# Creating three 3's and four 4's, respectively
rep(3, 3)
rep(4, 4)

# Creating a vector with the first three even numbers and the first three odd numbers
seq(2, 6, by = 2)
seq(1, 5, by = 2)

# Re-creating the previous four vectors using the 'c' command
c(3, 3, 3)
c(4, 4, 4, 4)

c(2, 4, 6)
c(1, 3, 5)
```

In **R**, We can create vectors using some simple commands. 


## 1.2 The Algebra of Vectors

```{r}
# create the vectors x, y, and z
(x <- 1:7)
(y <- seq(2, 14, by = 2))
(z <- c(1, 1, 2))

# Add x to y and print

print(x + y)

# Multiply z by 2 and print
print(2 * z)

# Multiply x and y by each other and print
print(x * y)

# Add x to z, if possible, and print
print(x + z)

```

## 1.3 Creating Matrices in R
```{r}
# Create a matrix of all 1's and all 2's that are 2 by 3 and 3 by 2, respectively
matrix(1, 2, 3)

print(matrix(2, 3, 2))

# Create a matrix and changing the byrow designation.
matrix(c(1, 2, 3, 2), nrow = 2, ncol = 2, byrow = FALSE)
matrix(c(1, 2, 3, 2), nrow = 2, ncol = 2, byrow = TRUE)

# Add A to the previously-created matrix
A <- matrix(1, 2, 2)
A + matrix(c(1, 2, 3, 2), nrow = 2, ncol = 2, byrow = TRUE)
```


## 1.4 Matrix Multiplication as a Transformation

* Stretch 

* Shrink

* Reflect 

The following examples show how multiplication by a matrix can alter a vector. 

```{r}
# (a) Stretch: Need to stretch the x (first) componenet of a vector by factor of four: c(1, 1)
A <- matrix(c(4, 0, 0, 1), nrow = 2, ncol = 2)

A %*% c(1, 1)

# (b) Shrinks the y (second) componenet of a vector b <- c(1, 1) by 50%
B <- matrix(c(1, 0, 0, 0.5), nrow = 2, ncol =2 )
B %*% c(1, 1)

# (c) reflect a vector about the y-axis
A <- matrix(c(-1, 0, 0, 1), nrow = 2, ncol = 2)
A

A %*% c(2, 2)

# (d) reflect a vector about the x-axis
B <- matrix(c(1, 0, 0, -1), nrow = 2, ncol =2)
B %*% c(2, 2)

# (e) rotation

C <- matrix(c(-1, 0, 0, -1), nrow = 2, ncol =2 )
C %*% c(2, 2)

```

## 1.5 Matrix Multiplication 

* Matrix multiplication is a complex transformation. 

* Matrix multiplication - Order matters

* matrix inverse (x * 1/x = 1, A %*% Ainv = I)

```{r}
# Take the inverse of the 2 by 2 identity matrix
solve(diag(2))

# Take the inverse of the matrix A
A <- matrix(c(1, -1, 2, 2), nrow = 2, ncol = 2, byrow = FALSE)
Ainv <- solve(A)

# Multiply A by its inverse on the left
Ainv%*%A

# Multiply A by its inverse on the right
A%*%Ainv

```


# 2. Matrix-Vector Equations

WNBA Data: Massey Matrix 12 x 12
```{r}
M



```

