---
title: "Object-Oriented Programming in R: S3 and R6"
author: "Jeff Li"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
---

"I maintained my edge by always being a student; you will always have something new to learn". - Jackie Joyner Kersee


**Course Description**

"Object-oriented programming (OOP) lets you specify relationships between functions and the objects that they can act on, helping you manage complexity in your code. This intermediate level course provides an introduction to OOP and to use the S3 and R6 systems. The S3 system is a great day-to-day R programming tool that simplifies some of the functions that you write. The R6 system is especially useful for industry-specific analyses, working with web APIs, and building GUIs. The course concludes with an interview with Winston Chang, creator of the R6 package.."

Ref: Cotton, Richie, (2018). "Introduction to Time Series Analysis", https://www.datacamp.com/.

Note: Some course materials and data have been revised for training by Jeff Li. 

# (I) Load required libraries
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(pryr)
library(R6)
library(assertive)
```

# 1. Introduction to Object-Oriented Programming 

Learn what object-oriented programming (OOP) consists of, when to use it, and what OOP systems are available in R. You'll also learn how R identifies different types of variable, using classes, types, and modes.

* With *functional programming*, think about the *funciton first*

* With *object-oriented programming (OOP)* think about the *data structures first*

* Don't use OOP for general purpose data analyses

* Do use OOP when you have a limited number of complex objects


## 1.1 Should I OOP?

Object-oriented programming (OOP) is very powerful, but not appropriate for every data analysis workflow. Which of the following scenarios are a good fit for using object-oriented programming?

* Cleaning up a dirty dataset. --NO

* Writing an interface to the Internet Movie Database API. ---Yes

* Creating objects to work with cartographic data for spatial analysis. --Yes 

* Using ggplot2 to visualize your dataset. ---No 


## 1.2 Working with objects 

Commom R objects include *numeric, logical, and character vectors*, as well as *data.frame*.  

One of the priciples of OOP is that functions can behave differently for different kinds of object. 

The *summary()* function is a good example of this. Since different types of variable need to be summarized in different ways, the output that is displayed varies depending upon what you pass into it. 

```{r}
# Create several variables
a_numeric_vector <- rlnorm(50) # log normal distribution 

a_factor <- factor(
  sample(c(LETTERS[1:5], NA), 50, replace = TRUE)
)

a_data_frame <- data.frame(
  n = a_numeric_vector, 
  f = a_factor
)

a_linear_model <- lm(dist ~ speed,data = cars)

# Call summary on these objects
summary(a_numeric_vector)

summary(a_factor)

summary(a_data_frame)

summary(a_linear_model)
```

## 1.4 Nine Systems

S3, R6, S4, ReferenceClasses, OOP, mutatr, R5, R.OO, and proto

* Use S3 regularly

* Use R6 when more power is needed

* Use S4 for Bioconductor

* Maybe use ReferenceClasses


## 1.5 How does R Distinguish variables? 

* *class()* is your first choice for determining the kind of variable

* *typeoff()* is also occasionally useful

* *mode()* and *storage.mode()* are old functions; don't use them. 

In R, there are three kinds of functions: 

* *closure*: most of the functions that we come across

* *builtin*: special evaluation mechanism to make them go faster. Like: *length()*

* *special*: lanshouage constructs, like if and while. 


```{r}
# Define a type_info function
type_info <- function(x) {
  c(
    class = class(x), 
    typeof = typeof(x), 
    mode  = mode(x), 
    storage.mode = storage.mode(x)
  )
}

# Create list of example variables
some_vars <- list(
  an_integer_vector = rpois(24, lambda = 5),
  a_numeric_vector = rbeta(24, shape1 = 1, shape2 = 1),
  an_integer_array = array(rbinom(24, size = 8, prob = 0.5), dim = c(2, 3, 4)),
  a_numeric_array = array(rweibull(24, shape = 1, scale = 1), dim = c(2, 3, 4)),
  a_data_frame = data.frame(int = rgeom(24, prob = 0.5), num = runif(24)),
  a_factor = factor(month.abb),
  a_formula = y ~ x,
  a_closure_function = mean,
  a_builtin_function = length,
  a_special_function = `if`
)

# Loop over some_vars calling type_info() on each element to explore them
lapply(some_vars, type_info)
```

## 1.6 Assigning Classes

* We can override the class()

* This won't break existing functionality


# 2. Using S3 

S3 is a very simple object-oriented system that lets you define different behavior for functions, depending upon their input argument. This chapter explains how to use S3, and how generics and methods work.


## 2.1 Generics and methods

Function overloading: input-dependent function behavior

function: generic + method. Methods are named generic.class. For example: print.Date, summary.factor, unique.array. 

```{r}
args(print)
args(print.Date)
print.function
print.Date
```


## 2.2 What's in a Name? 

S3 uses a strict naming conversion: all S3 methods have a name of the form: **generic.class**. 

Early days of the S language, many of the functions have a name containing a dot without being an S3 method. For example, all.equal() is actually an S3 generic, not a method. 

Two methods to check if a function is an S3 generic by calling *is_s3_generic()* from the pryr package, and just print it. 

Check if a function is an S3 method by calling *is_s3_method()*. 

```{r}
is_s3_generic("t")
t

is_s3_method("t")
is_s3_method("t.matrix")
is_s3_method("t.test")
is_s3_generic("t.test") # early days of the S languange code. 

is_s3_method("is.complex")
is_s3_generic("is.complex")
is.complex

is_s3_method('seq.Date')
is_s3_generic("seq")
seq

is_s3_method("is.na.data.frame")
is_s3_generic("is.na")
is.na

is_s3_generic("sort")
is_s3_generic("order")
is_s3_method("order")
```


## 2.3 Creating a Generic Function  and an S3 method 

To create an S3 functions, the first step is to write the generic. 

```{r}
# Create get_n_elements
get_n_elements <- function(x, ...) {
  UseMethod("get_n_elements")
}

# Create a data.frame method for get_n_elements
get_n_elements.data.frame <- function(x, ...) {
  length(unlist(x))
}

# Call the method on the sleep dataset
get_n_elements(sleep)

# Create a default method for get_n_elements
get_n_elements.default <- function(x, ...) {
  length(unlist(x))
}

# create  a vector
v <- 1:10
get_n_elements(v)

# test a matrix
mat <- matrix(1:10, nrow =2)
get_n_elements(mat)

```

## 2.4 Finding available methods for an S3 generic and for a class 

to find all the available methods for an S3 generic, call the *methods()* function

```{r}
is_s3_generic("wilcox.test")
wilcox.test

# Avialable for "wilcox.text"
methods(generic.function = "wilcox.test")


# available methods for a class "lm"
methods(class = "lm")
is_s3_method("summary.lm")
is_s3_method(   "plot.lm")
is_s3_method(  "anova.lm")

.S3methods(class = "lm")
.S4methods(class = "lm")
```


## 2.5 Method lookup for primitive generics

Some core functionality of R is defined using *primitive* functions, which use a special technique for accessing C-code. 

```{r}
exp
sin

`+`
`*`

`if`

`for`

# list S3 primitive generics
.S3PrimitiveGenerics

methods(generic.function = "dim")
methods(generic.function = "names<-")

is_s3_generic("as.Date") # "as.Date" is not an S# primitive generics

```

Primitive generics don't throw an error when no method is found.  A primitive S3 generics lookup process to find methods: first using the *class* as normal, but nothing is found, the internal C-code function will be called.  

```{r}
methods(generic.function = "length")
length
```

```{r}
# is.na() is  a primitive generic. 
methods(generic.function = "is.na")

is.na(list(TRUE, FALSE, NA))
```

## 2.6 Too much class
```{r}
# define multiple classes
x <- c(1, 3, 6, 10, 15)

# set multiple classes by assigning a character vector to class(). Order from more specific to more general as moving from left to right, so we begin with the behavior most targetted to the object

class(x) <- c("triangular_numbers", "natural_numbers", "numeric") # Triangular number obtained by continued summation of the natural numbers 1, 2, 3, 4, 5, etc. 

# class of x
class(x)

# 

is.numeric

is.numeric(x)

# Use inherits() to test for arbitrary classes
inherits(x = x, what = "triangular_numbers")
inherits(x = x, what = "natural_numbers")
inherits(x = x, what = "numeric")

# Use NextMethod() to chain method calls

# define a generic function
what_am_i <- function(x, ...) {
  UseMethod("what_am_i")
}

# defined chain method calls
what_am_i.triangular_numbers <- function(x, ...) {
  message("I'm triangular numbers")
  NextMethod("what_am_i")
}

what_am_i.natural_numbers <- function(x, ...) {
  message("I'm natural numbers")
  NextMethod("what_am_i")
}

what_am_i.numeric <- function(x, ...) {
  message("I'm numeric")
}

# use the defined chain method
what_am_i(x)

```

Very Classy Exercise 
```{r}
# define kitty
kitty <- "Miaow!"

# Assign classes
class(kitty) <- c("cat", "mammal", "character")

# Does kitty inherit from cat/mammal/character vector?
inherits(kitty, "cat")
inherits(x = kitty, what = "mammal")
inherits(x = kitty, what = "character")

is.character(kitty)

# Does kitty inherit from dog? 
inherits(x = kitty, what = "dog")
```

## 2.7 Writing the Next Method

when objects have multiple classes, we can call methods for several of these classes using *NextMethod()*. 


```{r}
# Inpect the workspace
ls.str()

# cat method
what_am_i <- function(x, ...) {
  UseMethod("what_am_i")
}

# cat method 
what_am_i.cat <- function(x, ...) {
  message("I'm a cat")
  NextMethod("what_am_i")
}

# mammal method
what_am_i.mammal <- function(x, ...) {
  message("I'm a mammal")
  NextMethod("what_am_i")
}

# character method
what_am_i.character <- function(x, ...) {
  message("I'm a character vector")
}

# call what_am_i()
what_am_i(kitty)

```

We can call multiple methods by using NextMethod(). 


# 3. Using R6 

* How to define R6 classes and create R6 objects. 

* Learn the structure of R6 classes

* How to separate the user interface from the implementation details


## 3.1 The Object Factory

**class generators** are templates for objects. a.k.a factories

```{r}
# Load the R6 package ot work with R6!
library(R6)

# Define class generator with R6Class()
thing_factory <- R6Class(classname = "Thing", 
  # Data fields stored in private list
  private = list(
    a_field = "a value", 
    another_field = 123
  )
)

class(thing_factory)

# Create objects with factory's new() method
a_thing <- thing_factory$new()
another_thing <- thing_factory$new()
yet_another_thing <- thing_factory$new()

class(a_thing)

```


## 3.2 Specifying the Microwave Oven Class

To create R6 objects, first create a class generator, sometimes known as a factory. 

```{r}
# Define microwave_oven_factory 
microwave_oven_factory <- R6Class(
  classname = "MicrowaveOven",
  private = list(power_rating_watts = 800))

class(microwave_oven_factory)

# view the microwave_oven_factory
microwave_oven_factory

# Make a new microwave oven  (a new object)
microwave_oven <- microwave_oven_factory$new()


```

## 3.3 Hiding Complexity with encapsulation

* Encapsulation = Separating  "implementation"  || "user interface "

* Store data in private list

* Store methods in public list

* Use private$ to access private elements

* Use self$ to access public elements


```{r}
microwave_oven_factory <- R6Class(
  classname = "MicrowaveOven", 
  private = list(
    power_rating_watts = 800, 
    door_is_open = FALSE
  ), 
  public = list(
    open_door = function() {
      private$door_is_open <- TRUE
    }
  )
)
```

## 3.4 Leaning to Cook 

Using *initialize()*, we can set private fields when objects are created. 

```{r}
# Add a cook method to the factory definition
microwave_oven_factory <- R6Class(
  classname = "MicrowaveOven", 
  
  private = list(
    power_rating_watts = 800,
    door_is_open = FALSE
  ), 
  
  # User-facing functionality is added to the public element
  public = list(
    # cook function
    cook = function(time_seconds) {
      Sys.sleep(time_seconds)
      print("Your food is cooked!")
    }, 
    # open door function
    open_door = function() {
      private$door_is_open = TRUE
    }, 
    # close door function
    close_door = function() {
      private$door_is_open = FALSE
    }, 
    
    # add initializa() method
    initialize = function(power_rating_watts, door_is_open) {
      if(!missing(power_rating_watts)) {
        private$power_rating_watts = power_rating_watts
      }
      if(!missing(door_is_open)) {
        private$door_is_open = door_is_open
      }
  
    }
  )
)

# Create microwave oven object
a_microwave_oven <- microwave_oven_factory$new(power_rating_watts = 650, 
                                               door_is_open = TRUE)
a_microwave_oven

# Call cook method for 1 second
a_microwave_oven$cook(time_seconds = 1)
```

## 3.5 Getting and Setting with Active Bindings

* getting = read the data field

* setting = write the data field

* Active Bindings: defined like functions; accessed like data variables


Control private access with active bindings.  Use assertive to check binding inputs. 


## 3.6 Read the Rating

The data stored by an R6 object is deliberately hidden away from the user by keeping it in the *private* element. This is the principle of *encapsulation*. 

If we want OT provide access to any of the data values (private), we can use an active binding. These are functions that behave like variables. 

Active bindings are stored in the *active* element of an R6 object. To create an active bindings to get a private data fields (i.e.a "read-only" binding), we can create a function with no arguments that simply returns the private element. 

The active binding is called like a data variable, not a function. 

```{r}
# Add a binding for power rating
microwave_oven_factory <- R6Class(
  classname = "MicrowaveOven", 
  
  private = list(
    ..power_rating_watts = 750
  ),
  
  active = list(
    # add the binding here
    power_rating_watts = function() {
      private$..power_rating_watts
    }
  )
)

# Make a microwave
a_microwave_oven <- microwave_oven_factory$new()

# Get the power rating 
a_microwave_oven$power_rating_watts   # active bindings let users examine data fields
```

## 3.7 Control the Power

Active bindings can also be used to set private fields. In this case, the binding functions should accept a single argument, names "value". 

```{r}
# Add a binding for power rating
microwave_oven_factory <- R6Class(
  classname = "MicrowaveOven", 
  private = list(
    ..power_rating_watts = 800, 
    ..power_level_watts = 800
  ),
  
  # add active list containing an active binding
  active = list(
    power_level_watts = function(value) {
      if(missing(value)) {
        private$..power_level_watts
      } else {
        assert_is_a_number(value)
        assert_all_are_in_closed_range(value, lower = 0, upper = private$..power_rating_watts)
        private$..power_level_watts <- value
      }
    }
  )
  
  )

# Make a microwave
a_microwave_oven <- microwave_oven_factory$new()

# Get the power level
a_microwave_oven$power_level_watts

# Try to set the power level to "400"
a_microwave_oven$power_level_watts <- "400"

# Try to set the power level to 1600 watts
a_microwave_oven$power_level_watts <- 1600

# Set the power level to 400 watts
a_microwave_oven$power_level_watts <- 400

a_microwave_oven$power_level_watts
```


# 4. R6 Inheritance

Learn how to inherit from an R6 class, and how the relationship between parent and child classes works.

## 4.1 Propagating Functionality with Inheritance 

* Propagate functionality using inheritance

* Use the inherit arg to R6Class()

* Children get their parent's functionality

*...but the converse is not true


## 4.2 Specifying a fancy microwave oven

Inheritance is used to propagate - this is, copy - functionality from one class to another. 

```{r}
# Define the microwave oven class 
microwave_oven_factory <- R6Class(
  classname = "MicrowaveOven", 
  
  private = list(
    ..power_rating_watts = 800, 
    ..power_level_watts = 800, 
    ..door_is_open = FALSE
  ),
  
  
   # User-facing functionality is added to the public element
  public = list(
    # cook function
    cook = function(time_seconds) {
      Sys.sleep(time_seconds)
      print("Your food is cooked!")
    }, 
    # open door function
    open_door = function() {
      private$door_is_open = TRUE
    }, 
    # close door function
    close_door = function() {
      private$door_is_open = FALSE
    } 
    
     # add initializa() method
    # ,initialize = function(power_rating_watts, door_is_open) {
    #   if(!missing(power_rating_watts)) {
    #     private$power_rating_watts = power_rating_watts
    #   }
    #   if(!missing(door_is_open)) {
    #     private$door_is_open = door_is_open
    #   } }
    ), 
  

  # add active list containing an active binding
  active = list(
    power_level_watts = function(value) {
      if(missing(value)) {
        private$..power_level_watts
      } else {
        assert_is_a_number(value)
        assert_all_are_in_closed_range(value, lower = 0, upper = private$..power_rating_watts)
        private$..power_level_watts <- value
      }
    }, 
    
    power_rating_watts = function(){
      private$..power_rating_watts
    }
  )
  
  )

# Explore the microwave oven class
microwave_oven_factory

# Define a fancy microwave class inheriting from microwave oven
fancy_microwave_oven_factory <- R6Class(
  classname = "FancyMicrowaveOven", 
  inherit = microwave_oven_factory
)

fancy_microwave_oven_factory

# Ceate two objects
a_microwave_oven <- microwave_oven_factory$new()
a_fancy_microwave_oven <- fancy_microwave_oven_factory$new()
a_microwave_oven
a_fancy_microwave_oven

inherits(a_fancy_microwave_oven, "MicrowaveOven")
inherits(a_fancy_microwave_oven, "R6")
inherits(a_microwave_oven, "R6")
```


## 4.3 Making a Fancy Microwave Oven
Inheritance means that the methods of the child class are exact copies of those in the parent class. 

```{r}
# Instantiate both types of microwave
a_microwave_oven <- microwave_oven_factory$new()
a_fancy_microwave_oven <- fancy_microwave_oven_factory$new()

# Get power rating for each microwave
microwave_power_rating <- a_microwave_oven$power_rating_watts
fancy_microwav_power_rating <- a_fancy_microwave_oven$power_rating_watts

# Verify that these are the same
identical(microwave_power_rating, fancy_microwav_power_rating)

# Cook with each microwave
a_microwave_oven$cook(time_seconds = 1)
a_fancy_microwave_oven$cook(time_seconds = 1)

```

